### 2015-08-17 betaRegression/betaRegressionExample.Rmd
added files

---
title: "negativeBinomial"
author: "John Tipton"
date: "August 14, 2015"
output: html_document
---

# Write a beta GLM in STAN

```{r}
library(rstan)
model <- "
// beta Regression
data {
  int<lower=1> N;                   // sample size
  int<lower=1> p;                   // p predictors
  vector<lower=0,upper=1>[N] y;     // response 
  matrix[N, p] X;                   // predictor matrix
}
parameters {
  vector[p] theta;                  // reg coefficients
  real<lower=0> phi;                // dispersion parameter
}
transformed parameters{
  vector[p] beta;
  beta <- theta * 10;               // matt trick if desired
}
model {
  // model calculations
  vector[N] Xbeta;                  // linear predictor
  vector<lower=0,upper=1>[N] mu;    // transformed linear predictor
  vector<lower=0>[N] A;             // parameter for beta distn
  vector<lower=0>[N] B;             // parameter for beta distn
  Xbeta <- X * beta;
  for (i in 1:N) { 
    mu[i] <- inv_logit(Xbeta[i]);   
  }
  A <- mu * phi;
  B <- (1.0 - mu) * phi;
  // priors
  theta ~ normal(0, 1);   
  phi ~ cauchy(0, 5);               // different options for phi  
  //phi ~ inv_gamma(.001, .001);
  //phi ~ uniform(0, 500);          // put upper on phi if using this
  // likelihood
  y ~ beta(A, B);
}
"
write(model, file = "../stanModels/betaRegression-glm.stan")
sm <- stan_model("../stanModels/betaRegression-glm.stan")
```

## Simulate some Data

```{r}
library(betareg)
set.seed(123)
N = 80
beta <- c(2, -1, 1, -2, -1)
x0 <- rep(1, N)
x1 <- c(rep(1, N/2), rep(0, N/2)) ## Indicator for net size
x2 <- rep(c(rep(1, 10), rep(0, 30)), 2) ## Indicator for July
x3 <- rep(c(rep(0, 10), rep(1, 20), rep(0, 10)), 2) ## Indicator for August
# x4 <- rep(c(rep(0, 30), rep(1, 10)), 2) ## Indicator for September
x12 <- x1 * x2 ## Interaction between net size and July
x13 <- x1 * x3 ## Interaction between net size and August
# X <- cbind(x0, x1, x2, x3, x12, x13)
X <- cbind(x1, x2, x3, x12, x13)

mu = plogis(X%*%beta)  # add noise if desired + rnorm(N, sd=.01)
phi = 10
A = mu*phi
B = (1-mu)*phi
y = rbeta(N, A, B)
layout(matrix(c(1, 1, 2, 3), 2, 2, byrow=TRUE))
hist(y, "FD")
hist(y[x1 == 0], "FD")
hist(y[x1 == 1], "FD")
```

## Fit the Model

```{r}
m <- sampling(sm, data = list(N = N, y = y, X = X, p=dim(X)[2]),
  pars = c("beta", "phi"),
  iter = 5000, chains = 4)
```

## Model output
```{r}
m
```

## Plot traceplots
```{r}
rstan::traceplot(m, inc_warmup = FALSE)
```

## Plot fitted vs. simulated values

```{r}
e <- rstan::extract(m, pars = c("beta", "phi"))
true_pars <- c(beta = beta, phi = phi)
library(betareg)
brmod <- betareg(y ~ . -1, data = data.frame(y, X))
summary(brmod)
coefs <- c(brmod$coefficients$mean, brmod$coefficients$precision)
par(mfrow = c(2, 3))
for(i in 1:2) {
  if(i == 1) {
#     for(j in 1:6) {
    for(j in 1:5) {
      plot(density(e[[i]][,j]), main = names(true_pars)[i + j - 1])
      abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
      abline(v = coefs[i + j - 1], lwd = 3, col = "red")
    }
  } else if(i == 2) {
    plot(density(e[[i]]), main = names(true_pars)[i + j - 1])
    abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
    abline(v = coefs[i + j - 1], lwd = 3, col = "red")
  }
}
# legend("topright", legend = c("posterior", "true", "MASS::glm.nb"),
#   col = c("black", "grey", "red"), lty = c(1, 2, 1), lwd = c(1, 1.3, 3),
#   bty = "n")
```

```{r}
library(ggmcmc)
library(coda)
s <-  mcmc.list(lapply(1:ncol(m), function(x) mcmc(as.array(m)[,x,])))
S <- ggs(s)

idx <- order(colMeans(e$beta))

ggs_caterpillar(S, family = "beta") + 
  geom_point(aes(y=1:5, x=beta[idx]), col = "red")
```

```{r}
hist(y[X[, 1] == 0])
hist(y[X[, 1] == 1])
```

# model for later comparison

# simulate yRep as in BDA appendix C; could also do in generated quantities block of stan code
for (s in 1:nSim){
  betaS = betareg_sim$beta[s,]
  muS = plogis(X%*%betaS)
  A = muS*betareg_sim$phi[s]
  B = (1-muS)*betareg_sim$phi[s]
  yRep[s,] = rbeta(dat$N, A, B)
}
str(yRep)

# example, as in BDA p.144, code from appendix C; probably not very useful
par(mfrow=c(5,4), mar=c(4,4,2,2))
hist(y, xlab='', main='y', 'FD')
for (s in 1:19) hist(yRep[s,], xlab='', main=paste('yRep',s), 'FD')
layout(1)

# the following plots get at the same thing
library(ggplot2); library(reshape2)
gdat = data.frame(X,y)
gdatyRep = melt(yRep)

# individual observation distibutions; note the theme is my own (available on github), but otherwise you'll have to define yours.
ggplot(aes(x=y), xlim=c(.2,1), data=gdat) + 
  stat_density(geom='line', lwd=2, alpha=.2) + 
  geom_line(aes(x=value,  group=as.factor(Var2), col=as.factor(Var2)), stat='density', data=gdatyRep, show_guide=F, alpha=.15) +
  ggtheme

# whole response distributions, comparison with betareg fits
ggplot(aes(x=y), xlim=c(.2,1), data=gdat) + 
  geom_line(aes(x=value,  group=as.factor(Var1), col=as.factor(Var1)), stat='density', data=gdatyRep, show_guide=F, alpha=.05) +
  stat_density(geom='line', lwd=2, alpha=.2) + 
  stat_density(aes(x=brmod$fitted), geom='line', col='darkred', lwd=2, alpha=.2) +
  stat_density(aes(x=value), geom='line', col='#FF5500', lwd=2, alpha=.2, data=gdatyRep) +
  ggtheme

# Examine quantiles
sapply(list(yRep=gdatyRep$value, brFitted=brmod$fitted.values, y=y), quantile, p=c(0,.1,.25,.5,.625,.75,.8,.9,.95,1))

### 2015-08-17 negativeBinomial/negativeBinomialExample.Rmd
added files

---
title: "negativeBinomial"
author: "John Tipton"
date: "August 14, 2015"
output: html_document
---

# Write a negative binomial GLM in STAN

```{r}
library(rstan)
model <- "
// Negative Binomial GLM
// see Gelman and Hill (2007)
data {
  int<lower=1> N;       // number of observations
  int<lower=0> p;
  matrix[N, p] X;       // predictor
  int<lower=0> y[N];    // response
} transformed data {
  vector[p] J;
  matrix[p, p] I_p;
  // calculate placeholder values
  J <- rep_vector(0, p);
  I_p <- diag_matrix(rep_vector(1, p));
}
parameters {
  real<lower=0> phi;   // Negative Binomial dispersion parameter
  vector[p] beta;      // Negative Binomial mean parameters
} model {
  // priors:
  phi ~ cauchy(0, 20);
  beta ~ multi_normal(J, 100 * I_p);
  // data model:
  y ~ neg_binomial_2_log(X * beta, phi);
}
"
write(model, file = "~/mysis/stanModels/overdispersedPoisson-glm.stan")
sm <- stan_model("~/mysis/stanModels/overdispersedPoisson-glm.stan")
```

## Simulate some Data

```{r}
set.seed(123)
N <- 80
phi <- 1
# beta <- c(1, 2, -1, 1, -2, -1)
beta <- c(2, -1, 1, -2, -1)
x0 <- rep(1, N)
x1 <- c(rep(1, N/2), rep(0, N/2)) ## Indicator for net size
x2 <- rep(c(rep(1, 10), rep(0, 30)), 2) ## Indicator for July
x3 <- rep(c(rep(0, 10), rep(1, 20), rep(0, 10)), 2) ## Indicator for August
# x4 <- rep(c(rep(0, 30), rep(1, 10)), 2) ## Indicator for September
x12 <- x1 * x2 ## Interaction between net size and July
x13 <- x1 * x3 ## Interaction between net size and August
# X <- cbind(x0, x1, x2, x3, x12, x13)
X <- cbind(x1, x2, x3, x12, x13)
y <- rnbinom(N, size=phi, mu=exp(X %*% beta))
layout(matrix(c(1, 1, 2, 3), 2, 2, byrow=TRUE))
hist(y)
hist(y[x1 == 0])
hist(y[x1 == 1])
```

## Fit the Model

```{r}
m <- sampling(sm, data = list(N = N, y = y, X = X, p=dim(X)[2]),
  pars = c("beta", "phi"),
  iter = 5000, chains = 4)
```

## Model output
```{r}
m
```

## Plot traceplots
```{r}
rstan::traceplot(m, inc_warmup = FALSE)
```

## Plot fitted vs. simulated values

```{r}
e <- rstan::extract(m, pars = c("beta", "phi"))
true_pars <- c(beta=beta, phi=phi)
m2 <- MASS::glm.nb(y ~ X-1)
coefs <- c(coef(m2), summary(m2)$theta)
par(mfrow = c(2, 3))
for(i in 1:2) {
  if(i == 1) {
#     for(j in 1:6) {
    for(j in 1:5) {
      plot(density(e[[i]][,j]), main = names(true_pars)[j])
      abline(v = true_pars[j], lwd = 2, col = "grey", lty = 2)
      abline(v = coefs[j], lwd = 3, col = "red")
    }
  } else if(i == 2) {
    plot(density(e[[i]]), main = names(true_pars)[j + 1])
    abline(v = true_pars[j + 1], lwd = 2, col = "grey", lty = 2)
    abline(v = coefs[j + 1], lwd = 3, col = "red")
  }
}
# legend("topright", legend = c("posterior", "true", "MASS::glm.nb"),
#   col = c("black", "grey", "red"), lty = c(1, 2, 1), lwd = c(1, 1.3, 3),
#   bty = "n")
```

```{r}
library(ggmcmc)
library(coda)
s <-  mcmc.list(lapply(1:ncol(m), function(x) mcmc(as.array(m)[,x,])))
S <- ggs(s)

idx <- order(colMeans(e$beta))

ggs_caterpillar(S, family = "beta") + 
  geom_point(aes(y=1:5, x=beta[idx]), col = "red")
```

### 2015-08-17 countModel/countData.Rmd
update neg binomial model and added real count data example

---
title: "negativeBinomial"
author: "John Tipton"
date: "August 14, 2015"
output: html_document
---

## Load the data
Note: ttestData.csv is missing two sites in the two-way anova data...

```{r}
library(ggplot2)
set.seed(123)
data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40, header = TRUE)
# data_one_way$full.5 ## count per m^2 for full net
# data_one_way$half ## count per m^2 for half net
count <- c(data_one_way$full.5,  ## count for full net
            4 * data_one_way$half.5) ## count for half net adjusted for radius
date <- rep(as.factor(data_one_way$month), 2)
net <- as.factor(c(rep("full", 40), rep("half", 40)))
data <- data.frame(count=count, date=date, net=net)
X <- model.matrix(~ net * date)
ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ .)
```

# Write a negative binomial GLM in STAN

```{r, echo=FALSE}
library(rstan, verbose=FALSE)
# model <- "
# // Negative Binomial GLM
# // see Gelman and Hill (2007)
# data {
#   int<lower=1> N;       // number of observations
#   int<lower=0> p;
#   matrix[N, p] X;       // predictor
#   int<lower=0> y[N];    // response
# } transformed data {
#   vector[p] J;
#   matrix[p, p] I_p;
#   // calculate placeholder values
#   J <- rep_vector(0, p);
#   I_p <- diag_matrix(rep_vector(1, p));
# }
# parameters {
#   real<lower=0> phi;   // Negative Binomial dispersion parameter
#   vector[p] beta;      // Negative Binomial mean parameters
# } model {
#   // priors:
#   phi ~ cauchy(0, 20);
#   beta ~ multi_normal(J, 100 * I_p);
#   // data model:
#   y ~ neg_binomial_2_log(X * beta, phi);
# }
# "
# write(model, file = "~/mysis/stanModels/negbin-glm.stan")
# sm <- stan_model("~/mysis/stanModels/negbin-glm.stan")
sm <- stan("~/mysis/stanModels/negbin-glm.stan", 
           data=list(N = length(count), y = count, X = X, p=dim(X)[2]),
           chains=0)
```
## Fit the Model

```{r, echo=FALSE}
library(parallel)
estimate_params <- function(model, data, pars = pars, iter=5000){
  sflist <- mclapply(1:4, mc.cores = 4,
             function(i) stan(fit = model, data = data, pars = pars,
                              chains = 1, chain_id = i,
                              iter=iter,
                              refresh = -1))
  fit <- sflist2stanfit(sflist)
  return(fit)
}
# m <- sampling(sm, data = list(N = length(count), y = count, X = X, p=dim(X)[2]),
#   pars = c("beta", "phi"),
#   iter = 5000, chains = 4)
m <- estimate_params(sm, data = list(N = length(count), y = count, X = X, p=dim(X)[2]),  pars = c("beta", "phi"))
```

## Model output
```{r}
m
```

## Plot traceplots
```{r}
rstan::traceplot(m, inc_warmup = FALSE)
```

## Plot fitted vs. simulated values

```{r}
e <- rstan::extract(m, pars = c("beta", "phi"))
par(mfrow = c(2, 3))
for(i in 1:2) {
  if(i == 1) {
#     for(j in 1:6) {
    for(j in 1:5) {
      plot(density(e[[i]][,j]), main = paste("beta", j, sep=""))
    }
  } else if(i == 2) {
    plot(density(e[[i]]), main = "phi")
  }
}
```

```{r}
library(ggmcmc)
library(coda)
s <-  mcmc.list(lapply(1:ncol(m), function(x) mcmc(as.array(m)[,x,])))
S <- ggs(s)

ggs_caterpillar(S, family = "beta") 
```

```{r}
library(Gmisc)
tableData <- cbind(summary(s)$statistics[1:7, 1], summary(s)$quantiles[1:7, c(1, 5)])
htmlTable(x=round(tableData, digits=2),
          caption=paste("Regression parameter estimates and CI's"),
          header = c("Estimate", "Lower 95% CI", "Upper 95% CI"),
          rowlabel = "Variables",
          rnames = c("Intercept",
                    "Net Size",
                    "August",
                    "September",
                    "Net Size by August",
                    "Net Size by September",
                    "Overdispersion Parameter"))
```

### 2015-08-17 negativeBinomial/negativeBinomialExample.Rmd
update neg binomial model and added real count data example

@@
 "
-write(model, file = "~/mysis/stanModels/overdispersedPoisson-glm.stan")
-sm <- stan_model("~/mysis/stanModels/overdispersedPoisson-glm.stan")
+write(model, file = "~/mysis/stanModels/negbin-glm.stan")
+sm <- stan_model("~/mysis/stanModels/negbin-glm.stan")
 ```

### 2015-08-24 countModel/countData.Rmd
updated simulated and real data models

@@
 data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40, header = TRUE)
-# data_one_way$full.5 ## count per m^2 for full net
+# data_one_way$full ## count per m^2 for full net
 # data_one_way$half ## count per m^2 for half net

### 2015-08-24 sexRatioModel/sexRatioModel.Rmd
updated simulated and real data models

---
title: "Beta Regression"
author: "John Tipton"
date: "August 14, 2015"
output: html_document
---

## Note, there are propotions missing and proportions greater than 1

# Write a beta GLM in STAN

```{r}
library(rstan)
# model <- "
# // beta Regression
# data {
#   int<lower=1> N_obs;                   // sample size
#   int<lower=1> N_mis;                   // sample size
#   int<lower=1> p;                       // p predictors
#   vector<lower=0,upper=1>[N_obs] y_obs; // response 
#   matrix[N_obs, p] X_obs;               // predictor matrix
#   matrix[N_mis, p] X_mis;               // predictor matrix
# }
# parameters {
#   vector<lower=0,upper=1>[N_mis] y_mis; // missing response 
#   vector[p] theta;                      // reg coefficients
#   real<lower=0> phi;                    // dispersion parameter
# }
# transformed parameters{
#   vector[p] beta;
#   beta <- theta * 10;               // matt trick if desired
# }
# model {
#   // model calculations
#   vector[N_obs] Xbeta_obs;              // linear predictor
#   vector[N_mis] Xbeta_mis;              // linear predictor
#   vector<lower=0,upper=1>[N_obs] mu_obs;// transformed linear predictor
#   vector<lower=0,upper=1>[N_mis] mu_mis;// transformed linear predictor
#   vector<lower=0>[N_obs] A_obs;         // parameter for beta distn
#   vector<lower=0>[N_mis] A_mis;         // parameter for beta distn
#   vector<lower=0>[N_obs] B_obs;         // parameter for beta distn
#   vector<lower=0>[N_mis] B_mis;         // parameter for beta distn
#   Xbeta_obs <- X_obs * beta;
#   Xbeta_mis <- X_mis * beta;
#   for (i in 1:N_obs) { 
#     mu_obs[i] <- inv_logit(Xbeta_obs[i]);   
#   }
#   for (i in 1:N_mis) { 
#     mu_mis[i] <- inv_logit(Xbeta_mis[i]);   
#   }
#   A_obs <- mu_obs * phi;
#   A_mis <- mu_mis * phi;
#   B_obs<- (1.0 - mu_obs) * phi;
#   B_mis<- (1.0 - mu_mis) * phi;
#   // priors
#   theta ~ normal(0, 1);   
#   phi ~ cauchy(0, 5);               // different options for phi  
#   //phi ~ inv_gamma(.001, .001);
#   //phi ~ uniform(0, 500);          // put upper on phi if using this
#   // likelihood
#   y_obs ~ beta(A_obs, B_obs);
#   y_mis ~ beta(A_mis, B_mis);
# }
# "
# write(model, file = "~/mysis/stanModels/betaRegressionMissing-glm.stan")
sm <- stan_model("~/mysis/stanModels/betaRegression-glm.stan")
```

## Load the Data

```{r}
library(ggplot2)
set.seed(123)

fullData <- read.csv("~/mysis/data/lengthData.csv", skip=4)
substr(as.character(fullData$Gender), 0, 1)
fullData$Station

males <- rep(0, 80)
females <- rep(0, 80)
unknowns <- rep(0, 80)
size <- rep(0, 80)
date_tmp <- rep(0, 80)

idx <- 1
for(i in unique(fullData$Sample.Label)){
  tmp_data = subset(fullData, fullData$Sample.Label == i)
  tmp = substr(as.character(tmp_data$Gender), 0, 1)
  males[idx] <- sum(tmp == "M")
  females[idx] <- sum(tmp == "F")
  unknowns[idx] <- sum(tmp == "U")
  size[idx] <- tmp_data$Size..m.[1]
  date_tmp[idx] <- tmp_data$Date[1]
  idx <- idx + 1
}
prop <- females / (females + males)
prop_corrected <- (females + 0.5 * unknowns) / 
                  (females + males + unknowns)
prop[prop == 0] <- 0.01
prop_corrected[prop_corrected == 0] <- 0.01
## Correct for mislabeling
date_tmp[70] <- 4
date_tmp[72] <- 4
date <- rep(0, 80)
date[date_tmp == 1] <- 1
date[date_tmp == 2] <- 1
date[date_tmp == 3] <- 2
date[date_tmp == 4] <- 3

# data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40, header = TRUE, stringsAsFactors=FALSE)

# ratio <- c(as.numeric(data_one_way$full.3),    ## proportion of females full net
#              as.numeric(data_one_way$half.3)) ## proportion of females half net
# N <- length(ratio)

# data_one_way$Sample.Date
## Set improper values to NA

date <- as.factor(date)
net <- as.factor(size)
data <- data.frame(prop=prop, date=date, net=net)
X <- model.matrix(~ net * date)
ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .)

# date <- rep(as.factor(data_one_way$month), 2)
# net <- as.factor(c(rep("full", 40), rep("half", 40)))
# data <- data.frame(count=count, date=date, net=net)
# X <- model.matrix(~ net * date)
# ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .)

```

## Fit the Model

```{r}
m <- sampling(sm, data = list(N=80, y=prop, X=X, p=dim(X)[2]),
  pars = c("beta", "phi"),
  iter = 5000, chains = 4)

m_corrected <- sampling(sm, data = list(N=80, y=prop_corrected, 
                                        X=X, p=dim(X)[2]),
  pars = c("beta", "phi"),
  iter = 5000, chains = 4)
```

## Model output
```{r}
m
```

## Plot traceplots
```{r}
rstan::traceplot(m, inc_warmup = FALSE)

rstan::traceplot(m_corrected, inc_warmup = FALSE)
```

## Plot fitted vs. simulated values

```{r}
e <- rstan::extract(m, pars = c("beta", "phi"))
library(betareg)
brmod <- betareg(y ~ . -1, data = data.frame(y=prop, X))
summary(brmod)
coefs <- c(brmod$coefficients$mean, brmod$coefficients$precision)
par(mfrow = c(3, 3))
for(i in 1:2) {
  if(i == 1) {
    for(j in 1:6) {
    # for(j in 1:5) {
      plot(density(e[[i]][,j]), main = paste("beta", j, sep=""))
      # abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
      abline(v = coefs[i + j - 1], lwd = 3, col = "red")
    }
  } else if(i == 2) {
    plot(density(e[[i]]), main = paste("phi"))
    # abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
    abline(v = coefs[i + j - 1], lwd = 3, col = "red")
  }
}

e_corrected <- rstan::extract(m_corrected, pars = c("beta", "phi"))
library(betareg)
brmod_corrected <- betareg(y ~ . -1, data = data.frame(y=prop_corrected, X))
summary(brmod_corrected)
coefs_corrected <- c(brmod_corrected$coefficients$mean, brmod_corrected$coefficients$precision)
par(mfrow = c(3, 3))
for(i in 1:2) {
  if(i == 1) {
    for(j in 1:6) {
    # for(j in 1:5) {
      plot(density(e_corrected[[i]][,j]), main = paste("beta", j, sep=""))
      # abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
      abline(v = coefs_corrected[i + j - 1], lwd = 3, col = "red")
    }
  } else if(i == 2) {
    plot(density(e_corrected[[i]]), main = paste("phi"))
    # abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
    abline(v = coefs_corrected[i + j - 1], lwd = 3, col = "red")
  }
}
```

```{r}
library(ggmcmc)
library(coda)
s <-  mcmc.list(lapply(1:ncol(m), function(x) mcmc(as.array(m)[,x,])))
S <- ggs(s)

ggs_caterpillar(S, family = "beta")

s_corrected <-  mcmc.list(lapply(1:ncol(m_corrected), function(x) mcmc(as.array(m_corrected)[,x,])))
S_corrected <- ggs(s_corrected)

ggs_caterpillar(S_corrected, family = "beta") 
```

# model for later comparison

# simulate yRep as in BDA appendix C; could also do in generated quantities block of stan code
for (s in 1:nSim){
  betaS = betareg_sim$beta[s,]
  muS = plogis(X%*%betaS)
  A = muS*betareg_sim$phi[s]
  B = (1-muS)*betareg_sim$phi[s]
  yRep[s,] = rbeta(dat$N, A, B)
}
str(yRep)

# example, as in BDA p.144, code from appendix C; probably not very useful
par(mfrow=c(5,4), mar=c(4,4,2,2))
hist(y, xlab='', main='y', 'FD')
for (s in 1:19) hist(yRep[s,], xlab='', main=paste('yRep',s), 'FD')
layout(1)

# the following plots get at the same thing
library(ggplot2); library(reshape2)
gdat = data.frame(X,y)
gdatyRep = melt(yRep)

# individual observation distibutions; note the theme is my own (available on github), but otherwise you'll have to define yours.
ggplot(aes(x=y), xlim=c(.2,1), data=gdat) + 
  stat_density(geom='line', lwd=2, alpha=.2) + 
  geom_line(aes(x=value,  group=as.factor(Var2), col=as.factor(Var2)), stat='density', data=gdatyRep, show_guide=F, alpha=.15) +
  ggtheme

# whole response distributions, comparison with betareg fits
ggplot(aes(x=y), xlim=c(.2,1), data=gdat) + 
  geom_line(aes(x=value,  group=as.factor(Var1), col=as.factor(Var1)), stat='density', data=gdatyRep, show_guide=F, alpha=.05) +
  stat_density(geom='line', lwd=2, alpha=.2) + 
  stat_density(aes(x=brmod$fitted), geom='line', col='darkred', lwd=2, alpha=.2) +
  stat_density(aes(x=value), geom='line', col='#FF5500', lwd=2, alpha=.2, data=gdatyRep) +
  ggtheme

# Examine quantiles
sapply(list(yRep=gdatyRep$value, brFitted=brmod$fitted.values, y=y), quantile, p=c(0,.1,.25,.5,.625,.75,.8,.9,.95,1))

### 2015-08-31 lengthModel/lengthModel.Rmd
bump

--- 2015-08-24 countModel/countData.Rmd
@@
 ## Load the data
-Note: ttestData.csv is missing two sites in the two-way anova data...
 
@@
 set.seed(123)
-data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40, header = TRUE)
+data_length <- read.csv("~/mysis/data/lengthData.csv", skip = 4, header = TRUE)
+net <- data_length$Size..m.
+length <- data_length$Length..mm.
+gender <- as.factor(substr(as.character(data_length$Gender), 0, 1))
 # data_one_way$full ## count per m^2 for full net

### 2015-08-31 sexRatioModel/sexRatioModelNegBin.Rmd
bump

---
title: "Beta Regression"
author: "John Tipton"
date: "August 14, 2015"
output: html_document
---

# Catch by gender 
Not sure of the story here. There is definitely an effect of gender, but I suspect this is due to non-random misclassification into the unknowns (i.e. the unknowns are more likely to be females (males) than males (females))

# Write a beta GLM in STAN

```{r}
library(rstan)
sm <- stan_model("~/mysis/stanModels/negbin-glm.stan")
```

## Load the Data

```{r}
library(ggplot2)
set.seed(123)

fullData <- read.csv("~/mysis/data/lengthData.csv", skip=4)

males <- rep(0, 80)
females <- rep(0, 80)
juveniles <- rep(0, 80)
unknowns <- rep(0, 80)
size <- rep(0, 80)
date_tmp <- rep(0, 80)

idx <- 1
for(i in unique(fullData$Sample.Label)){
  tmp_data = subset(fullData, fullData$Sample.Label == i)
  tmp = substr(as.character(tmp_data$Gender), 0, 1)
  males[idx] <- sum(tmp == "M")
  juveniles[idx] <- sum(tmp == "J")
  females[idx] <- sum(tmp == "F")
  unknowns[idx] <- sum(tmp == "U")
  size[idx] <- tmp_data$Size..m.[1]
  date_tmp[idx] <- tmp_data$Date[1]
  idx <- idx + 1
}

## construct count vector
count <- c(males, females, juveniles, unknowns)
gender <- factor(rep(1:4, each=80))
## Correct for mislabeling
date_tmp[70] <- 4
date_tmp[72] <- 4
date <- rep(0, 80)
date[date_tmp == 1] <- 1
date[date_tmp == 2] <- 1
date[date_tmp == 3] <- 2
date[date_tmp == 4] <- 3

date <- as.factor(date)
net <- as.factor(size)
count[rep(net, times=4) == 0.5] <- count[rep(net, times=4) == 0.5] * 4
data <- data.frame(count=count, date=rep(date, times=4), net=rep(net, times=4), gender=gender)
# X <- model.matrix(~ net * date + net * gender + gender * date, data=data)
X <- model.matrix(~ net + net * gender + gender * date, data=data)
ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ gender)
ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ date)
ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(gender ~ date)

```

## Fit the Model

```{r}
m <- sampling(sm, data = list(N=320, y=count, X=X, p=dim(X)[2]),
  pars = c("beta", "phi"),
  iter = 5000, chains = 4)
```

## Model output
```{r}
m
```

## Plot traceplots
```{r}
rstan::traceplot(m, inc_warmup = FALSE)
```

## Plot fitted vs. simulated values

```{r}
e <- rstan::extract(m, pars = c("beta", "phi"))
library(MASS)
nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X))
summary(nbmod)
coefs <- c(nbmod$coefficients, nbmod$theta)
par(mfrow = c(3, 3))
for(i in 1:2) {
  if(i == 1){
    for(j in 1:dim(e[[i]])[2]) {
      plot(density(e[[i]][,j]), main = paste("beta", j, sep=""))
      abline(v = coefs[j], lwd = 3, col = "red")
    }
  } else if(i == 2) {
    plot(density(e[[i]]), main = paste("phi"))
    abline(v = coefs[dim(e[[i]])[2] + 1], lwd = 3, col = "red")
  }
}

```

```{r}
library(ggmcmc)
library(coda)
s <-  mcmc.list(lapply(1:ncol(m), function(x) mcmc(as.array(m)[,x,])))
S <- ggs(s)

ggs_caterpillar(S, family = "beta")

```

### 2015-09-03 lengthModel/lengthModel.Rmd
updated length model

--- 2015-08-24 countModel/countData.Rmd
@@
 
-## Load the data
-Note: ttestData.csv is missing two sites in the two-way anova data...
+## Load the data - are these dates errors or are the samples that were found later in the season
 
@@
 library(ggplot2)
+library(stringr)
 set.seed(123)
-data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40, header = TRUE)
-# data_one_way$full ## count per m^2 for full net
-# data_one_way$half ## count per m^2 for half net
-count <- c(data_one_way$full.5,  ## count for full net
-            4 * data_one_way$half.5) ## count for half net adjusted for radius
-date <- rep(as.factor(data_one_way$month), 2)
-net <- as.factor(c(rep("full", 40), rep("half", 40)))
-data <- data.frame(count=count, date=date, net=net)
-X <- model.matrix(~ net * date)
-ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ .)
+data_length <- read.csv("~/mysis/data/lengthData.csv", skip = 4, header = TRUE)
+net <- data_length$Size..m.
+y <- data_length$Length..mm.
+gender <- as.factor(substr(str_trim(as.character(data_length$Gender)), 0, 1))
+                       ## remove whitespace from string, then truncate into 
+                       ## broad gender classes, not subclasses
+
+date <- as.numeric(data_length$Date)
+date[date == 2] <- 1      ## collapse into monthly values
+date[date == 3] <- 2
+date[date == 4] <- 3
+date <- as.factor(date)
+plot(as.numeric(date), type='l', main = "Is this a data error?")
+data <- data.frame(y=y, date=date, net=net, gender=gender)
+X <- model.matrix(~ net + date + gender, data=data)
+ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ date)
 ```
 
-# Write a negative binomial GLM in STAN
+# Write the model
+$t$ = time
+$i$ = net size
+$j$ = age class
+$k$ = replicate
+
+$$
+y_{tijk} \sim \mbox{N}(\mu_{tij}, \sigma^2_{tij})
+$$
 
@@
 library(rstan, verbose=FALSE)
-# model <- "
-# // Negative Binomial GLM
-# // see Gelman and Hill (2007)
-# data {
-#   int<lower=1> N;       // number of observations
-#   int<lower=0> p;
-#   matrix[N, p] X;       // predictor
-#   int<lower=0> y[N];    // response
-# } transformed data {
-#   vector[p] J;
-#   matrix[p, p] I_p;
-#   // calculate placeholder values
-#   J <- rep_vector(0, p);
-#   I_p <- diag_matrix(rep_vector(1, p));
-# }
-# parameters {
-#   real<lower=0> phi;   // Negative Binomial dispersion parameter
-#   vector[p] beta;      // Negative Binomial mean parameters
-# } model {
-#   // priors:
-#   phi ~ cauchy(0, 20);
-#   beta ~ multi_normal(J, 100 * I_p);
-#   // data model:
-#   y ~ neg_binomial_2_log(X * beta, phi);
-# }
-# "
-# write(model, file = "~/mysis/stanModels/negbin-glm.stan")
-# sm <- stan_model("~/mysis/stanModels/negbin-glm.stan")
-sm <- stan("~/mysis/stanModels/negbin-glm.stan", 
-           data=list(N = length(count), y = count, X = X, p=dim(X)[2]),
+model <- "
+// regression
+data {
+  int<lower=1> N;       // number of observations
+  int<lower=1> p;       // number of predictors
+  matrix[N, p] X;       // predictor
+  vector[N] y;          // response
+} parameters {
+  real<lower=0> s;      // regression standard deviation
+  vector[p] beta;       // mean parameters
+} model {
+  // priors:
+  //s ~ cauchy(0, 20);
+  // beta ~ multi_normal(0, 100);
+  // data model:
+  y ~ normal(X * beta, s);
+}
+"
+write(model, file = "~/mysis/stanModels/length.stan")
+# sm <- stan_model("~/mysis/stanModels/length.stan")
+sm <- stan("~/mysis/stanModels/length.stan", 
+           data=list(N=length(y), p=dim(X)[2], y=y, X=X),
            chains=0)
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, cach=TRUE}
 library(parallel)
@@
 }
-# m <- sampling(sm, data = list(N = length(count), y = count, X = X, p=dim(X)[2]),
-#   pars = c("beta", "phi"),
-#   iter = 5000, chains = 4)
-m <- estimate_params(sm, data = list(N = length(count), y = count, X = X, p=dim(X)[2]),  pars = c("beta", "phi"))
+# m <- sampling(sm, data = list(N=length(y), p=dim(X)[2], y=y, X=X),
+#   pars = c("beta", "s"),
+#   iter = 1000, chains = 4)
+m <- estimate_params(sm, data = list(N=length(y), p=dim(X)[2], y=y, X=X),
+                     pars = c("beta", "s"))
 ```
@@
 ```{r}
-e <- rstan::extract(m, pars = c("beta", "phi"))
-par(mfrow = c(2, 3))
+e <- rstan::extract(m, pars = c("beta", "s"))
+par(mfrow = c(3, 3))
 for(i in 1:2) {
   if(i == 1) {
-#     for(j in 1:6) {
-    for(j in 1:5) {
+    for(j in 1:dim(e[[i]])[2]) {
       plot(density(e[[i]][,j]), main = paste("beta", j, sep=""))
@@
   } else if(i == 2) {
-    plot(density(e[[i]]), main = "phi")
+    plot(density(e[[i]]), main = "s")
   }
@@
 library(Gmisc)
-tableData <- cbind(summary(s)$statistics[1:7, 1], summary(s)$quantiles[1:7, c(1, 5)])
+tableData <- cbind(summary(s)$statistics[1:8, 1], summary(s)$quantiles[1:8, c(1, 5)])
 htmlTable(x=round(tableData, digits=2),
@@
                     "September",
-                    "Net Size by August",
-                    "Net Size by September",
-                    "Overdispersion Parameter"))
+                    "Juvenile",
+                    "Male",
+                    "Unknown",
+                    "standard deviation"))
 ```

### 2015-09-03 lengthModel/lengthModel.Rmd
bump

@@
 date <- as.factor(date)
-plot(as.numeric(date), type='l', main = "Is this a data error?")
+plot(as.numeric(date), type='l', main = "Is this a data error? - yes?")
+## correct date mislabeling
+date <- as.numeric(data_length$Date)
+date[7330:7519] <- 4
+date[7559:7773] <- 4
+date[date == 2] <- 1      ## collapse into monthly values
+date[date == 3] <- 2
+date[date == 4] <- 3
+date <- as.factor(date)
+plot(as.numeric(date), type='l', main = "Data looks better")
 data <- data.frame(y=y, date=date, net=net, gender=gender)
@@
   vector[p] beta;             // mean parameters
+  vector<lower=0>[p] gamma2;  // LASSO mixing parameters
+  real<lower=0> lambda2;      // LASSO shrinkage parameters
 } model {
@@
   //s ~ cauchy(0, 20);
-  // beta ~ multi_normal(0, 100);
+  lambda2 ~ gamma(1, 1);      // Prior for weak LASSO shrinkage 
+  gamma2 ~ exponential(lambda2 / 2.0);
+  for(i in 2:p){
+    beta[i] ~ normal(0, s * sqrt(gamma2[i]));
+  }
+
   // data model:
@@
 
-```{r, echo=FALSE, cach=TRUE}
+```{r, echo=FALSE, cache=TRUE}
 library(parallel)
@@
 m <- estimate_params(sm, data = list(N=length(y), p=dim(X)[2], y=y, X=X),
-                     pars = c("beta", "s"))
+                     pars = c("beta", "s", "gamma2", "lambda2"))
 ```
@@
 library(Gmisc)
-tableData <- cbind(summary(s)$statistics[1:8, 1], summary(s)$quantiles[1:8, c(1, 5)])
+tableData <- cbind(summary(s)$statistics[1:(e[[i]])[2] + 1), 1], summary(s)$quantiles[1:(e[[i]])[2] + 1), c(1, 5)])
 htmlTable(x=round(tableData, digits=2),

### 2015-09-04 lengthModel/lengthModel.Rmd
updated length model

@@
 library(Gmisc)
-tableData <- cbind(summary(s)$statistics[1:(e[[i]])[2] + 1), 1], summary(s)$quantiles[1:(e[[i]])[2] + 1), c(1, 5)])
+tableData <- cbind(summary(s)$statistics[1:(dim(e[[1]])[2] + 1), 1],
+                   summary(s)$quantiles[1:(dim(e[[1]])[2] + 1), c(1, 5)])
 htmlTable(x=round(tableData, digits=2),

### 2015-09-04 lengthModel/lengthModelHeterogeneous.Rmd
update

--- 2015-09-04 lengthModel/lengthModel.Rmd
@@
 library(stringr)
+library(mgcv)
 set.seed(123)
@@
 ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ date)
+idx <- attr(uniquecombs(X), "index")
 ```
@@
   int<lower=1> p;             // number of predictors
+  int<lower=1> q;             // number of levels
   matrix[N, p] X;             // predictor
   vector[N] y;                // response
+  int idx[N];                 // indicator variable of treatment
 } parameters {  
-  real<lower=0> s;            // regression standard deviation
+  vector<lower=0>[q] s;       // heterogeneous regression standard deviation
   vector[p] beta;             // mean parameters
-  vector<lower=0>[p] gamma2;  // LASSO mixing parameters
-  real<lower=0> lambda2;      // LASSO shrinkage parameters
 } model {
@@
   //s ~ cauchy(0, 20);
-  lambda2 ~ gamma(1, 1);      // Prior for weak LASSO shrinkage 
-  gamma2 ~ exponential(lambda2 / 2.0);
   for(i in 2:p){
-    beta[i] ~ normal(0, s * sqrt(gamma2[i]));
+    beta[i] ~ normal(0, 100);
   }
@@
   // data model:
-  y ~ normal(X * beta, s);
+  for(i in 1:N){
+    y[i] ~ normal(X[i] * beta, s[idx[i]]);
+  }
 }
 "
-write(model, file = "~/mysis/stanModels/length.stan")
-# sm <- stan_model("~/mysis/stanModels/length.stan")
-sm <- stan("~/mysis/stanModels/length.stan", 
-           data=list(N=length(y), p=dim(X)[2], y=y, X=X),
+write(model, file = "~/mysis/stanModels/lengthHeterogeneous.stan")
+# sm <- stan_model("~/mysis/stanModels/lengthHeterogeneous.stan")
+sm <- stan("~/mysis/stanModels/lengthHeterogeneous.stan", 
+           data=list(N=length(y), p=dim(X)[2], y=y, X=X, idx=idx, q=length(unique(idx))),
            chains=0)
@@
 }
-# m <- sampling(sm, data = list(N=length(y), p=dim(X)[2], y=y, X=X),
+# m <- sampling(sm, data = list(N=length(y), p=dim(X)[2], y=y, X=X, idx=idx, q=length(unique(idx))),
 #   pars = c("beta", "s"),
-#   iter = 1000, chains = 4)
-m <- estimate_params(sm, data = list(N=length(y), p=dim(X)[2], y=y, X=X),
-                     pars = c("beta", "s", "gamma2", "lambda2"))
+#   iter = 1000, chains = 1)
+m <- estimate_params(sm, data = list(N=length(y), p=dim(X)[2], y=y, X=X, idx=idx, q=length(unique(idx))),
+                     pars = c("beta", "s"))
 ```
@@
 ```
+
+```{r}
+summary(lm(y ~ net + date + gender))
+plot(lm(y ~ net + date + gender))
+library(MASS)
+summary(rlm(y ~ net + date + gender))
+plot(rlm(y ~ net + date + gender))
+```

### 2015-09-09 lengthModel/lengthModelHeterogeneous.Rmd
started results writeup

@@
 date <- as.factor(date)
+station <- data_length$Station
 plot(as.numeric(date), type='l', main = "Is this a data error? - yes?")

### 2015-09-09 results/mysisResults.Rmd
started results writeup

---
title: "Mysis analysis"
author: "John Tipton"
date: "September 9, 2015"
output: html_document
---
```{r loadLibraries, echo=FALSE, include=FALSE, message=FALSE}
library(knitr)
library(ggplot2)
library(Gmisc)
```
## Model for Total counts for each net
```{r totalCount, echo=FALSE, include=FALSE, message=FALSE}
set.seed(123)
data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40, header = TRUE)
count <- c(data_one_way$full.5,  ## count for full net
            4 * data_one_way$half.5) ## count for half net adjusted for radius
date <- rep(as.factor(data_one_way$month), 2)
station <- rep(data_one_way$Station, 2)
net <- as.factor(c(rep("full", 40), rep("half", 40)))
data <- data.frame(count=count, date=date, net=net)
X <- model.matrix(~ net * date)
ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ .)
library(MASS)
summary(nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X)))
summary(qpmod <- glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(2.6)))
summary(glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(0.1)))
summary(glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(10)))
coefs <- c(nbmod$coefficients, nbmod$theta)
```

To test for a change in total counts between the two net sizes, we construct a negative binomal model that examines the effects of net size, date of sampling, the interaction between net size and sampling data while including a random effect for station. The results shown below show that there is no effect on the number of counts observed between the two net sizes after adjusting for the size of the nets.

```{r countModelResults, echo=FALSE}
tableData <- cbind(summary(nbmod)$coeff[1:6, ])
colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
                         "Net Size * August", "Net Size * September")
# htmlTable(x=round(tableData, digits=2),
#           caption=paste("Negative Binomial Regression parameter estimates"),
#           header = c("  Estimate  ", "  Std. Error  ", "  z value  ", "  p-value  "),
#           rowlabel = "Variable",
#           rnames = c("Intercept",
#                     "Net Size",
#                     "August",
#                     "September",
#                     "Net Size * August",
#                     "Net Size * September"))
kable(tableData, digits = 4, format="pandoc")
```

### 2015-09-10 lengthModel/lengthModelT.Rmd
updated results

--- 2015-09-04 lengthModel/lengthModel.Rmd
@@
 library(stringr)
+library(parallel)
+library(rstan, verbose=FALSE)
 set.seed(123)
@@
 ```{r, echo=FALSE}
-library(rstan, verbose=FALSE)
 model <- "
-// regression
+// student t regression
 data {
@@
   real<lower=0> s;            // regression standard deviation
+  real<lower=0> nu;           // student t degrees of freedom
   vector[p] beta;             // mean parameters
@@
   // data model:
-  y ~ normal(X * beta, s);
+#   for(i in 1:N){
+#     y[i] ~ student_t(nu, X[i] * beta, s);
+#   }
+  y ~ student_t(nu, X * beta, s);
 }
 "
-write(model, file = "~/mysis/stanModels/length.stan")
-# sm <- stan_model("~/mysis/stanModels/length.stan")
-sm <- stan("~/mysis/stanModels/length.stan", 
+write(model, file = "~/mysis/stanModels/lengthT.stan")
+# sm <- stan_model("~/mysis/stanModels/lengthT.stan")
+sm <- stan("~/mysis/stanModels/lengthT.stan", 
            data=list(N=length(y), p=dim(X)[2], y=y, X=X),
@@
 ```{r, echo=FALSE, cache=TRUE}
-library(parallel)
-estimate_params <- function(model, data, pars = pars, iter=5000){
+estimate_params <- function(model, data, pars = pars, iter=1000){
   sflist <- mclapply(1:4, mc.cores = 4,
@@
 # m <- sampling(sm, data = list(N=length(y), p=dim(X)[2], y=y, X=X),
-#   pars = c("beta", "s"),
-#   iter = 1000, chains = 4)
+#   pars = c("beta", "s"), iter = 100, chains = 1)
 m <- estimate_params(sm, data = list(N=length(y), p=dim(X)[2], y=y, X=X),
-                     pars = c("beta", "s", "gamma2", "lambda2"))
+                     pars = c("beta", "s", "gamma2", "lambda2", "nu"))
 ```

### 2015-09-10 results/mysisResults.Rmd
updated results

@@
 library(Gmisc)
+library(MASS)
+library(stringr)
+library(mgcv)
+library(lme4)
 ```
 ## Model for Total counts for each net
-```{r totalCount, echo=FALSE, include=FALSE, message=FALSE}
+```{r totalCount, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE}
 set.seed(123)
@@
 ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ .)
-library(MASS)
 summary(nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X)))
-summary(qpmod <- glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(2.6)))
-summary(glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(0.1)))
-summary(glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(10)))
+# summary(qpmod <- glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(2.6)))
+# summary(glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(0.1))) 
+# summary(glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(10)))
 coefs <- c(nbmod$coefficients, nbmod$theta)
@@
 
-To test for a change in total counts between the two net sizes, we construct a negative binomal model that examines the effects of net size, date of sampling, the interaction between net size and sampling data while including a random effect for station.
-The results shown below show that there is no effect on the number of counts observed between the two net sizes after adjusting for the size of the nets.
+To test for a change in total counts between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
+The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
 
-```{r countModelResults, echo=FALSE}
+```{r countModelResults, echo=FALSE, cache=TRUE}
 tableData <- cbind(summary(nbmod)$coeff[1:6, ])
@@
 ```
+
+## Model for counts by gender
+```{r, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE}
+set.seed(123)
+fullData <- read.csv("~/mysis/data/lengthData.csv", skip=4)
+
+males <- rep(0, 80)
+females <- rep(0, 80)
+juveniles <- rep(0, 80)
+unknowns <- rep(0, 80)
+size <- rep(0, 80)
+date_tmp <- rep(0, 80)
+
+idx <- 1
+for(i in unique(fullData$Sample.Label)){
+  tmp_data = subset(fullData, fullData$Sample.Label == i)
+  tmp = substr(as.character(tmp_data$Gender), 0, 1)
+  males[idx] <- sum(tmp == "M")
+  juveniles[idx] <- sum(tmp == "J")
+  females[idx] <- sum(tmp == "F")
+  unknowns[idx] <- sum(tmp == "U")
+  size[idx] <- tmp_data$Size..m.[1]
+  date_tmp[idx] <- tmp_data$Date[1]
+  idx <- idx + 1
+}
+
+## construct count vector
+count <- c(males, females, juveniles, unknowns)
+gender <- factor(rep(1:4, each=80))
+## Correct for mislabeling
+date_tmp[70] <- 4
+date_tmp[72] <- 4
+date <- rep(0, 80)
+date[date_tmp == 1] <- 1
+date[date_tmp == 2] <- 1
+date[date_tmp == 3] <- 2
+date[date_tmp == 4] <- 3
+
+date <- as.factor(date)
+net <- as.factor(size)
+count[rep(net, times=4) == 0.5] <- count[rep(net, times=4) == 0.5] * 4
+data <- data.frame(count=count, date=rep(date, times=4), net=rep(net, times=4), gender=gender)
+# X <- model.matrix(~ net * date + net * gender + gender * date, data=data)
+X <- model.matrix(~ net * gender + date, data=data)
+ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ gender)
+ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ date)
+ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(gender ~ date)
+nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X))
+```
+
+To test for a change in counts broken down by gender category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
+The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
+
+```{r sexResults, echo=FALSE, cache=TRUE}
+
+tableData <- cbind(summary(nbmod)$coeff)
+colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
+rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
+                         "Unknown Gender", "August", "September", 
+                         "Net Size * Females", "Net Size * Juveniles", 
+                         "Net Size * Unknowns")
+
+kable(tableData, digits = 4, format="pandoc")
+```
+
+## Model for proportion of juveniles
+```{r, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE}
+set.seed(123)
+data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40,
+                         header = TRUE, stringsAsFactors=FALSE)
+
+prop <- c(as.numeric(data_one_way$full.2),    ## proportion of females full net
+             as.numeric(data_one_way$half.2)) ## proportion of females half net
+N <- length(prop)
+## Set improper values to NA
+
+date <- rep(as.factor(data_one_way$month), 2)
+net <- as.factor(c(rep("full", 40), rep("half", 40)))
+data <- data.frame(prop=prop, date=date, net=net)
+X <- model.matrix(~ net * date)
+ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .)
+library(betareg)
+brmod <- betareg(prop ~ . -1, data = data.frame(prop, X))
+summary(brmod)
+```
+
+To test for a change in the proportion of juveniles caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
+The results shown below show that there is no evidence of an effect on the proportion of juveniles caught between the two net sizes (p=`r I(round(summary(brmod)$coef$mean[2, 4], 4))`).
+
+```{r juvenileResults, echo=FALSE, cache=TRUE}
+tableData <- cbind(summary(brmod)$coeff$mean)
+colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
+rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
+                         "Net Size * August", "Net Size * September")
+kable(tableData, digits = 4, format="pandoc")
+```
+
+## Length model
+
+```{r, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE}
+set.seed(123)
+data_length <- read.csv("~/mysis/data/lengthData.csv", skip = 4, header = TRUE)
+net <- data_length$Size..m.
+y <- data_length$Length..mm.
+gender <- as.factor(substr(str_trim(as.character(data_length$Gender)), 0, 1))
+                       ## remove whitespace from string, then truncate into 
+                       ## broad gender classes, not subclasses
+
+date <- as.numeric(data_length$Date)
+date[date == 2] <- 1      ## collapse into monthly values
+date[date == 3] <- 2
+date[date == 4] <- 3
+date <- as.factor(date)
+station <- data_length$Station
+# plot(as.numeric(date), type='l', main = "Is this a data error? - yes?")
+## correct date mislabeling
+date <- as.numeric(data_length$Date)
+date[7330:7519] <- 4
+date[7559:7773] <- 4
+date[date == 2] <- 1      ## collapse into monthly values
+date[date == 3] <- 2
+date[date == 4] <- 3
+date <- as.factor(date)
+# plot(as.numeric(date), type='l', main = "Data looks better")
+data <- data.frame(y=y, date=date, net=net, gender=gender)
+X <- model.matrix(~ net + date + gender, data=data)
+ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ date)
+idx <- attr(uniquecombs(X), "index")
+summary(lmmod <- lm(y ~ net + date + gender))
+library(lme4)
+summary(lmemod <- lmer(y ~ net + date + gender + (1|station)))
+rlmmod <- rlm(y ~ net + date + gender)
+```
+
+```{r lengthResults, echo=FALSE, cache=TRUE}
+tableData <- cbind(summary(lmmod)$coeff)
+colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
+rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
+                         "Juvenile", "Male", "Unknown Gender")
+kable(tableData, digits = 4, format="pandoc")
+
+# plot(lmmod)
+# summary(rlmmod)
+# plot(rlmmod)
+# kruskal.test(y ~ net)
+```

### 2015-09-21 results/mysisResults.Rmd
updated results

@@
 ```{r loadLibraries, echo=FALSE, include=FALSE, message=FALSE}
+options(digits=3)
 library(knitr)
@@
 # X <- model.matrix(~ net * date + net * gender + gender * date, data=data)
-X <- model.matrix(~ net * gender + date, data=data)
+X <- model.matrix(~ net * gender + net * date + gender * date, data=data)
 ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ gender)
@@
 
-To test for a change in counts broken down by gender category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
-The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
-
+To test for a change in counts broken down by gender category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling, the interaction between net size and sampling date, and the interaction between sampling date and gender.
+The results shown below show that there is no evidence of an interaction between net size and sampling date and no evidence of an interaction between net size and gender.
 ```{r sexResults, echo=FALSE, cache=TRUE}
-
 tableData <- cbind(summary(nbmod)$coeff)
@@
                          "Net Size * Females", "Net Size * Juveniles", 
-                         "Net Size * Unknowns")
+                         "Net Size * Unknowns", "Net Size * August",
+                         "Net Size * September",
+                         "Females * August", "Juveniles * August", 
+                         "Unknown Gender * August",
+                         "Females * September", "Juveniles * September", 
+                         "Unknown Gender * September")
+kable(tableData, digits = 4, format="pandoc")
+```
 
+```{r, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE}
+X <- model.matrix(~ net * gender + gender * date, data=data)
+nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X))
+```
+Thus, we refit the model without the interaction between net size and sampling date.
+On this reduced model, the results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
+```{r sexResultsReduced, echo=FALSE, cache=TRUE}
+tableData <- cbind(summary(nbmod)$coeff)
+colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
+rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
+                         "Unknown Gender", "August", "September", 
+                         "Net Size * Females", "Net Size * Juveniles", 
+                         "Net Size * Unknowns", 
+                         "Females * August", "Juveniles * August", 
+                         "Unknown Gender * August",
+                         "Females * September", "Juveniles * September", 
+                         "Unknown Gender * September")
 kable(tableData, digits = 4, format="pandoc")
@@
 
-date <- as.numeric(data_length$Date)
-date[date == 2] <- 1      ## collapse into monthly values
-date[date == 3] <- 2
-date[date == 4] <- 3
-date <- as.factor(date)
+# date <- as.numeric(data_length$Date)
+# date[date == 2] <- 1      ## collapse into monthly values
+# date[date == 3] <- 2
+# date[date == 4] <- 3
+# date <- as.factor(date)
 station <- data_length$Station
@@
 # plot(as.numeric(date), type='l', main = "Data looks better")
-data <- data.frame(y=y, date=date, net=net, gender=gender)
+data <- data.frame(y=y, date=date, net=as.factor(net), gender=gender)
 X <- model.matrix(~ net + date + gender, data=data)
+levels(data$date) <- c("July", "August", "Septempber")
+levels(data$net) <- c("Small Net", "Large Net")
+levels(data$gender) <- c("Female", "Juvenile", "Male", "Unknown")
 ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ date) 
+ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ gender)
+ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(date ~ gender)
 idx <- attr(uniquecombs(X), "index")
-summary(lmmod <- lm(y ~ net + date + gender))
+summary(lmmod <- lm(y ~ net + date * gender))
 library(lme4)
-summary(lmemod <- lmer(y ~ net + date + gender + (1|station)))
-rlmmod <- rlm(y ~ net + date + gender)
+# summary(lmemod <- lmer(y ~ net * date + net * gender + date * gender + (1|station)))
+# summary(lmemod <- lmer(y ~ net + date * gender + (1|station)))
+# summary(rlmmod <- rlm(y ~ net * date + net * gender + date * gender))
+summary(rlmmod <- rlm(y ~ net + date * gender))
+tvalues <- summary(rlmmod)$coefficients[, 3]
+df <- summary(rlmmod)$df[2]
+pvalues <- pt(abs(tvalues), df, lower.tail=FALSE) * 2
 ```
@@
 ```{r lengthResults, echo=FALSE, cache=TRUE}
-tableData <- cbind(summary(lmmod)$coeff)
+tableData <- cbind(summary(rlmmod)$coeff, pvalues)
 colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
-                         "Juvenile", "Male", "Unknown Gender")
+rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
+                         "Unknown Gender", "August", "September", 
+                         "Females * August", "Juveniles * August", 
+                         "Unknown Gender * August",
+                         "Females * September", "Juveniles * September", 
+                         "Unknown Gender * September")
+# rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
+#                          "Unknown Gender", "August", "September", 
+#                          "Net Size * Females", "Net Size * Juveniles", 
+#                          "Net Size * Unknowns", "Net Size * August",
+#                          "Net Size * September",
+#                          "Females * August", "Juveniles * August", 
+#                          "Unknown Gender * August",
+#                          "Females * September", "Juveniles * September", 
+#                          "Unknown Gender * September")
 kable(tableData, digits = 4, format="pandoc")
@@
 # plot(rlmmod)
-# kruskal.test(y ~ net)
+# kruskal.test(y ~ net + date + gender)
 ```
+
+```{r effect size, echo=FALSE, eval=FALSE}
+SD <- sqrt(((sum(net == 1.0) - 1) * sd(y[net == 1.0])^2 + 
+              (sum(net == 0.5) - 1) * sd(y[net == 0.5])^2) / 
+             (sum(net == 1.0) + sum(net == 0.5)))
+
+d <- (mean(y[net == 1.0]) - mean(y[net == 0.5])) / SD
+```
+We see that there is a significant effect of net size on mean length caught (p = `r I(pvalues[2])`) after controlling for date, gender class, and an interaction between date and gender class.
+Although this is statisticallly significant, the effect size is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(y))`).
+given a large sample size, a hypothesis test will show staistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05).
+Therefore, the practical effect of a difference in mean length of `r I(summary(rlmmod)$coefficients[2, 1])`mm on a species with a mean length of `r I(mean(y))`mm is small (this is the smallest effect of all the effects estimated by almost a factor of two) and a difference in means of this size is not of practical interest.

### 2015-09-23 results/mysisResults.Rmd
updated results

@@
 data <- data.frame(count=count, date=date, net=net)
+levels(data$date) <- c("July", "August", "Septempber")
+levels(data$net) <- c("Small Net", "Large Net")
 X <- model.matrix(~ net * date)
-ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ .)
 summary(nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X)))
@@
 coefs <- c(nbmod$coefficients, nbmod$theta)
+```
+
+```{r plotCount, echo=FALSE, warning=FALSE, message=FALSE}
+ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ .) + 
+  ggtitle("Histogram of normalized counts by net size")
+ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Histogram of normalized counts by net size and date")
 ```
@@
 data <- data.frame(count=count, date=rep(date, times=4), net=rep(net, times=4), gender=gender)
+levels(data$date) <- c("July", "August", "Septempber")
+levels(data$net) <- c("Small Net", "Large Net")
+levels(data$gender) <- c("Female", "Juvenile", "Male", "Unknown")
 # X <- model.matrix(~ net * date + net * gender + gender * date, data=data)
 X <- model.matrix(~ net * gender + net * date + gender * date, data=data)
-ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ gender)
-ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ date)
-ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(gender ~ date)
+
 nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X))
+```
+
+```{r plot SexCount, echo=FALSE, warning=FALSE, message=FALSE}
+ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ gender) + 
+  ggtitle("Histogram of normalized counts by sex and net size")
+# ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
+#   ggtitle("Histogram of normalized counts by net size")
+ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(gender ~ date) + 
+  ggtitle("Histogram of normalized counts by sex and date")
 ```
@@
 data <- data.frame(prop=prop, date=date, net=net)
+levels(data$date) <- c("July", "August", "Septempber")
+levels(data$net) <- c("Small Net", "Large Net")
 X <- model.matrix(~ net * date)
-ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .)
 library(betareg)
@@
 summary(brmod)
+```
+
+```{r plotJuviCount, echo=FALSE, warning=FALSE, message=FALSE}
+ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .) + 
+  ggtitle("Histogram of proportion of juveniles by net size")
 ```
@@
 levels(data$gender) <- c("Female", "Juvenile", "Male", "Unknown")
-ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ date) 
-ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ gender)
-ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(date ~ gender)
 idx <- attr(uniquecombs(X), "index")
@@
 pvalues <- pt(abs(tvalues), df, lower.tail=FALSE) * 2
+```
+
+```{r plotLength, echo=FALSE, warning=FALSE, message=FALSE}
+ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Histogram of lengths by net size")
+ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ gender) + 
+  ggtitle("Histogram of lengths by sex and net size")
+ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(date ~ gender) + 
+  ggtitle("Histogram of lengths by sex and date")
 ```

### 2015-10-12 results/mysisResults.Rmd
updated writeup

--- 2015-09-10 results/mysisResults.Rmd
@@
 ---
-```{r loadLibraries, echo=FALSE, include=FALSE, message=FALSE}
-library(knitr)
-library(ggplot2)
-library(Gmisc)
-library(MASS)
-library(stringr)
-library(mgcv)
-library(lme4)
+This document shows the analyses performed in Johnson et. al. We start by loading packages and define helper functions
+
+```{r loadLibraries, echo=TRUE, include=FALSE, message=FALSE}
+options(digits=3)
+set.seed(123)
+require(knitr)
+require(ggplot2)
+require(Gmisc)
+require(MASS)
+require(stringr)
+require(mgcv)
+require(lme4)
+
 ```
+
 ## Model for Total counts for each net
-```{r totalCount, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE}
-set.seed(123)
-data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40, header = TRUE)
-count <- c(data_one_way$full.5,  ## count for full net
-            4 * data_one_way$half.5) ## count for half net adjusted for radius
-date <- rep(as.factor(data_one_way$month), 2)
-station <- rep(data_one_way$Station, 2)
-net <- as.factor(c(rep("full", 40), rep("half", 40)))
-data <- data.frame(count=count, date=date, net=net)
-X <- model.matrix(~ net * date)
-ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ .)
-summary(nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X)))
-# summary(qpmod <- glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(2.6)))
-# summary(glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(0.1))) 
-# summary(glmmPQL(count ~ net * date, random=~1|station, family=neg.bin(10)))
+We begin the analysis by loading the data used for measuring total count.
+At each sampling event $i$ we recorded the number of mysis shrimp caught and important covariates including date, site, net size, etc.
+```{r totalCount, echo=TRUE, include=TRUE, message=FALSE, cache=TRUE}
+load("~/mysis/data/mysisCountData.RData")                                 ## load count data
+mns <- by(mysisCountData[, "count"], mysisCountData[, "net"], mean)       ## mean of counts by net size
+vars <- by(mysisCountData[, "count"], mysisCountData[, "net"], var)       ## variance of counts by net size
+```
+
+After loading the data, we can look at histograms of data to get a better understanding of what variables effect mysis shrimp count.
+First we examine the total counts broken down by net size and sampling month
+
+```{r plotCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center"}
+ggplot(data=mysisCountData, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Histogram of normalized counts by net size and date")
+```
+
+From these histograms, we see little differences between the two net sizes, except for perhaps a spike in catches of about 120 by the small net in August.
+We test this formally by fitting a model to compare if the count of shrimp caught by the large net is the same as the count of shrimp caught by the large net, after normalizing for the area of the two nets and controlling for covariates.
+Because counts cannot be less that zero and often are right skewed, the assumption of Normality in the data is questionable.
+Hence, we don't want to use Gaussian methods like linear regression and ANOVA.
+Luckily, the class of models known a generalized linear models (or glms) offers a solution.
+These models allow for regression models that control for the effects of covariates like linear regression but do not assume the normal distribution.
+For count data, there are two natural distributions to use: the Poisson distribution and the negative binomial distribution.
+The Poisson distribution has the strong assumption that the mean equals the variance, which is often not met in practical datasets.
+In our data, the count means are $(`r I(mns)`)$ and the count variances are $(`r I(vars)`)$, for the small and large nets, respectively.
+Therefore we use a negative binomial model
+$$
+\label{neg:bin}
+y_i \sim \mbox{NegBin}(\mu_i, \phi),
+$$
+where $\mu$ is the mean of the negative binomial distribution and $\phi$ is an overdispersion parameter that allows for the mean and variance to be different.
+We model $\mu$ with covariates using a log link function
+$$ 
+\mu_i = log(\mathbf{X}_i \boldsymbol{\beta}),
+$$
+where $\mathbf{X}_i$ is the set of covariates for observation $i$.
+Then we perform inference on our coefficients $\boldsymbol{\beta}$, where the interpretation of $\beta_j$ is a percent change in count per unit change in $X_{ij}$.
+```{r totalCountModel, echo=TRUE, eval=TRUE, results="hide"}
+## construct covariates of net size, date and their interaction
+X <- model.matrix(~ net * date - 1, data=mysisCountData)  
+## fit negative binomial model
+summary(nbmod <- glm.nb(y ~ ., data = data.frame(y=mysisCountData$count, X)))
+## get coefficient estimates
 coefs <- c(nbmod$coefficients, nbmod$theta)           
@@
 
-```{r countModelResults, echo=FALSE, cache=TRUE}
+```{r countModelResults, echo=TRUE}
 tableData <- cbind(summary(nbmod)$coeff[1:6, ])
@@
                          "Net Size * August", "Net Size * September")
-# htmlTable(x=round(tableData, digits=2),
-#           caption=paste("Negative Binomial Regression parameter estimates"),
-#           header = c("  Estimate  ", "  Std. Error  ", "  z value  ", "  p-value  "),
-#           rowlabel = "Variable",
-#           rnames = c("Intercept",
-#                     "Net Size",
-#                     "August",
-#                     "September",
-#                     "Net Size * August",
-#                     "Net Size * September"))
 kable(tableData, digits = 4, format="pandoc")
@@
 
-## Model for counts by gender
-```{r, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE}
-set.seed(123)
-fullData <- read.csv("~/mysis/data/lengthData.csv", skip=4)
+## Model for counts by sex
 
-males <- rep(0, 80)
-females <- rep(0, 80)
-juveniles <- rep(0, 80)
-unknowns <- rep(0, 80)
-size <- rep(0, 80)
-date_tmp <- rep(0, 80)
+The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.
+To do this we use (\ref{eq:negBin}) above.
+We begin by loading the data where counts are divided into sex classes.
 
-idx <- 1
-for(i in unique(fullData$Sample.Label)){
-  tmp_data = subset(fullData, fullData$Sample.Label == i)
-  tmp = substr(as.character(tmp_data$Gender), 0, 1)
-  males[idx] <- sum(tmp == "M")
-  juveniles[idx] <- sum(tmp == "J")
-  females[idx] <- sum(tmp == "F")
-  unknowns[idx] <- sum(tmp == "U")
-  size[idx] <- tmp_data$Size..m.[1]
-  date_tmp[idx] <- tmp_data$Date[1]
-  idx <- idx + 1
-}
-
-## construct count vector
-count <- c(males, females, juveniles, unknowns)
-gender <- factor(rep(1:4, each=80))
-## Correct for mislabeling
-date_tmp[70] <- 4
-date_tmp[72] <- 4
-date <- rep(0, 80)
-date[date_tmp == 1] <- 1
-date[date_tmp == 2] <- 1
-date[date_tmp == 3] <- 2
-date[date_tmp == 4] <- 3
-
-date <- as.factor(date)
-net <- as.factor(size)
-count[rep(net, times=4) == 0.5] <- count[rep(net, times=4) == 0.5] * 4
-data <- data.frame(count=count, date=rep(date, times=4), net=rep(net, times=4), gender=gender)
-# X <- model.matrix(~ net * date + net * gender + gender * date, data=data)
-X <- model.matrix(~ net * gender + date, data=data)
-ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ gender)
-ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(net ~ date)
-ggplot(data=data, aes(count)) + geom_histogram() + facet_grid(gender ~ date)
-nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X))
+```{r, echo=TRUE, message=FALSE, results="hide"}
+## load data
+load("~/mysis/data/mysisSexCountData.RData")   
+## construct model matrix for covariates
+X <- model.matrix(~ net * gender + net * date + gender * date-1, 
+                  data=mysisSexCountData)
+## fit model
+nbmod <- glm.nb(y ~ ., data = data.frame(y=mysisSexCountData$count, X))
 ```
 
-To test for a change in counts broken down by gender category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
-The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
+To explore this data in more detail we examine two sets of histograms, the counts of each sex class broken down by date in figure1 and the counts of each sex class broken down by net size in figure2.
+From these we see...
 
-```{r sexResults, echo=FALSE, cache=TRUE}
+```{r plot SexCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold"}
+ggplot(data=mysisSexCountData, aes(count)) +         
+  geom_histogram() + facet_grid(net ~ gender) + 
+  ggtitle("Histogram of normalized counts by sex and net size")
+ggplot(data=mysisSexCountData, aes(count)) +
+  geom_histogram() + facet_grid(gender ~ date) + 
+  ggtitle("Histogram of normalized counts by sex and date")
+```
 
+To test for a change in counts broken down by gender category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling, the interaction between net size and sampling date, and the interaction between sampling date and gender.
+The results shown below show that there is no evidence of an interaction between net size and sampling date and no evidence of an interaction between net size and gender.
+
+```{r sexResults, echo=TRUE, cache=TRUE}
 tableData <- cbind(summary(nbmod)$coeff)
@@
                          "Net Size * Females", "Net Size * Juveniles", 
-                         "Net Size * Unknowns")
+                         "Net Size * Unknowns", "Net Size * August",
+                         "Net Size * September",
+                         "Females * August", "Juveniles * August", 
+                         "Unknown Gender * August",
+                         "Females * September", "Juveniles * September", 
+                         "Unknown Gender * September")
+kable(tableData, digits = 4, format="pandoc")
+```
 
+Thus, we refit the model without the interaction between net size and sampling date. To do this we start by loading the new covariate matrix $\mathbf{X}$.
+
+```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE}
+X <- model.matrix(~ net * gender + gender * date, data=mysisSexCountData)
+nbmod <- glm.nb(y ~ . -1, data = data.frame(y=mysisSexCountData$count, X))
+```
+
+On this reduced model, the results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
+```{r sexResultsReduced, echo=TRUE, cache=TRUE}
+tableData <- cbind(summary(nbmod)$coeff)
+colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
+rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
+                         "Unknown Gender", "August", "September", 
+                         "Net Size * Females", "Net Size * Juveniles", 
+                         "Net Size * Unknowns", 
+                         "Females * August", "Juveniles * August", 
+                         "Unknown Gender * August",
+                         "Females * September", "Juveniles * September", 
+                         "Unknown Gender * September")
 kable(tableData, digits = 4, format="pandoc")
@@
 ## Model for proportion of juveniles
-```{r, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE}
+```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE}
 set.seed(123)
@@
 
+```{r plotJuviCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center"}
+ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .) + 
+  ggtitle("Histogram of proportion of juveniles by net size")
+```
+
 To test for a change in the proportion of juveniles caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
@@
 
-```{r juvenileResults, echo=FALSE, cache=TRUE}
+```{r juvenileResults, echo=TRUE, cache=TRUE}
 tableData <- cbind(summary(brmod)$coeff$mean)
@@
 
-```{r, echo=FALSE, include=FALSE, message=FALSE, cache=TRUE}
-set.seed(123)
-data_length <- read.csv("~/mysis/data/lengthData.csv", skip = 4, header = TRUE)
-net <- data_length$Size..m.
-y <- data_length$Length..mm.
-gender <- as.factor(substr(str_trim(as.character(data_length$Gender)), 0, 1))
-                       ## remove whitespace from string, then truncate into 
-                       ## broad gender classes, not subclasses
+```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE}
+# data_length <- read.csv("~/mysis/data/lengthData.csv", skip = 4, header = TRUE)
+# net <- data_length$Size..m.
+# y <- data_length$Length..mm.
+# gender <- as.factor(substr(str_trim(as.character(data_length$Gender)), 0, 1))
+# station <- data_length$Station
+# date <- as.numeric(data_length$Date)
+# date[7330:7519] <- 4
+# date[7559:7773] <- 4
+# date[date == 2] <- 1      ## collapse into monthly values
+# date[date == 3] <- 2
+# date[date == 4] <- 3
+# date <- as.factor(date)
+# # plot(as.numeric(date), type='l', main = "Data looks better")
+# data <- data.frame(y=y, date=date, net=as.factor(net), gender=gender)
+# levels(data$date) <- c("July", "August", "Septempber")
+# levels(data$net) <- c("Small Net", "Large Net")
+# levels(data$gender) <- c("Female", "Juvenile", "Male", "Unknown")
 
-date <- as.numeric(data_length$Date)
-date[date == 2] <- 1      ## collapse into monthly values
-date[date == 3] <- 2
-date[date == 4] <- 3
-date <- as.factor(date)
-station <- data_length$Station
-# plot(as.numeric(date), type='l', main = "Is this a data error? - yes?")
-## correct date mislabeling
-date <- as.numeric(data_length$Date)
-date[7330:7519] <- 4
-date[7559:7773] <- 4
-date[date == 2] <- 1      ## collapse into monthly values
-date[date == 3] <- 2
-date[date == 4] <- 3
-date <- as.factor(date)
-# plot(as.numeric(date), type='l', main = "Data looks better")
-data <- data.frame(y=y, date=date, net=net, gender=gender)
-X <- model.matrix(~ net + date + gender, data=data)
-ggplot(data=data, aes(y)) + geom_histogram() + facet_grid(net ~ date)
+load("~/mysis/data/mysisLengthData.RData")
+X <- model.matrix(~ net + date + gender, data=mysisLengthData)
 idx <- attr(uniquecombs(X), "index")
-summary(lmmod <- lm(y ~ net + date + gender))
+summary(lmmod <- lm(y ~ net + date * gender, data=mysisLengthData))
 library(lme4)
-summary(lmemod <- lmer(y ~ net + date + gender + (1|station)))
-rlmmod <- rlm(y ~ net + date + gender)
+# summary(lmemod <- lmer(y ~ net * date + net * gender +
+#                          date * gender + (1|station), data=mysisLengthData))
+# summary(lmemod <- lmer(y ~ net + date * gender + (1|station),
+#                        data=mysisLengthData))
+# summary(rlmmod <- rlm(y ~ net * date + net * gender + 
+#                         date * gender, data=mysisLengthData))))
+summary(rlmmod <- rlm(y ~ net + date * gender, data=mysisLengthData))
+tvalues <- summary(rlmmod)$coefficients[, 3]
+df <- summary(rlmmod)$df[2]
+pvalues <- pt(abs(tvalues), df, lower.tail=FALSE) * 2
 ```
 
-```{r lengthResults, echo=FALSE, cache=TRUE}
-tableData <- cbind(summary(lmmod)$coeff)
+```{r plotLength, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold"}
+ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ date) +
+  ggtitle("Lengths by net size")
+ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ gender) + 
+  ggtitle("Lengths by sex and net size")
+ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(date ~ gender) + 
+  ggtitle("Lengths by sex and date")
+```
+
+```{r lengthResults, echo=TRUE, cache=TRUE}
+tableData <- cbind(summary(rlmmod)$coeff, pvalues)
 colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
-                         "Juvenile", "Male", "Unknown Gender")
+rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
+                         "Unknown Gender", "August", "September", 
+                         "Females * August", "Juveniles * August", 
+                         "Unknown Gender * August",
+                         "Females * September", "Juveniles * September", 
+                         "Unknown Gender * September")
+
 kable(tableData, digits = 4, format="pandoc")
+```
 
-# plot(lmmod)
-# summary(rlmmod)
-# plot(rlmmod)
-# kruskal.test(y ~ net)
+```{r effect size, echo=TRUE, eval=FALSE}
+SD <- sqrt(((sum(mysisLengthData$net == "Large Net") - 1) *
+              sd(mysisLengthData$y[mysisLengthData$net == "Large Net"])^2 + 
+              (sum(mysisLengthData$net == "Small Net") - 1) *
+              sd(mysisLengthData$y[mysisLengthData$net == "Small Net"])^2) / 
+             (sum(mysisLengthData$net == "Large Net") + 
+                sum(mysisLengthData$net == "Small Net")))
+
+d <- (mean(mysisLengthData$y[mysisLengthData$net == "Large Net"]) - 
+        mean(mysisLengthData$y[mysisLengthData$net == "Small Net"])) / SD
 ```
+We see that there is a significant effect of net size on mean length caught (p = `r I(pvalues[2])`) after controlling for date, gender class, and an interaction between date and gender class.
+Although this is statisticallly significant, the effect size is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`).
+given a large sample size, a hypothesis test will show staistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05).
+Therefore, the practical effect of a difference in mean length of `r I(summary(rlmmod)$coefficients[2, 1])`mm on a species with a mean length of `r I(mean(mysisLengthData$y))`mm is small (this is the smallest effect of all the effects estimated by almost a factor of two) and a difference in means of this size is not of practical interest.

### 2015-10-12 _drafts/mysisAnalysis.Rmd
updated gh pages

--- 2015-10-12 results/mysisResults.Rmd
@@
 require(knitr)
+knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
 require(ggplot2)
@@
 ```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE}
-set.seed(123)
-data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40,
-                         header = TRUE, stringsAsFactors=FALSE)
-
-prop <- c(as.numeric(data_one_way$full.2),    ## proportion of females full net
-             as.numeric(data_one_way$half.2)) ## proportion of females half net
-N <- length(prop)
-## Set improper values to NA
-
-date <- rep(as.factor(data_one_way$month), 2)
-net <- as.factor(c(rep("full", 40), rep("half", 40)))
-data <- data.frame(prop=prop, date=date, net=net)
-levels(data$date) <- c("July", "August", "Septempber")
-levels(data$net) <- c("Small Net", "Large Net")
+load("~/mysis/data/mysisSexCountData.RData")
 X <- model.matrix(~ net * date)

### 2015-10-12 betaRegression/betaRegressionExample.Rmd

@@
   vector[p] beta;
+  vector[N] A;             // parameter for beta distn
+  vector[N] B;             // parameter for beta distn
+  vector[N] mu;    // transformed linear predictor
+  vector[N] Xbeta;                  // linear predictor
   beta <- theta * 10;               // matt trick if desired
-}
-model {
-  // model calculations
-  vector[N] Xbeta;                  // linear predictor
-  vector<lower=0,upper=1>[N] mu;    // transformed linear predictor
-  vector<lower=0>[N] A;             // parameter for beta distn
-  vector<lower=0>[N] B;             // parameter for beta distn
+//  vector<lower=0,upper=1>[N] mu;    // transformed linear predictor
   Xbeta <- X * beta;
@@
   B <- (1.0 - mu) * phi;
+}
+model {
+  // model calculations
   // priors
@@
   y ~ beta(A, B);
+} generated quantities {
+vector[N] log_lik;
+  for (n in 1:N){
+    log_lik[n] <- beta_log(y[n], A[n], B[n]);
+  }
 }
 "
-write(model, file = "../stanModels/betaRegression-glm.stan")
-sm <- stan_model("../stanModels/betaRegression-glm.stan")
+write(model, file = "~/mysis/stanModels/betaRegression-glm.stan")
+sm <- stan_model("~/mysis/stanModels/betaRegression-glm.stan")
 ```

### 2015-10-12 countModel/countData.Rmd

@@
 e <- rstan::extract(m, pars = c("beta", "phi"))
+library(MASS)
+nbmod <- glm.nb(y ~ . -1, data = data.frame(y=count, X))
+summary(nbmod)
+coefs <- c(nbmod$coefficients, nbmod$theta)
 par(mfrow = c(2, 3))
@@
       plot(density(e[[i]][,j]), main = paste("beta", j, sep=""))
+      abline(v = coefs[j], lwd = 3, col = "red")
     }
@@
     plot(density(e[[i]]), main = "phi")
+    abline(v = coefs[dim(e[[i]])[2] + 1], lwd = 3, col = "red")
   }

### 2015-10-12 juvenileRatioModel/juvenileRatioModel.Rmd

--- 2015-08-17 betaRegression/betaRegressionExample.Rmd
@@
 ---
-title: "negativeBinomial"
+title: "Beta Regression"
 author: "John Tipton"
@@
 ---
-
 # Write a beta GLM in STAN
@@
 library(rstan)
-model <- "
-// beta Regression
-data {
-  int<lower=1> N;                   // sample size
-  int<lower=1> p;                   // p predictors
-  vector<lower=0,upper=1>[N] y;     // response 
-  matrix[N, p] X;                   // predictor matrix
-}
-parameters {
-  vector[p] theta;                  // reg coefficients
-  real<lower=0> phi;                // dispersion parameter
-}
-transformed parameters{
-  vector[p] beta;
-  beta <- theta * 10;               // matt trick if desired
-}
-model {
-  // model calculations
-  vector[N] Xbeta;                  // linear predictor
-  vector<lower=0,upper=1>[N] mu;    // transformed linear predictor
-  vector<lower=0>[N] A;             // parameter for beta distn
-  vector<lower=0>[N] B;             // parameter for beta distn
-  Xbeta <- X * beta;
-  for (i in 1:N) { 
-    mu[i] <- inv_logit(Xbeta[i]);   
-  }
-  A <- mu * phi;
-  B <- (1.0 - mu) * phi;
-  // priors
-  theta ~ normal(0, 1);   
-  phi ~ cauchy(0, 5);               // different options for phi  
-  //phi ~ inv_gamma(.001, .001);
-  //phi ~ uniform(0, 500);          // put upper on phi if using this
-  // likelihood
-  y ~ beta(A, B);
-}
-"
-write(model, file = "../stanModels/betaRegression-glm.stan")
-sm <- stan_model("../stanModels/betaRegression-glm.stan")
+sm <- stan_model("~/mysis/stanModels/betaRegression-glm.stan")
 ```
 
-## Simulate some Data
+## Load the Data
 
 ```{r}
-library(betareg)
+library(ggplot2)
 set.seed(123)
-N = 80
-beta <- c(2, -1, 1, -2, -1)
-x0 <- rep(1, N)
-x1 <- c(rep(1, N/2), rep(0, N/2)) ## Indicator for net size
-x2 <- rep(c(rep(1, 10), rep(0, 30)), 2) ## Indicator for July
-x3 <- rep(c(rep(0, 10), rep(1, 20), rep(0, 10)), 2) ## Indicator for August
-# x4 <- rep(c(rep(0, 30), rep(1, 10)), 2) ## Indicator for September
-x12 <- x1 * x2 ## Interaction between net size and July
-x13 <- x1 * x3 ## Interaction between net size and August
-# X <- cbind(x0, x1, x2, x3, x12, x13)
-X <- cbind(x1, x2, x3, x12, x13)
+data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40, header = TRUE, stringsAsFactors=FALSE)
 
-mu = plogis(X%*%beta)  # add noise if desired + rnorm(N, sd=.01)
-phi = 10
-A = mu*phi
-B = (1-mu)*phi
-y = rbeta(N, A, B)
-layout(matrix(c(1, 1, 2, 3), 2, 2, byrow=TRUE))
-hist(y, "FD")
-hist(y[x1 == 0], "FD")
-hist(y[x1 == 1], "FD")
+prop <- c(as.numeric(data_one_way$full.2),    ## proportion of females full net
+             as.numeric(data_one_way$half.2)) ## proportion of females half net
+N <- length(prop)
+## Set improper values to NA
+
+date <- rep(as.factor(data_one_way$month), 2)
+net <- as.factor(c(rep("full", 40), rep("half", 40)))
+data <- data.frame(prop=prop, date=date, net=net)
+X <- model.matrix(~ net * date)
+ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .)
 ```
@@
 ```{r}
-m <- sampling(sm, data = list(N = N, y = y, X = X, p=dim(X)[2]),
+m <- sampling(sm, data = list(N=N, y = prop, X = X, p=dim(X)[2]),
   pars = c("beta", "phi"),
@@
 e <- rstan::extract(m, pars = c("beta", "phi"))
-true_pars <- c(beta = beta, phi = phi)
 library(betareg)
-brmod <- betareg(y ~ . -1, data = data.frame(y, X))
+brmod <- betareg(prop ~ . -1, data = data.frame(prop, X))
 summary(brmod)
 coefs <- c(brmod$coefficients$mean, brmod$coefficients$precision)
-par(mfrow = c(2, 3))
+par(mfrow = c(3, 3))
 for(i in 1:2) {
   if(i == 1) {
-#     for(j in 1:6) {
-    for(j in 1:5) {
-      plot(density(e[[i]][,j]), main = names(true_pars)[i + j - 1])
-      abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
+    for(j in 1:6) {
+    # for(j in 1:5) {
+      plot(density(e[[i]][,j]), main = paste("beta", j, sep=""))
+      # abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
       abline(v = coefs[i + j - 1], lwd = 3, col = "red")
@@
   } else if(i == 2) {
-    plot(density(e[[i]]), main = names(true_pars)[i + j - 1])
-    abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
+    plot(density(e[[i]]), main = "phi")
+    # abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
     abline(v = coefs[i + j - 1], lwd = 3, col = "red")
@@
 
-idx <- order(colMeans(e$beta))
+ggs_caterpillar(S, family = "beta")
 
-ggs_caterpillar(S, family = "beta") + 
-  geom_point(aes(y=1:5, x=beta[idx]), col = "red")
-```
-
-```{r}
-hist(y[X[, 1] == 0])
-hist(y[X[, 1] == 1])
 ```

### 2015-10-12 lengthModel/lengthModel.Rmd

@@
   // data model:
-  y ~ normal(X * beta, s);
+  for(i in 1:N){
+    y[i] ~ normal(X[i] * beta, s[idx[i]]);
+  }
 }

### 2015-10-12 overdispersedPoisson/overdispersedPoissonExample.Rmd

---
title: "overdispersedPoisson"
author: "John Tipton"
date: "August 14, 2015"
output: html_document
---

# Write an overdispersed Poisson GLM in STAN

```{r}
library(rstan)
model <- "
// overdispersed Poisson parameterized 
// see Gelman and Hill (2007)
data {
  int<lower=1> N;       // number of observations
  int<lower=0> p;
  matrix[N, p] X;       // predictor
  int<lower=0> y[N];    // response
} transformed data {
  vector[p] J;
  matrix[p, p] I_p;
  // calculate placeholder values
  J <- rep_vector(0, p);
  I_p <- diag_matrix(rep_vector(1, p));
}
parameters {
  real<lower=0> tau;   // Poisson dispersion parameter
  vector[p] beta;      // Poisson mean parameters
  vector[N] epsilon;   // Poisson deviances
} transformed parameters {
  vector[N] means;     // Poisson means
  means <- exp(X * beta + epsilon);
} model {
  // priors:
  tau ~ cauchy(0, 20);
  beta ~ multi_normal(J, 100 * I_p);
  // data model:
  for(i in 1:N){
    epsilon[i] ~ normal(0, tau);
  }
  y ~ poisson_log(means);
}
"
write(model, file = "../stanModels/overdispersedPoisson-glm.stan")
sm <- stan_model("../stanModels/overdispersedPoisson-glm.stan")
```

## Simulate some Data

```{r}
set.seed(123)
N <- 80
tau <- 1
# beta <- c(1, 2, -1, 1, -2, -1)
beta <- c(2, -1, 1, -2, -1)
x0 <- rep(1, N)
x1 <- c(rep(1, N/2), rep(0, N/2)) ## Indicator for net size
x2 <- rep(c(rep(1, 10), rep(0, 30)), 2) ## Indicator for July
x3 <- rep(c(rep(0, 10), rep(1, 20), rep(0, 10)), 2) ## Indicator for August
# x4 <- rep(c(rep(0, 30), rep(1, 10)), 2) ## Indicator for September
x12 <- x1 * x2 ## Interaction between net size and July
x13 <- x1 * x3 ## Interaction between net size and August
# X <- cbind(x0, x1, x2, x3, x12, x13)
X <- cbind(x1, x2, x3, x12, x13)
y <- rpois(N, exp(X %*% beta + rnorm(N, 0, tau)))
layout(matrix(c(1, 1, 2, 3), 2, 2, byrow=TRUE))
hist(y)
hist(y[x1 == 0])
hist(y[x1 == 1])
```

## Fit the Model

```{r}
m <- sampling(sm, data = list(N = N, y = y, X = X, p=dim(X)[2]),
  pars = c("beta", "tau"),
  iter = 5000, chains = 4)
```

## Model output
```{r}
m
```

## Plot traceplots
```{r}
rstan::traceplot(m, inc_warmup = FALSE)
```

## Plot fitted vs. simulated values

```{r}
e <- rstan::extract(m, pars = c("beta", "tau"))
true_pars <- c(beta = beta, tau = tau)
library(lme4)
obs_effect<-1:N
overdisp.fit <- glmer(y ~ X - 1 + (1|obs_effect), family=poisson)
coefs <- c(overdisp.fit@beta, overdisp.fit@theta)
par(mfrow = c(3, 3))
for(i in 1:2) {
  if(i == 1) {
#     for(j in 1:6) {
    for(j in 1:5) {
      plot(density(e[[i]][,j]), main = names(true_pars)[i + j - 1])
      abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
      abline(v = coefs[i + j - 1], lwd = 3, col = "red")
    }
  } else if(i == 2) {
    plot(density(e[[i]]), main = names(true_pars)[i + j - 1])
    abline(v = true_pars[i + j - 1], lwd = 2, col = "grey", lty = 2)
    abline(v = coefs[i + j - 1], lwd = 3, col = "red")
  }
}
# legend("topright", legend = c("posterior", "true", "MASS::glm.nb"),
#   col = c("black", "grey", "red"), lty = c(1, 2, 1), lwd = c(1, 1.3, 3),
#   bty = "n")
```

```{r}
library(ggmcmc)
library(coda)
s <-  mcmc.list(lapply(1:ncol(m), function(x) mcmc(as.array(m)[,x,])))
S <- ggs(s)

idx <- order(colMeans(e$beta))

ggs_caterpillar(S, family = "beta") + 
  geom_point(aes(y=idx, x=beta), col = "red")
```

```{r}
hist(y[X[, 1] == 0])
hist(y[X[, 1] == 1])
```

### 2015-10-12 results/mysisResults.Rmd

@@
 
-prop <- c(as.numeric(data_one_way$full.2),    ## proportion of females full net
-             as.numeric(data_one_way$half.2)) ## proportion of females half net
+prop <- c(as.numeric(data_one_way$full.2),    ## proportion of juveniles full net
+             as.numeric(data_one_way$half.2)) ## proportion of juveniles half net
 N <- length(prop)

### 2015-10-12 _drafts/mysisAnalysis.Rmd

--- 2015-10-12 results/mysisResults.Rmd
@@
----
 title: "Mysis analysis"
 author: "John Tipton"
-date: "September 9, 2015"
-output: html_document
+date: "October 12, 2015"
 ---
@@
                          "Net Size * August", "Net Size * September")
-kable(tableData, digits = 4, format="pandoc")
+kable(tableData, digits = 4, format="html")
 ```
@@
                          "Unknown Gender * September")
-kable(tableData, digits = 4, format="pandoc")
+kable(tableData, digits = 4, format="html")
 ```
@@
 
-kable(tableData, digits = 4, format="pandoc")
+kable(tableData, digits = 4, format="html")
 ```

### 2015-12-16 responseToReviewers/mysisReviewerResponse.Rmd
updated writeup, started response to reviewers

---
output: word_document
---

title: "Response to reviewer's statistical comments on the paper: GLR-D-15-00174 - Effect of net size on estimates of abundance, size, age and sex ratio of Mysis diluviana Journal of Great Lakes Research"
author: "John Tipton"
date: "October 12, 2015"
---

```{r loadLibraries, echo=FALSE, include=FALSE, message=FALSE}
## We start by loading packages and define helper functions
options(digits=3)
set.seed(123)
library(knitr)
knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
library(ggplot2)
library(Gmisc)
library(MASS)
library(stringr)
library(mgcv)
library(lme4)
```

After loading the data, we can look at histograms of data to get a better understanding of what variables effect mysis shrimp count. First we examine the total counts broken down by net size and sampling month

```{r plotCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", eval=FALSE}
ggplot(data=mysisCountData, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
  ggtitle("Histogram of normalized counts by net size and date")
```

From these histograms, we see little differences between the two net sizes, except for perhaps a spike in catches of about 120 by the small net in August. We test this formally by fitting a model to compare if the count of shrimp caught by the large net is the same as the count of shrimp caught by the large net, after normalizing for the area of the two nets and controlling for covariates. Because counts cannot be less that zero and often are right skewed, the assumption of Normality in the data is questionable. Hence, we don't want to use Gaussian methods like linear regression and ANOVA. But, for completeness, let's look a a paired t-test and see what the results look like with the knowledge that this analysis is less than satisfactory.

```{r pariedTTest}
y1 <- mysisCountData$count[mysisCountData$net == "Small Net"]
y2 <- mysisCountData$count[mysisCountData$net == "Large Net"]
pairttest <- t.test(y1, y2, paired=TRUE)
```

From the paired t-test, we find that there is no significant difference ($t$ = `r pairttest$statistic`, $df$ = `r pairttest$parameter`, $p$-value = `r pairttest$p.value` ) in expected counts (after accounting for the different surface areas between the nets)

Luckily, the class of models known a generalized linear models (or glms) offers a solution. These models allow for regression models that control for the effects of covariates like linear regression but do not assume the normal distribution. For count data, there are two natural distributions to use: the Poisson distribution and the negative binomial distribution. The Poisson distribution has the strong assumption that the mean equals the variance, which is often not met in practical datasets. In our data, the count means are $(`r I(mns)`)$ and the count variances are $(`r I(vars)`)$, for the small and large nets, respectively. Therefore we use a negative binomial model
$$
\label{neg:bin}
y_i \sim \mbox{NegBin}(\mu_i, \phi),
$$
where $\mu$ is the mean of the negative binomial distribution and $\phi$ is an overdispersion parameter that allows for the mean and variance to be different. We model $\mu$ with covariates using a log link function
$$ 
\mu_i = log(\mathbf{X}_i \boldsymbol{\beta}),
$$
where $\mathbf{X}_i$ is the set of covariates for observation $i$. Then we perform inference on our coefficients $\boldsymbol{\beta}$, where the interpretation of $\beta_j$ is a percent change in count per unit change in $X_{ij}$.

```{r totalCountModel, echo=TRUE, eval=TRUE, results="hide"}
## construct covariates of net size, date and their interaction
X <- model.matrix(~ net * date - 1, data=mysisCountData)  
## fit negative binomial model
summary(nbmod <- glm.nb(y ~ ., data = data.frame(y=mysisCountData$count, X)))
## get coefficient estimates
coefs <- c(nbmod$coefficients, nbmod$theta)           
```

To test for a change in total counts between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date. The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).

```{r countModelResults, echo=TRUE, eval=FALSE}
tableData <- cbind(summary(nbmod)$coeff[1:6, ])
colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
                         "Net Size * August", "Net Size * September")
kable(tableData, digits = 4, format="html")
```

Associate Editor:
This is a useful analysis of mysid sampling methods that has implications to Great Lakes and beyond. The paper has received two careful reviews. Both reviewers were generally positive, but have some suggestions for improvements.

### Both reviewers wanted to know why you did not use a paired t test when comparing the two net types, which given the paired approach to sampling would appear to be the most appropriate method to analyze the data. If you have reasons not to use that approach, please elaborate in your response letter and add a sentence or two to the methods.

We begin the analysis by loading the data used for measuring total count. At each sampling event $i$ we recorded the number of mysis shrimp caught and important covariates including date, site, net size, etc.

```{r totalCount, echo=TRUE, include=TRUE, message=FALSE, cache=TRUE, eval=TRUE}
load("~/mysis/data/mysisCountData.RData")                                 ## load count data
```

## Another issue to consider is the catch of a few large and small individuals in the larger nets.  You should be able to test if that is simply due to the larger sample size obtained in the larger net. You suggest that may be the case.

### Finally, the length analysis is using every measured mysids as an independent sampling unit.  I think it would be more appropriate to use a paired t test on the mean lengths for that comparison as well, using each net instead of each mysid as the sampling unit. There is always the possibility that different size groups occur in different areas of the reservoir, and therefore that each mysids measured are not independent from the other mysids measured in a given net haul. Please consider that in your response.

Lars

## Reviewers' comments:

## Reviewer #1: Review of GLR-D-15-00174

### 8)      Line 145, 150 I'm not sure you controlled for these factors, rather it seems like they were factors in the analysis?

### 9)      Line 149 I don't think the mean length is the correct variable of interest.  The size distributions could be very different and provide similar mean lengths.  Rather, I think an analysis of the size structure is appropriate, perhaps using some sort of contingency table to evaluate the homogeneity of proportions.  I also think the analysis of population structure (male-female, juvenile) and male:female ratio would be more appropriately done by testing for homegenity of proportions and then looking at standardized deviation to see where the nets differed (if at all).

### 10)     The analysis of abundance (and other factors) was done basically to see if this study was done with one net or the other, would the results over the reservoir differ.  This is an interesting question and an important one to address.  But the net tows were done in paired combinations, so there is more information here.  I would like to see something like a paired t-test (using each individual set of net tows as a sample), or a regression of values from one net plotted against the other from each sample to see if there was any bias based on the paired tows.

### 11)     Line 163 and 166  I think you mean density and not number?

### 12)     Please provide F values, df so the stats can be evaluated

### 13)     Line 177 also caught more of the smallest individuals

## Reviewer #2: This manuscript has one straightforward message that will be useful to researchers interested in the collection of mysids, in that it supports the use of smaller diameter nets, which will be convenient from small vessels often used in lakes these animals have spread into. Overall this manuscript is acceptable for publication with some revisions. The details of the statistical tests need to be expanded upon and some language that downplays some of the size-distribution differences with different net sizes should be discussed in greater detail.

Major points:

### 2) The manuscript is short so you have plenty of space to expand upon your statistical methodology which isn't made clear, and your presentation of only p values is insufficient to assess your analysis.

### 4) Downplaying the differences in size-distribution associated with the net sizes doesn't make the paper more persuasive. There are plenty of reasons where researchers may target the tiny fraction of smallest or largest individuals which are being missed by the smaller nets. Discuss this difference and end with the point you make that if possible, use the larger nets, but for most work, the smaller nets are an acceptable substitute.

Detailed comments:
L38 Only 3 dates were sampled.

### L40 Size structure was slightly different (more at tails of size-distribution for large nets). Without this fact the last sentence of the abstract may not be clear.

### L122 use ind or N rather than lowercase n

### L130 use s or seconds

### L141 Be consistent   N/m2, number/m2

### L142-151. You can be more concise here for some content and have to expand on other. You always have the same effects so you don't have to relist them. State the data isn't normal so a negative binomial transformation was used for regression on abundance, sex ratio…etc. See next note regarding statistical methods. You would be better off explaining the rationale for methods and the format in more detail rather than just the factors. For instance there is no detail given on the negative binomial model which can be constructed and assessed in a range of ways (likelihood format, AIC etc). Be descriptive here. If you used a specific stat software package, this would be the place to note it.

### L145-148 Why wouldn't you use a paired test? You collected each at the same time, so you could control much of the variability in catch (patchiness) by comparing them to each other (you comment L153-155 on this). Your size data shows multiple cohorts so it isn't surprising that you have non-normal data but the distributions are similar so should be easily discriminated using pairwise testing.

### L162-170. You give p values but no specifics as to method. The negative binomial model requires a range of output to assess the fit. The p values by themselves are not necessarily relevant so F (or z) values or r-mean squares should be included as well.

### L156. It is unusual for an author to cite unpublished data on their own paper. Ideally, you could easily add an additional graph that shows the series given the few number of tables and figures.
L161. Are you alluding to a drawback in your methods here given the lack of identification of sex increasing during the study?
### L167 Don't repeat yourself. "between net size and sampling date" redundant. Instead, "effect or interactions".
### L168 As above, you can include all of these as one statement: "density or juvenile proportion".
L173 "likely not of practical interest".
### L175. So did you check for homogeneity of variance? A larger range but similar mode suggests that the larger nets had increased variance. This doesn't necessarily affect your results but should be noted. This might be relevant for any attempt at bioenergetics modelling for instance. The quartile plots in Fig. 2 suggest that the smaller net mean lengths are skewed downward (lower quartile larger than upper) which is reversed in the larger nets (except in July, which is more likely driven by the bimodal distribution in this month and suggests (there is no testing?) that the distributions are not normal and means are unlikely to be relevant for this month). You might be better off testing the bimodal distribution as two distributions (mode to mode). This may be a result of the different efficiencies of the two net sizes (you make no note of flowmeter readings and in your next section you calculate efficiency using areal ratio). It also might be worth more text describing the different distributions of %juveniles since the august and September distributions are very different even though their modes are similar (the distribution tail).

### L183. "probably not biologically relevant". Since you do not give these distributions (beyond the graphs), this isn't necessarily your call since some researchers may be focused on the very tails of the distribution that you dismiss. Instead, note that they are very small differences in numbers and that researchers interested in these larger individuals may wish to utilize the larger nets. Instead, move the sentence that is "Thus, these two nets…" up to be the second sentence (you state that there is essentially no difference in their distribution), then follow up with the caveat with respect to the larger size distribution afterwards.

### L187. The slight bias where the smaller nets are missing the largest individuals may by more relevant in late October or November when many programs sample to get egg production rates. Larger individuals appear to be able to avoid the smaller nets (they are certainly more powerful swimmers) and may also detect a difference in the net-mouth "head-wave" if your efficiencies are different.

### L203. As noted previously, be careful with your wording. Your equipment is similar to certain studies, but it can just as easily be said that your mesh size (for instance) is smaller than the majority that specifically are used to target mysids. Your comment on the size preventing loss of small mysids may be spurious. Given your distributions (and mean size of ~10mm) did you even see any mysids less than 1 mm in size (or even up to 3 mm)? Fig. 3 suggests you didn't, so you shouldn't expect net extrusion at a 3:1 ratio. If the mesh size was used to be consistent with other monitoring methods, then just state that, which is a perfectly fine reasoning and supported by your results.
L214 "Our findings…"
### L217 "While the larger net captured more…" - you have waited until here to note that there are two possible reasons (increased sampling area or reduced net avoidance). This should be in your first sentence of your discussion since it is the one factor which is clearly different between the sampling methods. Your second point about the large net also catching larger numbers of smaller individuals only suggests a sampling area effect (wording should be "did not support"). Both factors can be happening with the different size animals - larger mature adults are significantly more powerful swimmers than the smaller juveniles. Your last point in this paragraph should be expanded upon to note the comments listed above (that some researchers may be targeting the largest individuals to estimate population growth rates).
L229. Remove "Conclusions" - it is the last paragraph anyway.
L230. The term "biologically relevant" is loaded and likely to create disagreement even if there isn't any in the findings. You don't know what is biologically relevant to mysids or to other trophic levels that are affected by them (e.g. large individuals are known to have a significant impact on trophic dynamics beyond just their biomass (a la Fry and Quinones 1994, Ellis et al. 2011) and the presence of mysids will shift size structure of the entire plankton when present. Size is important.
L236. This sentence essentially repeats the last in the previous paragraph. Combine them and expand on as noted in the L217 comment.

Table 1: You can save a huge amount of space in your table by just using "Conical" instead of "Conical plankton" and "Pyramid" instead of "Inverted pyramid" on two lines and note in the caption (you should also explain "framed net"). This table can easily be portrait with this change thus freeing up more space. This table is a major source of your extremely large reference section. This would be better suited for a supplementary appendix since it is not directly essential to the finding of this study.

Fig. 1. If you are going to note lat-long lines on your map, you should use a different line marking (e.g. dashed).
Fig. 2 and Fig. 3 captions don't match your graphs because the next two figures pages are reversed.
Fig. 2. The dots are means and the lines are modes? This isn't made clear. Axes titles may have to be larger.
Fig. 3 The 1.0 m diameter net bars should be white to be consistent with the other graph. Font may have to be larger for the axes and the inset pie-graphs.

This document shows the analyses performed in Johnson et. al.

## Model for counts by sex

The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates. To do this we use (\ref{eq:negBin}) above. We begin by loading the data where counts are divided into sex classes.

```{r, echo=TRUE, message=FALSE, results="hide", eval=FALSE}
## load data
load("~/mysis/data/mysisSexCountData.RData")   
## construct model matrix for covariates
X <- model.matrix(~ net * gender + net * date + gender * date-1, 
                  data=mysisSexCountData)
## fit model
nbmod <- glm.nb(y ~ ., data = data.frame(y=mysisSexCountData$count, X))
```

To explore this data in more detail we examine two sets of histograms, the counts of each sex class broken down by date in figure1 and the counts of each sex class broken down by net size in figure2. From these we see...

```{r plot SexCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold", eval=FALSE}
ggplot(data=mysisSexCountData, aes(count)) +         
  geom_histogram() + facet_grid(net ~ gender) + 
  ggtitle("Histogram of normalized counts by sex and net size")
ggplot(data=mysisSexCountData, aes(count)) +
  geom_histogram() + facet_grid(gender ~ date) + 
  ggtitle("Histogram of normalized counts by sex and date")
```

To test for a change in counts broken down by gender category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling, the interaction between net size and sampling date, and the interaction between sampling date and gender. The results shown below show that there is no evidence of an interaction between net size and sampling date and no evidence of an interaction between net size and gender.

```{r sexResults, echo=TRUE, cache=TRUE, eval=FALSE}
tableData <- cbind(summary(nbmod)$coeff)
colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
                         "Unknown Gender", "August", "September", 
                         "Net Size * Females", "Net Size * Juveniles", 
                         "Net Size * Unknowns", "Net Size * August",
                         "Net Size * September",
                         "Females * August", "Juveniles * August", 
                         "Unknown Gender * August",
                         "Females * September", "Juveniles * September", 
                         "Unknown Gender * September")
kable(tableData, digits = 4, format="html")
```

Thus, we refit the model without the interaction between net size and sampling date. To do this we start by loading the new covariate matrix $\mathbf{X}$.

```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE, eval=FALSE}
X <- model.matrix(~ net * gender + gender * date, data=mysisSexCountData)
nbmod <- glm.nb(y ~ . -1, data = data.frame(y=mysisSexCountData$count, X))
```

On this reduced model, the results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
```{r sexResultsReduced, echo=TRUE, cache=TRUE, eval=FALSE}
tableData <- cbind(summary(nbmod)$coeff)
colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
                         "Unknown Gender", "August", "September", 
                         "Net Size * Females", "Net Size * Juveniles", 
                         "Net Size * Unknowns", 
                         "Females * August", "Juveniles * August", 
                         "Unknown Gender * August",
                         "Females * September", "Juveniles * September", 
                         "Unknown Gender * September")
kable(tableData, digits = 4, format="html")
```

## Model for proportion of juveniles
```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
set.seed(123)
data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40,
                         header = TRUE, stringsAsFactors=FALSE)

prop <- c(as.numeric(data_one_way$full.2),    ## proportion of females full net
             as.numeric(data_one_way$half.2)) ## proportion of females half net
N <- length(prop)
## Set improper values to NA

date <- rep(as.factor(data_one_way$month), 2)
net <- as.factor(c(rep("full", 40), rep("half", 40)))
data <- data.frame(prop=prop, date=date, net=net)
levels(data$date) <- c("July", "August", "Septempber")
levels(data$net) <- c("Small Net", "Large Net")
X <- model.matrix(~ net * date)
library(betareg)
brmod <- betareg(prop ~ . -1, data = data.frame(prop, X))
summary(brmod)
```

```{r plotJuviCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", eval=FALSE}
ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .) + 
  ggtitle("Histogram of proportion of juveniles by net size")
```

To test for a change in the proportion of juveniles caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date. The results shown below show that there is no evidence of an effect on the proportion of juveniles caught between the two net sizes (p=`r I(round(summary(brmod)$coef$mean[2, 4], 4))`).

```{r juvenileResults, echo=TRUE, cache=TRUE, eval=FALSE}
tableData <- cbind(summary(brmod)$coeff$mean)
colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
                         "Net Size * August", "Net Size * September")
kable(tableData, digits = 4, format="html")
```

## Length model

```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
# data_length <- read.csv("~/mysis/data/lengthData.csv", skip = 4, header = TRUE)
# net <- data_length$Size..m.
# y <- data_length$Length..mm.
# gender <- as.factor(substr(str_trim(as.character(data_length$Gender)), 0, 1))
# station <- data_length$Station
# date <- as.numeric(data_length$Date)
# date[7330:7519] <- 4
# date[7559:7773] <- 4
# date[date == 2] <- 1      ## collapse into monthly values
# date[date == 3] <- 2
# date[date == 4] <- 3
# date <- as.factor(date)
# # plot(as.numeric(date), type='l', main = "Data looks better")
# data <- data.frame(y=y, date=date, net=as.factor(net), gender=gender)
# levels(data$date) <- c("July", "August", "Septempber")
# levels(data$net) <- c("Small Net", "Large Net")
# levels(data$gender) <- c("Female", "Juvenile", "Male", "Unknown")

load("~/mysis/data/mysisLengthData.RData")
X <- model.matrix(~ net + date + gender, data=mysisLengthData)
idx <- attr(uniquecombs(X), "index")
summary(lmmod <- lm(y ~ net + date * gender, data=mysisLengthData))
library(lme4)
# summary(lmemod <- lmer(y ~ net * date + net * gender +
#                          date * gender + (1|station), data=mysisLengthData))
# summary(lmemod <- lmer(y ~ net + date * gender + (1|station),
#                        data=mysisLengthData))
# summary(rlmmod <- rlm(y ~ net * date + net * gender + 
#                         date * gender, data=mysisLengthData))))
summary(rlmmod <- rlm(y ~ net + date * gender, data=mysisLengthData))
tvalues <- summary(rlmmod)$coefficients[, 3]
df <- summary(rlmmod)$df[2]
pvalues <- pt(abs(tvalues), df, lower.tail=FALSE) * 2
```

```{r plotLength, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold", eval=FALSE}
ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ date) +
  ggtitle("Lengths by net size")
ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ gender) + 
  ggtitle("Lengths by sex and net size")
ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(date ~ gender) + 
  ggtitle("Lengths by sex and date")
```

```{r lengthResults, echo=TRUE, cache=TRUE, eval=FALSE}
tableData <- cbind(summary(rlmmod)$coeff, pvalues)
colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
                         "Unknown Gender", "August", "September", 
                         "Females * August", "Juveniles * August", 
                         "Unknown Gender * August",
                         "Females * September", "Juveniles * September", 
                         "Unknown Gender * September")

kable(tableData, digits = 4, format="html")
```

```{r effect size, echo=TRUE, eval=FALSE}
SD <- sqrt(((sum(mysisLengthData$net == "Large Net") - 1) *
              sd(mysisLengthData$y[mysisLengthData$net == "Large Net"])^2 + 
              (sum(mysisLengthData$net == "Small Net") - 1) *
              sd(mysisLengthData$y[mysisLengthData$net == "Small Net"])^2) / 
             (sum(mysisLengthData$net == "Large Net") + 
                sum(mysisLengthData$net == "Small Net")))

d <- (mean(mysisLengthData$y[mysisLengthData$net == "Large Net"]) - 
        mean(mysisLengthData$y[mysisLengthData$net == "Small Net"])) / SD
```
We see that there is a significant effect of net size on mean length caught (p = `r I(pvalues[2])`) after controlling for date, gender class, and an interaction between date and gender class. Although this is statisticallly significant, the effect size is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`). given a large sample size, a hypothesis test will show staistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05). Therefore, the practical effect of a difference in mean length of `r I(summary(rlmmod)$coefficients[2, 1])`mm on a species with a mean length of `r I(mean(mysisLengthData$y))`mm is small (this is the smallest effect of all the effects estimated by almost a factor of two) and a difference in means of this size is not of practical interest.

### 2015-12-17 responseToReviewers/mysisReviewerResponse.Rmd
updated work

@@
 
-After loading the data, we can look at histograms of data to get a better understanding of what variables effect mysis shrimp count.
-First we examine the total counts broken down by net size and sampling month
+Associate Editor:
+This is a useful analysis of mysid sampling methods that has implications to Great Lakes and beyond.
+The paper has received two careful reviews.
+Both reviewers were generally positive, but have some suggestions for improvements.
+
+### Both reviewers wanted to know why you did not use a paired t test when comparing the two net types, which given the paired approach to sampling would appear to be the most appropriate method to analyze the data.
+If you have reasons not to use that approach, please elaborate in your response letter and add a sentence or two to the methods.
+
+We begin the analysis by loading the data used for measuring total count.
+At each sampling event we recorded the number of mysis shrimp caught and important covariates including date, site, net size, etc.
+
+```{r totalCount, echo=TRUE, include=TRUE, message=FALSE, cache=TRUE, eval=TRUE}
+load("~/mysis/data/mysisCountData.RData")                                 ## load count data
+```
+
+By looking at the distributions of counts across time, we see that they are similar between net sizes, but the normality of the counts is questionable.
 
@@
 ```
-
-From these histograms, we see little differences between the two net sizes, except for perhaps a spike in catches of about 120 by the small net in August.
-We test this formally by fitting a model to compare if the count of shrimp caught by the large net is the same as the count of shrimp caught by the large net, after normalizing for the area of the two nets and controlling for covariates.
 Because counts cannot be less that zero and often are right skewed, the assumption of Normality in the data is questionable.
@@
 
-From the paired t-test, we find that there is no significant difference ($t$ = `r pairttest$statistic`, $df$ = `r pairttest$parameter`, $p$-value = `r pairttest$p.value` ) in expected counts (after accounting for the different surface areas between the nets)
+From the paired t-test, we find that there is no significant difference ($t$ = `r pairttest$statistic`, $df$ = `r pairttest$parameter`, $p$-value = `r pairttest$p.value` ) in expected counts.
 
-Luckily, the class of models known a generalized linear models (or glms) offers a solution.
-These models allow for regression models that control for the effects of covariates like linear regression but do not assume the normal distribution.
-For count data, there are two natural distributions to use: the Poisson distribution and the negative binomial distribution.
-The Poisson distribution has the strong assumption that the mean equals the variance, which is often not met in practical datasets.
-In our data, the count means are $(`r I(mns)`)$ and the count variances are $(`r I(vars)`)$, for the small and large nets, respectively.
-Therefore we use a negative binomial model
-$$
-\label{neg:bin}
-y_i \sim \mbox{NegBin}(\mu_i, \phi),
-$$
-where $\mu$ is the mean of the negative binomial distribution and $\phi$ is an overdispersion parameter that allows for the mean and variance to be different.
-We model $\mu$ with covariates using a log link function
-$$ 
-\mu_i = log(\mathbf{X}_i \boldsymbol{\beta}),
-$$
-where $\mathbf{X}_i$ is the set of covariates for observation $i$.
-Then we perform inference on our coefficients $\boldsymbol{\beta}$, where the interpretation of $\beta_j$ is a percent change in count per unit change in $X_{ij}$.
+We can also look at a paired t-test for the mean length of the nets, although this loses alot of information because we are only testing the sample level means without accounting for the differences in uncertainties about the site level means due to different sample sizes between nets
 
-```{r totalCountModel, echo=TRUE, eval=TRUE, results="hide"}
-## construct covariates of net size, date and their interaction
-X <- model.matrix(~ net * date - 1, data=mysisCountData)  
-## fit negative binomial model
-summary(nbmod <- glm.nb(y ~ ., data = data.frame(y=mysisCountData$count, X)))
-## get coefficient estimates
-coefs <- c(nbmod$coefficients, nbmod$theta)           
+```{r lengthData}
+load("~/mysis/data/mysisLengthData.RData")
 ```
 
-To test for a change in total counts between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
-The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
+First lets look at some histograms and get an idea of what the data look like.
 
-```{r countModelResults, echo=TRUE, eval=FALSE}
-tableData <- cbind(summary(nbmod)$coeff[1:6, ])
+```{r, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold", eval=FALSE}
+ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ date) +
+  ggtitle("Lengths by net size")
+```
+
+When looking at the distribution of lengths between the two net sizes, we see that the histograms look quite similar across the different dates.
+What stands out is that the small net catches about a quarter of the number of shrimp, as expected by the difference in net sizes, whereas the general shapes of the histograms appear quite similar across net sizes, suggesting there is not a difference in shrimp length distribution between nets across time.
+
+```{r, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold", eval=FALSE}
+ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ gender) + 
+  ggtitle("Lengths by sex and net size")
+```
+
+When we plot the distribution of lengths between the net sizes with respect to sex class, the distributions look quite similar as well.
+This suggests that there is not a lot of difference in length distribution of shrimp between the two net sizes when broken down by sex class.
+
+```{r pairedTTestLength}
+y1 <- by(mysisLengthData$y[mysisLengthData$net == "Small Net"], 
+         mysisLengthData$label[mysisLengthData$net == "Small Net"], mean)
+y1 <- y1[!is.na(y1)]
+y2 <- by(mysisLengthData$y[mysisLengthData$net == "Large Net"], 
+         mysisLengthData$label[mysisLengthData$net == "Large Net"], mean)
+y2 <- y2[!is.na(y2)]
+pairttestlength <- t.test(y1, y2, paired=TRUE)
+layout(matrix(1:2,2,1))
+hist(y1)
+hist(y2)
+```
+
+From the paired t-test, we find that there is a significant difference ($t$ = `r pairttest$statistic`, $df$ = `r pairttest$parameter`, $p$-value = `r pairttest$p.value` ) in expected counts, although the $p$-value is close to the 0.05 level.
+Despite the statistically significant difference, the practical difference is quite small as judged by the histograms and small difference in shrimp size.
+
+To better account for the different sample sizes and other ....
+
+```{r}
+
+X <- model.matrix(~ net + date + gender, data=mysisLengthData)
+idx <- attr(uniquecombs(X), "index")
+summary(lmmod <- lm(y ~ net + date + gender + station, data=mysisLengthData))
+library(lme4)
+# summary(lmemod <- lmer(y ~ net * date + net * gender +
+#                          date * gender + (1|station), data=mysisLengthData))
+# summary(lmemod <- lmer(y ~ net + date * gender + (1|station),
+#                        data=mysisLengthData))
+# summary(rlmmod <- rlm(y ~ net * date + net * gender + 
+#                         date * gender, data=mysisLengthData))))
+summary(rlmmod <- rlm(y ~ net + date + gender + station, data=mysisLengthData))
+tvalues <- summary(rlmmod)$coefficients[, 3]
+df <- summary(rlmmod)$df[2]
+pvalues <- pt(abs(tvalues), df, lower.tail=FALSE) * 2
+```
+
+```{r lengthResults, echo=TRUE, cache=TRUE, eval=FALSE}
+tableData <- cbind(summary(rlmmod)$coeff, pvalues)
 colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
 rownames(tableData) <- c("Intercept", "Net Size",  "August", "September", 
-                         "Net Size * August", "Net Size * September")
+                         "Juveniles", "Males", "Unknown Gender",
+                         "Station 10", "Station 2", "Station 3", "Station 4",
+                         "Station 5", "Station 6", "Station 7", "Station 8",
+                         "Station 9")
+
 kable(tableData, digits = 4, format="html")
@@
 
-Associate Editor:
-This is a useful analysis of mysid sampling methods that has implications to Great Lakes and beyond.
-The paper has received two careful reviews.
-Both reviewers were generally positive, but have some suggestions for improvements.
+Although the statistical test is significant, we have very large sample sizes and and the regression coefficients are small relative to the other sources of variation in the model.
+Of greater interest is whether the observed effect is of practical significance.
 
-### Both reviewers wanted to know why you did not use a paired t test when comparing the two net types, which given the paired approach to sampling would appear to be the most appropriate method to analyze the data.
-If you have reasons not to use that approach, please elaborate in your response letter and add a sentence or two to the methods.
+```{r effect size, echo=TRUE, eval=TRUE}
+SD <- sqrt(((sum(mysisLengthData$net == "Large Net") - 1) *
+              sd(mysisLengthData$y[mysisLengthData$net == "Large Net"])^2 + 
+              (sum(mysisLengthData$net == "Small Net") - 1) *
+              sd(mysisLengthData$y[mysisLengthData$net == "Small Net"])^2) / 
+             (sum(mysisLengthData$net == "Large Net") + 
+                sum(mysisLengthData$net == "Small Net")))
 
-We begin the analysis by loading the data used for measuring total count.
-At each sampling event $i$ we recorded the number of mysis shrimp caught and important covariates including date, site, net size, etc.
+d <- (mean(mysisLengthData$y[mysisLengthData$net == "Large Net"]) - 
+        mean(mysisLengthData$y[mysisLengthData$net == "Small Net"])) / SD
+```
 
-```{r totalCount, echo=TRUE, include=TRUE, message=FALSE, cache=TRUE, eval=TRUE}
-load("~/mysis/data/mysisCountData.RData")                                 ## load count data
-```
+We see that there is a significant effect of net size on mean length caught (p = `r I(pvalues[2])`) after controlling for date, gender class, and sampling location.
+Although this is statisticallly significant, the effect size is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`).
+Given a large sample size, a hypothesis test will show staistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05).
+Therefore, the practical effect of a difference in mean length of `r I(summary(rlmmod)$coefficients[2, 1])`mm on a species with a mean length of `r I(mean(mysisLengthData$y))`mm is small (this is the smallest effect of all the effects estimated by almost a factor of two) and a difference in means of this size is not of practical interest.
+Another measure of effect size is Cohen's $d$ which measures the difference in means relative to a pooled standard deviation.
+For our data, Cohen's $d$=`r I(d)`, which implies that the effect of net size is quite small in terms of practical significance.
 
@@
 You suggest that may be the case.
+
+In actuallity, this is quite a complicated statistical test as you are now testing the extreme observations of each sample.
+I prefer to look at the histograms and see that the distributions between net sizes for all of the different subsets of the data are consistent rather than do a formal statistical test (which is much more complicated).
+A quick and dirty extreme value analysis suggests that the model parameters for the generalized extreme value distributions are quite similar for the two net sizes, suggesting that both nets sample the large shrimp equally.
+
+```{r}
+gev.fit(xdat=mysisLengthData$y[mysisLengthData$net == "Small Net"])
+gev.fit(xdat=mysisLengthData$y[mysisLengthData$net == "Large Net"])
+```
 
@@
 
-Lars
+It is true that different sized groups occur at different areas in the reservior, but that is accounted for in the linear regression model (a paired t-test is a speical case of a linear regression model).
+When performing a paired t-test, it requires collapsing each site's length distribution to a mean, which necessarily loses information about the distribution of lengths at each site.
+The linear model above properly accounts for that extra variability in the lengths at each site.
 
@@
 
+By including these factors in the regression model, it controls for their influence.
+
 ### 9)      Line 149 I don't think the mean length is the correct variable of interest.
@@
 I also think the analysis of population structure (male-female, juvenile) and male:female ratio would be more appropriately done by testing for homegenity of proportions and then looking at standardized deviation to see where the nets differed (if at all).
+
+While a contigency table is a possible analysis, the beta regression model is a more flexible model.
+A contingency table to evaluate a homogeniety of proportions is a special case of the beta regression model (just as the t-test is a special case of a regression model).
+I prefer to analyze the data using a regression framework as it is a more general framework that can account for multiple covariates at once (jointly estimating all the effects) instead of having to construct a table and perform analyses one at a time (marginally estimate effects).
 
@@
 
-```{r, echo=TRUE, message=FALSE, results="hide", eval=FALSE}
+```{r, echo=TRUE, message=FALSE, results="hide", eval=TRUE}
 ## load data
@@
 
-```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE, eval=FALSE}
+```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
 X <- model.matrix(~ net * gender + gender * date, data=mysisSexCountData)
@@
 ```
-
-```{r lengthResults, echo=TRUE, cache=TRUE, eval=FALSE}
-tableData <- cbind(summary(rlmmod)$coeff, pvalues)
-colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
-                         "Unknown Gender", "August", "September", 
-                         "Females * August", "Juveniles * August", 
-                         "Unknown Gender * August",
-                         "Females * September", "Juveniles * September", 
-                         "Unknown Gender * September")
-
-kable(tableData, digits = 4, format="html")
-```
-
-```{r effect size, echo=TRUE, eval=FALSE}
-SD <- sqrt(((sum(mysisLengthData$net == "Large Net") - 1) *
-              sd(mysisLengthData$y[mysisLengthData$net == "Large Net"])^2 + 
-              (sum(mysisLengthData$net == "Small Net") - 1) *
-              sd(mysisLengthData$y[mysisLengthData$net == "Small Net"])^2) / 
-             (sum(mysisLengthData$net == "Large Net") + 
-                sum(mysisLengthData$net == "Small Net")))
-
-d <- (mean(mysisLengthData$y[mysisLengthData$net == "Large Net"]) - 
-        mean(mysisLengthData$y[mysisLengthData$net == "Small Net"])) / SD
-```
-We see that there is a significant effect of net size on mean length caught (p = `r I(pvalues[2])`) after controlling for date, gender class, and an interaction between date and gender class.
-Although this is statisticallly significant, the effect size is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`).
-given a large sample size, a hypothesis test will show staistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05).
-Therefore, the practical effect of a difference in mean length of `r I(summary(rlmmod)$coefficients[2, 1])`mm on a species with a mean length of `r I(mean(mysisLengthData$y))`mm is small (this is the smallest effect of all the effects estimated by almost a factor of two) and a difference in means of this size is not of practical interest.

### 2015-12-17 data/processData.R

@@ -1,0 +1,73 @@
+## File to process the CSV files to RData files
+
+##
+## Created 12_15_2015
+## Edited 12_15_2015
+##
+
+##
+## Count Data
+##
+
+data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40, header = TRUE)
+# data_one_way$full ## count per m^2 for full net
+# data_one_way$half ## count per m^2 for half net
+count <- c(data_one_way$full.5,  ## count for full net
+           4 * data_one_way$half.5) ## count for half net adjusted for radius
+date <- rep((data_one_way$month), 2)
+date[date == 7] <- "July"
+date[date == 8] <- "August"
+date[date == 9] <- "September"
+date <- factor(date, levels=c("July", "August", "September"))
+net <- factor(c(rep("Large Net", 40), rep("Small Net", 40)), levels=c("Small Net", "Large Net"))
+station <- factor(data_one_way$Station)
+mysisCountData <- data.frame(count=count, date=date, net=net, station=station)
+
+## Save RData file
+save(mysisCountData, file="~/mysis/data/mysisCountData.RData")   
+## remove all data to avoid conflicts
+rm(list=ls())
+
+## 
+## Length Data
+##
+
+data_length <- read.csv("~/mysis/data/lengthData.csv", skip = 4, header = TRUE)
+net <- data_length$Size..m.
+net[net== 0.5] <- "Small Net"
+net[net== 1.0] <- "Large Net"
+net <- factor(net, levels=c("Small Net", "Large Net"))
+y <- data_length$Length..mm.
+## remove whitespace from string, then truncate into 
+## broad gender classes, not subclasses
+library(stringr)
+gender <- as.factor(substr(str_trim(as.character(data_length$Gender)), 0, 1))
+
+date <- as.numeric(data_length$Date)
+date[date == 2] <- 1      ## collapse into monthly values
+date[date == 3] <- 2
+date[date == 4] <- 3
+date <- as.factor(date)
+plot(as.numeric(date), type='l', main = "Is this a data error? - yes?")
+## correct date mislabeling in the raw data
+date <- as.numeric(data_length$Date)
+date[7330:7519] <- 4
+date[7559:7773] <- 4
+date[date == 2] <- 1      ## collapse into monthly values
+date[date == 3] <- 2
+date[date == 4] <- 3
+# date <- as.factor(date)
+plot(as.numeric(date), type='l', main = "Data looks better")
+date[date == 1] <- "July"
[…]

### 2016-01-04 responseToReviewers/mysisReviewerResponse.Rmd
updated writeup

@@
 ---
+title: "Effect of net size on estimates of abundance, size, age and sex ratio of Mysis diluviana Journal of Great Lakes Research - Statistical Analysis"
+author: "John Tipton"
+date: "January 3, 2016"
 output: word_document
----
-
-title: "Response to reviewer's statistical comments on the paper: GLR-D-15-00174 - Effect of net size on estimates of abundance, size, age and sex ratio of Mysis diluviana Journal of Great Lakes Research"
-author: "John Tipton"
-date: "October 12, 2015"
 ---
@@
 library(knitr)
-knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
+# knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
+knitr::opts_chunk$set(echo=FALSE)
+knitr::opts_chunk$set(include=FALSE)
+knitr::opts_chunk$set(message=FALSE)
+knitr::opts_chunk$set(warning=FALSE)
 library(ggplot2)
@@
 library(lme4)
+library(ismev)
+library(car)
+library(betareg)
 ```
 
-Associate Editor:
-This is a useful analysis of mysid sampling methods that has implications to Great Lakes and beyond.
-The paper has received two careful reviews.
-Both reviewers were generally positive, but have some suggestions for improvements.
+We tested many of the population characteristics to see if there was any difference between the two sized nets.
+Our main questions of interest are 1) Is there a difference in abundance of Mysids caught between the two nets, 2) Is there a difference in mean length of Mysids caught between the two nets, and 3) Does the count when broken down by age and sex class differ between the two net sizes.
 
-### Both reviewers wanted to know why you did not use a paired t test when comparing the two net types, which given the paired approach to sampling would appear to be the most appropriate method to analyze the data.
-If you have reasons not to use that approach, please elaborate in your response letter and add a sentence or two to the methods.
+1) 
+We test if the abundance of Mysids caught differs between the two nets.
+By looking at the distributions of counts across time, we see that they are similar between net sizes, but the normality of the counts is questionable.
 
-We begin the analysis by loading the data used for measuring total count.
-At each sampling event we recorded the number of mysis shrimp caught and important covariates including date, site, net size, etc.
-
-```{r totalCount, echo=TRUE, include=TRUE, message=FALSE, cache=TRUE, eval=TRUE}
+```{r totalCount, eval=TRUE}
 load("~/mysis/data/mysisCountData.RData")                                 ## load count data
@@
 
-By looking at the distributions of counts across time, we see that they are similar between net sizes, but the normality of the counts is questionable.
-
-```{r plotCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", eval=FALSE}
+```{r plotCount, eval=TRUE, include=TRUE}
 ggplot(data=mysisCountData, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
@@
 ```
-Because counts cannot be less that zero and often are right skewed, the assumption of Normality in the data is questionable.
-Hence, we don't want to use Gaussian methods like linear regression and ANOVA.
-But, for completeness, let's look a a paired t-test and see what the results look like with the knowledge that this analysis is less than satisfactory.
+
+From the histogram, we see little differences between the two net sizes, except for perhaps a spike in catches of about 120 by the small net in August.
+Because counts cannot be less that zero and often are right skewed, the assumption of normality in the data is questionable.
+Hence, methods like linear regression, t-tests, and ANOVA that assume normal distributions are not ideal, but, for completeness we begin with a paired t-test.
 
@@
 
-We can also look at a paired t-test for the mean length of the nets, although this loses alot of information because we are only testing the sample level means without accounting for the differences in uncertainties about the site level means due to different sample sizes between nets
+```{r totalCountPoisson, message=FALSE}
+mns <- by(mysisCountData[, "count"], mysisCountData[, "net"], mean)       ## mean of counts by net size
+vars <- by(mysisCountData[, "count"], mysisCountData[, "net"], var)       ## variance of counts by net size
+```
+
+We test for a difference in counts between the nets formally by fitting a model to compare if the count of Mysids caught by the large net is the same as the count of Mysids caught by the large net, after normalizing for the area of the two nets and controlling for covariates.
+The class of models known as generalized linear models (glms) offers a rigorous solution for modeling count data.
+These models allow for regression models that control for the effects of covariates like linear regression but do not assume a normal distribution.
+For count data, there are two natural distributions to use: the Poisson distribution and the negative binomial distribution.
+The Poisson distribution has the strong assumption that the mean equals the variance, which is often not met in practical datasets.
+In our data, the count means are $(`r I(mns)`)$ and the count variances are $(`r I(vars)`)$, for the small and large nets, respectively.
+Therefore we use a negative binomial model
+$$
+\label{neg:bin}
+y_i \sim \mbox{NegBin}(\mu_i, \phi),
+$$
+where $\mu_i$ is the mean of the negative binomial distribution for sampling event $i$ and $\phi$ is an overdispersion parameter that allows for the mean and variance to be different.
+We model $\mu_i$ with covariates using the canonical log link function
+$$ 
+\mu_i = log(\mathbf{X}_i \boldsymbol{\beta}),
+$$
+where $\mathbf{X}_i$ is the set of covariates for observation $i$.
+Then we perform inference on our coefficients $\boldsymbol{\beta}$, where the interpretation of $\beta_j$ is a percent change in count per unit change in $X_{ij}$.
+
+```{r totalCountModel, eval=TRUE, results="hide"}
+summary(nbmod <- glm.nb(count ~ net + date + station, data=mysisCountData))
+## get coefficient estimates
+coefs <- c(nbmod$coefficients, nbmod$theta)           
+```
+
+To test for a change in total counts between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and sampling location.
+The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes ($z$-value = `r I(round(summary(nbmod)$coef[2, 3], 4))`, $p$-value=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
+
+```{r countModelResults, eval=FALSE}
+tableData <- cbind(summary(nbmod)$coeff[1:6, ])
+colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
+rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
+                         "Net Size * August", "Net Size * September")
+kable(tableData, digits = 4, format="html")
+```
+
+2) 
+The next population characteristic of interest is whether we are catching the same size Mysids with each net. First we examine some histograms and get an idea of what the length data look like.
 
@@
 
-First lets look at some histograms and get an idea of what the data look like.
-
-```{r, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold", eval=FALSE}
+```{r, eval=TRUE, include=TRUE}
 ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ date) +
@@
 
-```{r, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold", eval=FALSE}
+```{r, eval=TRUE, include=TRUE}
 ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ gender) + 
@@
 
+Next, we plot histograms of the sampling events (80 sampling events – 10 sites, 2 nets, 4 dates) as a paired data set.
+
 ```{r pairedTTestLength}
@@
 y1 <- y1[!is.na(y1)]
+
 y2 <- by(mysisLengthData$y[mysisLengthData$net == "Large Net"], 
@@
 pairttestlength <- t.test(y1, y2, paired=TRUE)
-layout(matrix(1:2,2,1))
-hist(y1)
-hist(y2)
+qplot(y1-y2, xlab="difference in means")
 ```
 
-From the paired t-test, we find that there is a significant difference ($t$ = `r pairttest$statistic`, $df$ = `r pairttest$parameter`, $p$-value = `r pairttest$p.value` ) in expected counts, although the $p$-value is close to the 0.05 level.
+From the histogram of paired differences, it looks like there is some light evidence that the smaller net catches slightly smaller shrimp, although the distribution does seems to be centered near 0.
+From the paired t-test, we find that there is a significant difference ($t$ = `r pairttestlength$statistic`, $df$ = `r pairttestlength$parameter`, $p$-value = `r pairttestlength$p.value` ) in expected counts, although the $p$-value is close to the 0.05 level.
 Despite the statistically significant difference, the practical difference is quite small as judged by the histograms and small difference in shrimp size.
 
-To better account for the different sample sizes and other ....
+```{r pairedTTestLengthtrim}
+y1trim <- by(mysisLengthData$y[mysisLengthData$net == "Small Net"],
+             mysisLengthData$label[mysisLengthData$net == "Small Net"], mean, trim=0.05)
+y1trim <- y1trim[!is.na(y1trim)]
+
+y2trim <- by(mysisLengthData$y[mysisLengthData$net == "Large Net"],
+             mysisLengthData$label[mysisLengthData$net == "Large Net"], mean, trim=0.05)
+y2trim <- y2trim[!is.na(y2trim)]
+pairttestlengthtrim <- t.test(y1trim, y2trim, paired=TRUE)
+qplot(y1trim-y2trim, xlab="difference in trimmed means")
+```
+
+One could argue that because the larger net catches four times the number of shrimp, the large net is more likely to catch shrimp at the extremes of the size classes (either really small or really large Mysids).
+To account for this, we perform a paired t-test using a trimmed mean, removing the smallest and largest 5% of the lengths before calculating the sample mean.
+From the paired t-test using the trimmed means, we find that there is not a significant difference ($t$ = `r pairttestlengthtrim$statistic`, $df$ = `r pairttestlengthtrim$parameter`, $p$-value = `r pairttestlengthtrim$p.value` ) in expected counts, although the $p$-value is close to the 0.05 level.
+As with the untrimmed means, the practical difference is quite small as judged by the histograms and small difference in shrimp size.
+
+To better account for the different sample sizes (the group means are from unbalanced sample sizes), we fit a linear model and a robust linear model (that assumes overdispersion in the data).
+We start by checking for heterogeniety in the group variances.
 
 ```{r}
+vars <- as.vector(by(mysisLengthData$y, mysisLengthData$label, var))
+qplot(vars, main = "var by sampling location", xlab="var")
+bart_test <- bartlett.test(mysisLengthData$y, mysisLengthData$label)
+levene_test <- leveneTest(mysisLengthData$y, mysisLengthData$label)
+```
 
-X <- model.matrix(~ net + date + gender, data=mysisLengthData)
-idx <- attr(uniquecombs(X), "index")
+When looking at the distribution of sampling variance by sampling occasion (10 locations, 2 nets, 4 different dates gives 80 total sampling ocassions), we see that there is some difference in variance of lengths by sampling occasion and Bartlett's test ($p$-value = `r bart_test$p.value`) and Levene's test ($p$-values = `r levene_test$Pr[1]`).
+
+To test if there is a difference in length of net sizes while accounting for heterogeneity, we use a weighted least squares regression where the weights are the inverse of the group variances.
+
+```{r}
+## weighted linear regression
+samps <- by(rep(1, length(mysisLengthData$y)), mysisLengthData$label, sum)
+w <- rep(0, length(mysisLengthData$y))
+for (i in 1:length(unique(mysisLengthData$label))) {
+  if (i==1) {
+    w[1:samps[1]] <- rep(1 / vars[i], samps[i])
+  } else {
+    w[(sum(samps[1:(i-1)])+1):sum(samps[1:i])] <- rep(1 / vars[i], samps[i])
+  }
+}
+summary(weightedlm <- lm(y ~ net + date + gender + station, 
+                          data=mysisLengthData, weights=w))
+```
+
+From the weighted least squares regression, we find that there is a statistically signficant difference in mean Mysid length caught between the two nets ($t$ = `r summary(weightedlm)$coefficients[2, 3]`, $p$-value = `r summary(weightedlm)$coefficients[2, 4]`).
+Now, after accounting for the location, date, sex distribution, and heterogeneity of variance we find that the difference in mean length caught between the two net sizes is `r summary(weightedlm)$coefficients[2, 1]` mm.
+
+As a final analysis, we apply a robust regression, using an M-estimator model that accounts for heterogeniety in variance and presence of outlying observations.
+
+```{r}
 summary(lmmod <- lm(y ~ net + date + gender + station, data=mysisLengthData))
-library(lme4)
-# summary(lmemod <- lmer(y ~ net * date + net * gender +
-#                          date * gender + (1|station), data=mysisLengthData))
-# summary(lmemod <- lmer(y ~ net + date * gender + (1|station),
-#                        data=mysisLengthData))
-# summary(rlmmod <- rlm(y ~ net * date + net * gender + 
-#                         date * gender, data=mysisLengthData))))
-summary(rlmmod <- rlm(y ~ net + date + gender + station, data=mysisLengthData))
-tvalues <- summary(rlmmod)$coefficients[, 3]
+summary(rlmmod <- rlm(y ~ net + date + gender + station, data=mysisLengthData, w=w))
+tvalues <- summary(rlmmod)$coefficients[2, 3]
 df <- summary(rlmmod)$df[2]
@@
 
-```{r lengthResults, echo=TRUE, cache=TRUE, eval=FALSE}
+```{r lengthResults, eval=TRUE}
 tableData <- cbind(summary(rlmmod)$coeff, pvalues)
-colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size",  "August", "September", 
-                         "Juveniles", "Males", "Unknown Gender",
-                         "Station 10", "Station 2", "Station 3", "Station 4",
-                         "Station 5", "Station 6", "Station 7", "Station 8",
-                         "Station 9")
-
-kable(tableData, digits = 4, format="html")
 ```
 
-Although the statistical test is significant, we have very large sample sizes and and the regression coefficients are small relative to the other sources of variation in the model.
-Of greater interest is whether the observed effect is of practical significance.
+For the robust linear model, we still find statistically significant differences in mean length between the net sizes ($t$-value = `r tvalues`, $p$-value = `r pvalues` on `r df` degrees of freedom), however the effect size `r summary(rlmmod)$coefficients[2, 1]` is still about the same as from the weighted linear model
 
-```{r effect size, echo=TRUE, eval=TRUE}
+```{r effect size, eval=TRUE}
 SD <- sqrt(((sum(mysisLengthData$net == "Large Net") - 1) *
@@
 
-We see that there is a significant effect of net size on mean length caught (p = `r I(pvalues[2])`) after controlling for date, gender class, and sampling location.
-Although this is statisticallly significant, the effect size is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`).
-Given a large sample size, a hypothesis test will show staistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05).
+Although the statistical tests using both the weighted linear model and the robust linear model show statistically significant differences in mean length caught between the net sizes, this is not unexpected because have very large sample sizes and and the regression coefficients are very small relative to the other sources of variation in the model.
+Of greater interest is whether the observed effect is of practical significance.
+
+The effect size of the mean length difference between the two net sizes is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`).
+Given a large sample size, a hypothesis test will show statistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05).
 Therefore, the practical effect of a difference in mean length of `r I(summary(rlmmod)$coefficients[2, 1])`mm on a species with a mean length of `r I(mean(mysisLengthData$y))`mm is small (this is the smallest effect of all the effects estimated by almost a factor of two) and a difference in means of this size is not of practical interest.
@@
 For our data, Cohen's $d$=`r I(d)`, which implies that the effect of net size is quite small in terms of practical significance.
+
+3) 
+The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.
+A visual inspection of counts by sex class shows no difference in counts by net size, but does show a change in counts over time.
+
+The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.
+
+```{r, eval=TRUE, include=TRUE}
+## load data
+load("~/mysis/data/mysisSexCountData.RData")   
+ggplot(data=mysisSexCountData, aes(count)) +         
+  geom_histogram() + facet_grid(net ~ gender) + 
+  ggtitle("Histogram of normalized counts by sex and net size")
+```
+
+From the histogram of counts broken down by sex class and net size, we see no outstanding differences of counts between net size.
+
+```{r, eval=TRUE, include=TRUE}
+ggplot(data=mysisSexCountData, aes(count)) +
+  geom_histogram() + facet_grid(gender ~ date) + 
+  ggtitle("Histogram of normalized counts by sex and date")
+```
+
+From the histogram of sex class by date, we see a pattern of juveniles maturing to males and females as time progresses and an increase in unknowns as juveniles grow in size but are not sexually differentiated.
+
+```{r, message=FALSE, results="hide"}
+## Normalize the counts <- divide by 4 because of the zeros for the small net
+mysisSexCountData$count[mysisSexCountData$net == "Large Net"] <- 
+  mysisSexCountData$count[mysisSexCountData$net == "Large Net"] / 4
+## Note: this gives warnings in the model as there are now non-integers
+summary(nbmod <- glm.nb(count ~ net + date + station + gender, data = mysisSexCountData))
+## Now fit the model with rounded counts
+mysisSexCountData$count[mysisSexCountData$net == "Large Net"] <- 
+  round(mysisSexCountData$count[mysisSexCountData$net == "Large Net"])
+## Note: this no longer gives a warning
+summary(nbmod <- glm.nb(count ~ net + date + station + gender, data = mysisSexCountData))
+```
+
+ To test for a change in counts broken down by sex category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling, sampling location, and sex class.
+The results show that there is no effect of net size on counts broken down by sex class ($z$-value = `r summary(nbmod)$coeff[2, 3]`, $p$-value = `r summary(nbmod)$coeff[2, 4]`).
+
+We are also interested if the proportion of juveniles is different between the two net sizes.
+Because the data are proportions, we cannot use methods that assume normally distributed data.
+One could use a contingency table to test homogeneity of proportions, but we prefer to use a glm known as beta regression (where a contingency table for homogeneity of proportions is a special case of beta regression).
+
+```{r, eval=TRUE, include=TRUE}
+load("~/mysis/data/mysisJuvenileData.RData") 
+ggplot(data=mysisJuvenileData, aes(prop)) + geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Histogram of proportion of juveniles by net size")
+```
+
+Looking at the histogram of proportions of juveniles by net size and by date, we see that there is more variability in the proportions for the small net (which is expected because the counts are smaller for the small net and therefore the proportions are more variable).
+We can see that the proportion of juveniles decreases through time and appears similar between the two net sizes.
+
+```{r, message=FALSE, eval=TRUE}
+brmod <- betareg(prop ~ net + date + station, data=mysisJuvenileData)
+```
+
+To test for a change in the proportion of juveniles caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and sampling location.
+The results shown below show that there is no evidence of an effect on the proportion of juveniles caught between the two net sizes (z=`r I(round(summary(brmod)$coef$mean[2, 3], 4))`, p=`r I(round(summary(brmod)$coef$mean[2, 4], 4))`).
+
+Also of interest is whether the ratio of females to males caught is the same between the two net sizes.
+Like the model for the proportion of juveniles, we could use a contingency table for heterogeneity of proportions, but we instead use the more flexible beta regression model.
+
+```{r, eval=TRUE, include=TRUE}
+load("~/mysis/data/mysisSexData.RData")
+ggplot(data=mysisSexData, aes(prop)) + geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Histogram of proportion of females by net size")
+```
+
+Looking at the histogram of proportions of females by net size and by date, we see that there is more variability in the proportions for the small net (which is expected because the counts are smaller for the small net and therefore the proportions are more variable).
+We can see that the proportion of females appears to increase through time and appears similar between the two net sizes.
+
+```{r, message=FALSE, eval=TRUE}
+brmodsex <- betareg(prop ~ net + date + station, data=mysisSexData)
+```
+
+To test for a change in the proportion of females caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and sampling location.
+The model shows that there is no evidence of an effect on the proportion of females caught between the two net sizes (z=`r I(round(summary(brmodsex)$coef$mean[2, 3], 4))`, p=`r I(round(summary(brmodsex)$coef$mean[2, 4], 4))`).
 
@@
 
-In actuallity, this is quite a complicated statistical test as you are now testing the extreme observations of each sample.
+Actually, this is quite a complicated statistical test as you are now testing the extreme observations of each sample.
 I prefer to look at the histograms and see that the distributions between net sizes for all of the different subsets of the data are consistent rather than do a formal statistical test (which is much more complicated).
@@
 ```{r}
-gev.fit(xdat=mysisLengthData$y[mysisLengthData$net == "Small Net"])
-gev.fit(xdat=mysisLengthData$y[mysisLengthData$net == "Large Net"])
+# gev.fit(xdat=mysisLengthData$y[mysisLengthData$net == "Small Net"])
+# gev.fit(xdat=mysisLengthData$y[mysisLengthData$net == "Large Net"])
 ```
@@
 
-It is true that different sized groups occur at different areas in the reservior, but that is accounted for in the linear regression model (a paired t-test is a speical case of a linear regression model).
+It is true that different sized groups occur at different areas in the reservoir, but that is accounted for in the linear regression model (a paired t-test is a special case of a linear regression model).
 When performing a paired t-test, it requires collapsing each site's length distribution to a mean, which necessarily loses information about the distribution of lengths at each site.
 The linear model above properly accounts for that extra variability in the lengths at each site.
+See the comments above.
 
-## Reviewers' comments:
-
-## Reviewer #1: Review of GLR-D-15-00174
+# Reviewer #1: Review of GLR-D-15-00174
 
@@
 
-While a contigency table is a possible analysis, the beta regression model is a more flexible model.
-A contingency table to evaluate a homogeniety of proportions is a special case of the beta regression model (just as the t-test is a special case of a regression model).
+While a contingency table is a possible analysis, the beta regression model is a more flexible model.
+A contingency table to evaluate a homogeneity of proportions is a special case of the beta regression model (just as the t-test is a special case of a regression model).
 I prefer to analyze the data using a regression framework as it is a more general framework that can account for multiple covariates at once (jointly estimating all the effects) instead of having to construct a table and perform analyses one at a time (marginally estimate effects).
@@
 
+We provide the data and the code in the appendix so the reader can evaluate the statistical analysis in its entirety.
+
 ### 13)     Line 177 also caught more of the smallest individuals
 
-## Reviewer #2: This manuscript has one straightforward message that will be useful to researchers interested in the collection of mysids, in that it supports the use of smaller diameter nets, which will be convenient from small vessels often used in lakes these animals have spread into.
+# Reviewer #2: This manuscript has one straightforward message that will be useful to researchers interested in the collection of mysids, in that it supports the use of smaller diameter nets, which will be convenient from small vessels often used in lakes these animals have spread into.
 Overall this manuscript is acceptable for publication with some revisions.
 The details of the statistical tests need to be expanded upon and some language that downplays some of the size-distribution differences with different net sizes should be discussed in greater detail.
-
-Major points:
 
@@
 Fig. 3 The 1.0 m diameter net bars should be white to be consistent with the other graph. Font may have to be larger for the axes and the inset pie-graphs.
-
-This document shows the analyses performed in Johnson et. al.
-
-## Model for counts by sex
-
-The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.
-To do this we use (\ref{eq:negBin}) above.
-We begin by loading the data where counts are divided into sex classes.
-
-```{r, echo=TRUE, message=FALSE, results="hide", eval=TRUE}
-## load data
-load("~/mysis/data/mysisSexCountData.RData")   
-## construct model matrix for covariates
-X <- model.matrix(~ net * gender + net * date + gender * date-1, 
-                  data=mysisSexCountData)
-## fit model
-nbmod <- glm.nb(y ~ ., data = data.frame(y=mysisSexCountData$count, X))
-```
-
-To explore this data in more detail we examine two sets of histograms, the counts of each sex class broken down by date in figure1 and the counts of each sex class broken down by net size in figure2.
-From these we see...
-
-```{r plot SexCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold", eval=FALSE}
-ggplot(data=mysisSexCountData, aes(count)) +         
-  geom_histogram() + facet_grid(net ~ gender) + 
-  ggtitle("Histogram of normalized counts by sex and net size")
-ggplot(data=mysisSexCountData, aes(count)) +
-  geom_histogram() + facet_grid(gender ~ date) + 
-  ggtitle("Histogram of normalized counts by sex and date")
-```
-
-To test for a change in counts broken down by gender category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling, the interaction between net size and sampling date, and the interaction between sampling date and gender.
-The results shown below show that there is no evidence of an interaction between net size and sampling date and no evidence of an interaction between net size and gender.
-
-```{r sexResults, echo=TRUE, cache=TRUE, eval=FALSE}
-tableData <- cbind(summary(nbmod)$coeff)
-colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
-                         "Unknown Gender", "August", "September", 
-                         "Net Size * Females", "Net Size * Juveniles", 
-                         "Net Size * Unknowns", "Net Size * August",
-                         "Net Size * September",
-                         "Females * August", "Juveniles * August", 
-                         "Unknown Gender * August",
-                         "Females * September", "Juveniles * September", 
-                         "Unknown Gender * September")
-kable(tableData, digits = 4, format="html")
-```
-
-Thus, we refit the model without the interaction between net size and sampling date. To do this we start by loading the new covariate matrix $\mathbf{X}$.
-
-```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
-X <- model.matrix(~ net * gender + gender * date, data=mysisSexCountData)
-nbmod <- glm.nb(y ~ . -1, data = data.frame(y=mysisSexCountData$count, X))
-```
-
-On this reduced model, the results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
-```{r sexResultsReduced, echo=TRUE, cache=TRUE, eval=FALSE}
-tableData <- cbind(summary(nbmod)$coeff)
-colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
-                         "Unknown Gender", "August", "September", 
-                         "Net Size * Females", "Net Size * Juveniles", 
-                         "Net Size * Unknowns", 
-                         "Females * August", "Juveniles * August", 
-                         "Unknown Gender * August",
-                         "Females * September", "Juveniles * September", 
-                         "Unknown Gender * September")
-kable(tableData, digits = 4, format="html")
-```
-
-## Model for proportion of juveniles
-```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
-set.seed(123)
-data_one_way <- read.csv("~/mysis/data/ttestData.csv", skip = 3, nrows = 40,
-                         header = TRUE, stringsAsFactors=FALSE)
-
-prop <- c(as.numeric(data_one_way$full.2),    ## proportion of females full net
-             as.numeric(data_one_way$half.2)) ## proportion of females half net
-N <- length(prop)
-## Set improper values to NA
-
-date <- rep(as.factor(data_one_way$month), 2)
-net <- as.factor(c(rep("full", 40), rep("half", 40)))
-data <- data.frame(prop=prop, date=date, net=net)
-levels(data$date) <- c("July", "August", "Septempber")
-levels(data$net) <- c("Small Net", "Large Net")
-X <- model.matrix(~ net * date)
-library(betareg)
-brmod <- betareg(prop ~ . -1, data = data.frame(prop, X))
-summary(brmod)
-```
-
-```{r plotJuviCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", eval=FALSE}
-ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .) + 
-  ggtitle("Histogram of proportion of juveniles by net size")
-```
-
-To test for a change in the proportion of juveniles caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
-The results shown below show that there is no evidence of an effect on the proportion of juveniles caught between the two net sizes (p=`r I(round(summary(brmod)$coef$mean[2, 4], 4))`).
-
-```{r juvenileResults, echo=TRUE, cache=TRUE, eval=FALSE}
-tableData <- cbind(summary(brmod)$coeff$mean)
-colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
-                         "Net Size * August", "Net Size * September")
-kable(tableData, digits = 4, format="html")
-```
-
-## Length model
-
-```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE, eval=TRUE}
-# data_length <- read.csv("~/mysis/data/lengthData.csv", skip = 4, header = TRUE)
-# net <- data_length$Size..m.
-# y <- data_length$Length..mm.
-# gender <- as.factor(substr(str_trim(as.character(data_length$Gender)), 0, 1))
-# station <- data_length$Station
-# date <- as.numeric(data_length$Date)
-# date[7330:7519] <- 4
-# date[7559:7773] <- 4
-# date[date == 2] <- 1      ## collapse into monthly values
-# date[date == 3] <- 2
-# date[date == 4] <- 3
-# date <- as.factor(date)
-# # plot(as.numeric(date), type='l', main = "Data looks better")
-# data <- data.frame(y=y, date=date, net=as.factor(net), gender=gender)
-# levels(data$date) <- c("July", "August", "Septempber")
-# levels(data$net) <- c("Small Net", "Large Net")
-# levels(data$gender) <- c("Female", "Juvenile", "Male", "Unknown")
-
-load("~/mysis/data/mysisLengthData.RData")
-X <- model.matrix(~ net + date + gender, data=mysisLengthData)
-idx <- attr(uniquecombs(X), "index")
-summary(lmmod <- lm(y ~ net + date * gender, data=mysisLengthData))
-library(lme4)
-# summary(lmemod <- lmer(y ~ net * date + net * gender +
-#                          date * gender + (1|station), data=mysisLengthData))
-# summary(lmemod <- lmer(y ~ net + date * gender + (1|station),
-#                        data=mysisLengthData))
-# summary(rlmmod <- rlm(y ~ net * date + net * gender + 
-#                         date * gender, data=mysisLengthData))))
-summary(rlmmod <- rlm(y ~ net + date * gender, data=mysisLengthData))
-tvalues <- summary(rlmmod)$coefficients[, 3]
-df <- summary(rlmmod)$df[2]
-pvalues <- pt(abs(tvalues), df, lower.tail=FALSE) * 2
-```
-
-```{r plotLength, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold", eval=FALSE}
-ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ date) +
-  ggtitle("Lengths by net size")
-ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ gender) + 
-  ggtitle("Lengths by sex and net size")
-ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(date ~ gender) + 
-  ggtitle("Lengths by sex and date")
-```

### 2016-01-04 _drafts/mysisAnalysis.Rmd
updated writeup

--- 2015-10-12 _drafts/mysisAnalysis.Rmd
@@
 ---
-title: "Mysis analysis"
+title: "Effect of net size on estimates of abundance, size, age and sex ratio of Mysis diluviana Journal of Great Lakes Research - Statistical Analysis"
 author: "John Tipton"
-date: "September 9, 2015"
+date: "January 3, 2016"
 output: html_document
 ---
-This document shows the analyses performed in Johnson et. al. We start by loading packages and define helper functions
 
-```{r loadLibraries, echo=TRUE, include=FALSE, message=FALSE}
+```{r loadLibraries, include=TRUE, message=FALSE}
+## We start by loading packages and define helper functions
 options(digits=3)
 set.seed(123)
-require(knitr)
-knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
-require(ggplot2)
-require(Gmisc)
-require(MASS)
-require(stringr)
-require(mgcv)
-require(lme4)
-
+library(knitr)
+# knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
+knitr::opts_chunk$set(echo=TRUE)
+knitr::opts_chunk$set(eval=TRUE)
+knitr::opts_chunk$set(tidy=TRUE)
+knitr::opts_chunk$set(include=TRUE)
+knitr::opts_chunk$set(message=FALSE)
+knitr::opts_chunk$set(warning=FALSE)
+library(ggplot2)
+library(Gmisc)
+library(MASS)
+library(stringr)
+library(mgcv)
+library(lme4)
+library(ismev)
+library(car)
+library(betareg)
 ```
 
-## Model for Total counts for each net
-We begin the analysis by loading the data used for measuring total count.
-At each sampling event $i$ we recorded the number of mysis shrimp caught and important covariates including date, site, net size, etc.
-```{r totalCount, echo=TRUE, include=TRUE, message=FALSE, cache=TRUE}
+We tested many of the population characteristics to see if there was any difference between the two sized nets.
+Our main questions of interest are 1) Is there a difference in abundance of Mysids caught between the two nets, 2) Is there a difference in mean length of Mysids caught between the two nets, and 3) Does the count when broken down by age and sex class differ between the two net sizes.
+
+1) 
+We test if the abundance of Mysids caught differs between the two nets.
+By looking at the distributions of counts across time, we see that they are similar between net sizes, but the normality of the counts is questionable.
+
+```{r totalCount, eval=TRUE}
 load("~/mysis/data/mysisCountData.RData")                                 ## load count data
+```
+
+```{r plotCount, eval=TRUE, include=TRUE}
+ggplot(data=mysisCountData, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Histogram of normalized counts by net size and date")
+```
+
+From the histogram, we see little differences between the two net sizes, except for perhaps a spike in catches of about 120 by the small net in August.
+Because counts cannot be less that zero and often are right skewed, the assumption of normality in the data is questionable.
+Hence, methods like linear regression, t-tests, and ANOVA that assume normal distributions are not ideal, but, for completeness we begin with a paired t-test.
+
+```{r pariedTTest}
+y1 <- mysisCountData$count[mysisCountData$net == "Small Net"]
+y2 <- mysisCountData$count[mysisCountData$net == "Large Net"]
+pairttest <- t.test(y1, y2, paired=TRUE)
+```
+
+From the paired t-test, we find that there is no significant difference ($t$ = `r pairttest$statistic`, $df$ = `r pairttest$parameter`, $p$-value = `r pairttest$p.value` ) in expected counts.
+
+```{r totalCountPoisson, message=FALSE}
 mns <- by(mysisCountData[, "count"], mysisCountData[, "net"], mean)       ## mean of counts by net size
@@
 
-After loading the data, we can look at histograms of data to get a better understanding of what variables effect mysis shrimp count.
-First we examine the total counts broken down by net size and sampling month
-
-```{r plotCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center"}
-ggplot(data=mysisCountData, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
-  ggtitle("Histogram of normalized counts by net size and date")
-```
-
-From these histograms, we see little differences between the two net sizes, except for perhaps a spike in catches of about 120 by the small net in August.
-We test this formally by fitting a model to compare if the count of shrimp caught by the large net is the same as the count of shrimp caught by the large net, after normalizing for the area of the two nets and controlling for covariates.
-Because counts cannot be less that zero and often are right skewed, the assumption of Normality in the data is questionable.
-Hence, we don't want to use Gaussian methods like linear regression and ANOVA.
-Luckily, the class of models known a generalized linear models (or glms) offers a solution.
-These models allow for regression models that control for the effects of covariates like linear regression but do not assume the normal distribution.
+We test for a difference in counts between the nets formally by fitting a model to compare if the count of Mysids caught by the large net is the same as the count of Mysids caught by the large net, after normalizing for the area of the two nets and controlling for covariates.
+The class of models known as generalized linear models (glms) offers a rigorous solution for modeling count data.
+These models allow for regression models that control for the effects of covariates like linear regression but do not assume a normal distribution.
 For count data, there are two natural distributions to use: the Poisson distribution and the negative binomial distribution.
@@
 $$
-where $\mu$ is the mean of the negative binomial distribution and $\phi$ is an overdispersion parameter that allows for the mean and variance to be different.
-We model $\mu$ with covariates using a log link function
+where $\mu_i$ is the mean of the negative binomial distribution for sampling event $i$ and $\phi$ is an overdispersion parameter that allows for the mean and variance to be different.
+We model $\mu_i$ with covariates using the canonical log link function
 $$ 
@@
 Then we perform inference on our coefficients $\boldsymbol{\beta}$, where the interpretation of $\beta_j$ is a percent change in count per unit change in $X_{ij}$.
-```{r totalCountModel, echo=TRUE, eval=TRUE, results="hide"}
-## construct covariates of net size, date and their interaction
-X <- model.matrix(~ net * date - 1, data=mysisCountData)  
-## fit negative binomial model
-summary(nbmod <- glm.nb(y ~ ., data = data.frame(y=mysisCountData$count, X)))
+
+```{r totalCountModel, eval=TRUE, results="hide"}
+summary(nbmod <- glm.nb(count ~ net + date + station, data=mysisCountData))
 ## get coefficient estimates
@@
 
-To test for a change in total counts between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
-The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
+To test for a change in total counts between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and sampling location.
+The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes ($z$-value = `r I(round(summary(nbmod)$coef[2, 3], 4))`, $p$-value=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
 
-```{r countModelResults, echo=TRUE}
+```{r countModelResults, eval=FALSE}
 tableData <- cbind(summary(nbmod)$coeff[1:6, ])
@@
                          "Net Size * August", "Net Size * September")
-kable(tableData, digits = 4, format="pandoc")
+kable(tableData, digits = 4, format="html")
 ```
 
-## Model for counts by sex
+2) 
+The next population characteristic of interest is whether we are catching the same size Mysids with each net. First we examine some histograms and get an idea of what the length data look like.
 
-The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.
-To do this we use (\ref{eq:negBin}) above.
-We begin by loading the data where counts are divided into sex classes.
-
-```{r, echo=TRUE, message=FALSE, results="hide"}
-## load data
-load("~/mysis/data/mysisSexCountData.RData")   
-## construct model matrix for covariates
-X <- model.matrix(~ net * gender + net * date + gender * date-1, 
-                  data=mysisSexCountData)
-## fit model
-nbmod <- glm.nb(y ~ ., data = data.frame(y=mysisSexCountData$count, X))
+```{r lengthData}
+load("~/mysis/data/mysisLengthData.RData")
 ```
 
-To explore this data in more detail we examine two sets of histograms, the counts of each sex class broken down by date in figure1 and the counts of each sex class broken down by net size in figure2.
-From these we see...
-
-```{r plot SexCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold"}
-ggplot(data=mysisSexCountData, aes(count)) +         
-  geom_histogram() + facet_grid(net ~ gender) + 
-  ggtitle("Histogram of normalized counts by sex and net size")
-ggplot(data=mysisSexCountData, aes(count)) +
-  geom_histogram() + facet_grid(gender ~ date) + 
-  ggtitle("Histogram of normalized counts by sex and date")
+```{r, eval=TRUE, include=TRUE}
+ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ date) +
+  ggtitle("Lengths by net size")
 ```
 
-To test for a change in counts broken down by gender category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling, the interaction between net size and sampling date, and the interaction between sampling date and gender.
-The results shown below show that there is no evidence of an interaction between net size and sampling date and no evidence of an interaction between net size and gender.
+When looking at the distribution of lengths between the two net sizes, we see that the histograms look quite similar across the different dates.
+What stands out is that the small net catches about a quarter of the number of shrimp, as expected by the difference in net sizes, whereas the general shapes of the histograms appear quite similar across net sizes, suggesting there is not a difference in shrimp length distribution between nets across time.
 
-```{r sexResults, echo=TRUE, cache=TRUE}
-tableData <- cbind(summary(nbmod)$coeff)
-colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
-                         "Unknown Gender", "August", "September", 
-                         "Net Size * Females", "Net Size * Juveniles", 
-                         "Net Size * Unknowns", "Net Size * August",
-                         "Net Size * September",
-                         "Females * August", "Juveniles * August", 
-                         "Unknown Gender * August",
-                         "Females * September", "Juveniles * September", 
-                         "Unknown Gender * September")
-kable(tableData, digits = 4, format="pandoc")
+```{r, eval=TRUE, include=TRUE}
+ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ gender) + 
+  ggtitle("Lengths by sex and net size")
 ```
 
-Thus, we refit the model without the interaction between net size and sampling date. To do this we start by loading the new covariate matrix $\mathbf{X}$.
+When we plot the distribution of lengths between the net sizes with respect to sex class, the distributions look quite similar as well.
+This suggests that there is not a lot of difference in length distribution of shrimp between the two net sizes when broken down by sex class.
 
-```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE}
-X <- model.matrix(~ net * gender + gender * date, data=mysisSexCountData)
-nbmod <- glm.nb(y ~ . -1, data = data.frame(y=mysisSexCountData$count, X))
+Next, we plot histograms of the sampling events (80 sampling events – 10 sites, 2 nets, 4 dates) as a paired data set.
+
+```{r pairedTTestLength}
+y1 <- by(mysisLengthData$y[mysisLengthData$net == "Small Net"], 
+         mysisLengthData$label[mysisLengthData$net == "Small Net"], mean)
+y1 <- y1[!is.na(y1)]
+
+y2 <- by(mysisLengthData$y[mysisLengthData$net == "Large Net"], 
+         mysisLengthData$label[mysisLengthData$net == "Large Net"], mean)
+y2 <- y2[!is.na(y2)]
+pairttestlength <- t.test(y1, y2, paired=TRUE)
+qplot(y1-y2, xlab="difference in means")
 ```
 
-On this reduced model, the results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes (p=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
-```{r sexResultsReduced, echo=TRUE, cache=TRUE}
-tableData <- cbind(summary(nbmod)$coeff)
-colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
-                         "Unknown Gender", "August", "September", 
-                         "Net Size * Females", "Net Size * Juveniles", 
-                         "Net Size * Unknowns", 
-                         "Females * August", "Juveniles * August", 
-                         "Unknown Gender * August",
-                         "Females * September", "Juveniles * September", 
-                         "Unknown Gender * September")
-kable(tableData, digits = 4, format="pandoc")
+From the histogram of paired differences, it looks like there is some light evidence that the smaller net catches slightly smaller shrimp, although the distribution does seems to be centered near 0.
+From the paired t-test, we find that there is a significant difference ($t$ = `r pairttestlength$statistic`, $df$ = `r pairttestlength$parameter`, $p$-value = `r pairttestlength$p.value` ) in expected counts, although the $p$-value is close to the 0.05 level.
+Despite the statistically significant difference, the practical difference is quite small as judged by the histograms and small difference in shrimp size.
+
+```{r pairedTTestLengthtrim}
+y1trim <- by(mysisLengthData$y[mysisLengthData$net == "Small Net"],
+             mysisLengthData$label[mysisLengthData$net == "Small Net"], mean, trim=0.05)
+y1trim <- y1trim[!is.na(y1trim)]
+
+y2trim <- by(mysisLengthData$y[mysisLengthData$net == "Large Net"],
+             mysisLengthData$label[mysisLengthData$net == "Large Net"], mean, trim=0.05)
+y2trim <- y2trim[!is.na(y2trim)]
+pairttestlengthtrim <- t.test(y1trim, y2trim, paired=TRUE)
+qplot(y1trim-y2trim, xlab="difference in trimmed means")
 ```
 
-## Model for proportion of juveniles
-```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE}
-load("~/mysis/data/mysisSexCountData.RData")
-X <- model.matrix(~ net * date)
-library(betareg)
-brmod <- betareg(prop ~ . -1, data = data.frame(prop, X))
-summary(brmod)
+One could argue that because the larger net catches four times the number of shrimp, the large net is more likely to catch shrimp at the extremes of the size classes (either really small or really large Mysids).
+To account for this, we perform a paired t-test using a trimmed mean, removing the smallest and largest 5% of the lengths before calculating the sample mean.
+From the paired t-test using the trimmed means, we find that there is not a significant difference ($t$ = `r pairttestlengthtrim$statistic`, $df$ = `r pairttestlengthtrim$parameter`, $p$-value = `r pairttestlengthtrim$p.value` ) in expected counts, although the $p$-value is close to the 0.05 level.
+As with the untrimmed means, the practical difference is quite small as judged by the histograms and small difference in shrimp size.
+
+To better account for the different sample sizes (the group means are from unbalanced sample sizes), we fit a linear model and a robust linear model (that assumes overdispersion in the data).
+We start by checking for heterogeniety in the group variances.
+
+```{r}
+vars <- as.vector(by(mysisLengthData$y, mysisLengthData$label, var))
+qplot(vars, main = "var by sampling location", xlab="var")
+bart_test <- bartlett.test(mysisLengthData$y, mysisLengthData$label)
+levene_test <- leveneTest(mysisLengthData$y, mysisLengthData$label)
 ```
 
-```{r plotJuviCount, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center"}
-ggplot(data=data, aes(prop)) + geom_histogram() + facet_grid(net ~ .) + 
-  ggtitle("Histogram of proportion of juveniles by net size")
+When looking at the distribution of sampling variance by sampling occasion (10 locations, 2 nets, 4 different dates gives 80 total sampling ocassions), we see that there is some difference in variance of lengths by sampling occasion and Bartlett's test ($p$-value = `r bart_test$p.value`) and Levene's test ($p$-values = `r levene_test$Pr[1]`).
+
+To test if there is a difference in length of net sizes while accounting for heterogeneity, we use a weighted least squares regression where the weights are the inverse of the group variances.
+
+```{r}
+## weighted linear regression
+samps <- by(rep(1, length(mysisLengthData$y)), mysisLengthData$label, sum)
+w <- rep(0, length(mysisLengthData$y))
+for (i in 1:length(unique(mysisLengthData$label))) {
+  if (i==1) {
+    w[1:samps[1]] <- rep(1 / vars[i], samps[i])
+  } else {
+    w[(sum(samps[1:(i-1)])+1):sum(samps[1:i])] <- rep(1 / vars[i], samps[i])
+  }
+}
+summary(weightedlm <- lm(y ~ net + date + gender + station, 
+                          data=mysisLengthData, weights=w))
 ```
 
-To test for a change in the proportion of juveniles caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and the interaction between net size and sampling date.
-The results shown below show that there is no evidence of an effect on the proportion of juveniles caught between the two net sizes (p=`r I(round(summary(brmod)$coef$mean[2, 4], 4))`).
+From the weighted least squares regression, we find that there is a statistically signficant difference in mean Mysid length caught between the two nets ($t$ = `r summary(weightedlm)$coefficients[2, 3]`, $p$-value = `r summary(weightedlm)$coefficients[2, 4]`).
+Now, after accounting for the location, date, sex distribution, and heterogeneity of variance we find that the difference in mean length caught between the two net sizes is `r summary(weightedlm)$coefficients[2, 1]` mm.
 
-```{r juvenileResults, echo=TRUE, cache=TRUE}
-tableData <- cbind(summary(brmod)$coeff$mean)
-colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "August", "September", 
-                         "Net Size * August", "Net Size * September")
-kable(tableData, digits = 4, format="pandoc")
-```
+As a final analysis, we apply a robust regression, using an M-estimator model that accounts for heterogeniety in variance and presence of outlying observations.
 
-## Length model
-
-```{r, echo=TRUE, include=FALSE, message=FALSE, cache=TRUE}
-# data_length <- read.csv("~/mysis/data/lengthData.csv", skip = 4, header = TRUE)
-# net <- data_length$Size..m.
-# y <- data_length$Length..mm.
-# gender <- as.factor(substr(str_trim(as.character(data_length$Gender)), 0, 1))
-# station <- data_length$Station
-# date <- as.numeric(data_length$Date)
-# date[7330:7519] <- 4
-# date[7559:7773] <- 4
-# date[date == 2] <- 1      ## collapse into monthly values
-# date[date == 3] <- 2
-# date[date == 4] <- 3
-# date <- as.factor(date)
-# # plot(as.numeric(date), type='l', main = "Data looks better")
-# data <- data.frame(y=y, date=date, net=as.factor(net), gender=gender)
-# levels(data$date) <- c("July", "August", "Septempber")
-# levels(data$net) <- c("Small Net", "Large Net")
-# levels(data$gender) <- c("Female", "Juvenile", "Male", "Unknown")
-
-load("~/mysis/data/mysisLengthData.RData")
-X <- model.matrix(~ net + date + gender, data=mysisLengthData)
-idx <- attr(uniquecombs(X), "index")
-summary(lmmod <- lm(y ~ net + date * gender, data=mysisLengthData))
-library(lme4)
-# summary(lmemod <- lmer(y ~ net * date + net * gender +
-#                          date * gender + (1|station), data=mysisLengthData))
-# summary(lmemod <- lmer(y ~ net + date * gender + (1|station),
-#                        data=mysisLengthData))
-# summary(rlmmod <- rlm(y ~ net * date + net * gender + 
-#                         date * gender, data=mysisLengthData))))
-summary(rlmmod <- rlm(y ~ net + date * gender, data=mysisLengthData))
-tvalues <- summary(rlmmod)$coefficients[, 3]
+```{r}
+summary(lmmod <- lm(y ~ net + date + gender + station, data=mysisLengthData))
+summary(rlmmod <- rlm(y ~ net + date + gender + station, data=mysisLengthData, w=w))
+tvalues <- summary(rlmmod)$coefficients[2, 3]
 df <- summary(rlmmod)$df[2]
@@
 
-```{r plotLength, echo=TRUE, warning=FALSE, message=FALSE, out.width="300px", fig.align="center", fig.show="hold"}
-ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ date) +
-  ggtitle("Lengths by net size")
-ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ gender) + 
-  ggtitle("Lengths by sex and net size")
-ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(date ~ gender) + 
-  ggtitle("Lengths by sex and date")
+```{r lengthResults, eval=TRUE}
+tableData <- cbind(summary(rlmmod)$coeff, pvalues)
 ```
 
-```{r lengthResults, echo=TRUE, cache=TRUE}
-tableData <- cbind(summary(rlmmod)$coeff, pvalues)
-colnames(tableData) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")
-rownames(tableData) <- c("Intercept", "Net Size", "Females", "Juveniles", 
-                         "Unknown Gender", "August", "September", 
-                         "Females * August", "Juveniles * August", 
-                         "Unknown Gender * August",
-                         "Females * September", "Juveniles * September", 
-                         "Unknown Gender * September")
+For the robust linear model, we still find statistically significant differences in mean length between the net sizes ($t$-value = `r tvalues`, $p$-value = `r pvalues` on `r df` degrees of freedom), however the effect size `r summary(rlmmod)$coefficients[2, 1]` is still about the same as from the weighted linear model
 
-kable(tableData, digits = 4, format="pandoc")
-```
-
-```{r effect size, echo=TRUE, eval=FALSE}
+```{r effect size, eval=TRUE}
 SD <- sqrt(((sum(mysisLengthData$net == "Large Net") - 1) *
@@
 ```
-We see that there is a significant effect of net size on mean length caught (p = `r I(pvalues[2])`) after controlling for date, gender class, and an interaction between date and gender class.
-Although this is statisticallly significant, the effect size is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`).
-given a large sample size, a hypothesis test will show staistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05).
+
+Although the statistical tests using both the weighted linear model and the robust linear model show statistically significant differences in mean length caught between the net sizes, this is not unexpected because have very large sample sizes and and the regression coefficients are very small relative to the other sources of variation in the model.
+Of greater interest is whether the observed effect is of practical significance.
+
+The effect size of the mean length difference between the two net sizes is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`).
+Given a large sample size, a hypothesis test will show statistical significance unless the population effect size is exactly zero (this explains why all of the p-values in the table above are less than 0.05).
 Therefore, the practical effect of a difference in mean length of `r I(summary(rlmmod)$coefficients[2, 1])`mm on a species with a mean length of `r I(mean(mysisLengthData$y))`mm is small (this is the smallest effect of all the effects estimated by almost a factor of two) and a difference in means of this size is not of practical interest.
+Another measure of effect size is Cohen's $d$ which measures the difference in means relative to a pooled standard deviation.
+For our data, Cohen's $d$=`r I(d)`, which implies that the effect of net size is quite small in terms of practical significance.
+
+3) 
+The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.
+A visual inspection of counts by sex class shows no difference in counts by net size, but does show a change in counts over time.
+
+The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.
+
+```{r, eval=TRUE, include=TRUE}
+## load data
+load("~/mysis/data/mysisSexCountData.RData")   
+ggplot(data=mysisSexCountData, aes(count)) +         
+  geom_histogram() + facet_grid(net ~ gender) + 
+  ggtitle("Histogram of normalized counts by sex and net size")
+```
+
+From the histogram of counts broken down by sex class and net size, we see no outstanding differences of counts between net size.
+
+```{r, eval=TRUE, include=TRUE}
+ggplot(data=mysisSexCountData, aes(count)) +
+  geom_histogram() + facet_grid(gender ~ date) + 
+  ggtitle("Histogram of normalized counts by sex and date")
+```
+
+From the histogram of sex class by date, we see a pattern of juveniles maturing to males and females as time progresses and an increase in unknowns as juveniles grow in size but are not sexually differentiated.
+
+```{r, message=FALSE, results="hide"}
+## Normalize the counts <- divide by 4 because of the zeros for the small net
+mysisSexCountData$count[mysisSexCountData$net == "Large Net"] <- 
+  mysisSexCountData$count[mysisSexCountData$net == "Large Net"] / 4
+## Note: this gives warnings in the model as there are now non-integers
+summary(nbmod <- glm.nb(count ~ net + date + station + gender, data = mysisSexCountData))
+## Now fit the model with rounded counts
+mysisSexCountData$count[mysisSexCountData$net == "Large Net"] <- 
+  round(mysisSexCountData$count[mysisSexCountData$net == "Large Net"])
+## Note: this no longer gives a warning
+summary(nbmod <- glm.nb(count ~ net + date + station + gender, data = mysisSexCountData))
+```
+
+ To test for a change in counts broken down by sex category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling, sampling location, and sex class.
+The results show that there is no effect of net size on counts broken down by sex class ($z$-value = `r summary(nbmod)$coeff[2, 3]`, $p$-value = `r summary(nbmod)$coeff[2, 4]`).
+
+We are also interested if the proportion of juveniles is different between the two net sizes.
+Because the data are proportions, we cannot use methods that assume normally distributed data.
+One could use a contingency table to test homogeneity of proportions, but we prefer to use a glm known as beta regression (where a contingency table for homogeneity of proportions is a special case of beta regression).
+
+```{r, eval=TRUE, include=TRUE}
+load("~/mysis/data/mysisJuvenileData.RData") 
+ggplot(data=mysisJuvenileData, aes(prop)) + geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Histogram of proportion of juveniles by net size")
+```
+
+Looking at the histogram of proportions of juveniles by net size and by date, we see that there is more variability in the proportions for the small net (which is expected because the counts are smaller for the small net and therefore the proportions are more variable).
+We can see that the proportion of juveniles decreases through time and appears similar between the two net sizes.
+
+```{r, message=FALSE, eval=TRUE}
+brmod <- betareg(prop ~ net + date + station, data=mysisJuvenileData)
+```
+
+To test for a change in the proportion of juveniles caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and sampling location.
+The results shown below show that there is no evidence of an effect on the proportion of juveniles caught between the two net sizes (z=`r I(round(summary(brmod)$coef$mean[2, 3], 4))`, p=`r I(round(summary(brmod)$coef$mean[2, 4], 4))`).
+
+Also of interest is whether the ratio of females to males caught is the same between the two net sizes.
+Like the model for the proportion of juveniles, we could use a contingency table for heterogeneity of proportions, but we instead use the more flexible beta regression model.
+
+```{r, eval=TRUE, include=TRUE}
+load("~/mysis/data/mysisSexData.RData")
+ggplot(data=mysisSexData, aes(prop)) + geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Histogram of proportion of females by net size")
+```
+
+Looking at the histogram of proportions of females by net size and by date, we see that there is more variability in the proportions for the small net (which is expected because the counts are smaller for the small net and therefore the proportions are more variable).
+We can see that the proportion of females appears to increase through time and appears similar between the two net sizes.
+
+```{r, message=FALSE, eval=TRUE}
+brmodsex <- betareg(prop ~ net + date + station, data=mysisSexData)
+```
+
+To test for a change in the proportion of females caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and sampling location.
+The model shows that there is no evidence of an effect on the proportion of females caught between the two net sizes (z=`r I(round(summary(brmodsex)$coef$mean[2, 3], 4))`, p=`r I(round(summary(brmodsex)$coef$mean[2, 4], 4))`).

### 2016-01-04 data/processData.R

@@ -71,3 +71,158 @@ save(mysisLengthData, file='~/mysis/data/mysisLengthData.RData')
 ## remove data to elimate conflicts
 rm(list=ls())
 
+## 
+## Juvenile ratio data
+##
+
+fullData <- read.csv("~/mysis/data/lengthData.csv", skip=4)
+
+males <- rep(0, 80)
+females <- rep(0, 80)
+juveniles <- rep(0, 80)
+unknowns <- rep(0, 80)
+size <- rep(0, 80)
+date_tmp <- rep(0, 80)
+station <- rep(0, 80)
+
+idx <- 1
+for(i in unique(fullData$Sample.Label)){
+  tmp_data = subset(fullData, fullData$Sample.Label == i)
+  tmp = substr(as.character(tmp_data$Gender), 0, 1)
+  males[idx] <- sum(tmp == "M")
+  juveniles[idx] <- sum(tmp == "J")
+  females[idx] <- sum(tmp == "F")
+  unknowns[idx] <- sum(tmp == "U")
+  size[idx] <- tmp_data$Size..m.[1]
+  date_tmp[idx] <- tmp_data$Date[1]
+  station[idx] <- tmp_data$Station[1]
+  idx <- idx + 1
+}
+prop <- juveniles / (juveniles + females + males + unknowns)
+
+## Correct for mislabeling
+date_tmp[70] <- 4
+date_tmp[72] <- 4
+date <- rep(0, 80)
+date[date_tmp == 1] <- "July"
+date[date_tmp == 2] <- "July"
+date[date_tmp == 3] <- "August"
+date[date_tmp == 4] <- "September"
+
+date <- factor(date, levels=c("July", "August", "September"))
+size[size == 0.5] <- "Small Net"
+size[size == 1.0] <- "Large Net"
+net <- factor(size, levels=c("Small Net", "Large Net"))
+station <- factor(station)
+mysisJuvenileData <- data.frame(prop=prop, date=date, net=net, station=station)
+
+## Save RData file
+save(mysisJuvenileData, file='~/mysis/data/mysisJuvenileData.RData')
+## remove data to elimate conflicts
+rm(list=ls())
+
+## 
+## Sex ratio data
+##
+
+fullData <- read.csv("~/mysis/data/lengthData.csv", skip=4)
+
+males <- rep(0, 80)
+females <- rep(0, 80)
+unknowns <- rep(0, 80)
[…]

### 2016-01-05 responseToReviewers/mysisReviewerResponse.Rmd
updated paper

@@
 
-We tested many of the population characteristics to see if there was any difference between the two sized nets.
-Our main questions of interest are 1) Is there a difference in abundance of Mysids caught between the two nets, 2) Is there a difference in mean length of Mysids caught between the two nets, and 3) Does the count when broken down by age and sex class differ between the two net sizes.
+To determine if there was any difference between the two sized nets, we have a few questions of interest.
+Our main questions of interest are 1) Is there a difference in density of *Mysids* caught between the two nets, 2) Is there a difference in mean length of *Mysids* caught between the two nets, and 3) Does the density when broken down by age and sex class differ between the two net sizes?
+Note that we will use count and density interchangeable in what follows because we normalize to a one $m^2$ area.
 
-1) 
-We test if the abundance of Mysids caught differs between the two nets.
-By looking at the distributions of counts across time, we see that they are similar between net sizes, but the normality of the counts is questionable.
+1) Is there a difference in density of *Mysids* caught between the two nets?
 
@@
 ggplot(data=mysisCountData, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
-  ggtitle("Histogram of normalized counts by net size and date")
+  ggtitle("Normalized counts by net size and date")
 ```
 
+By looking at the distributions of counts across time, we see that they are similar between net sizes, but the normality of the counts is questionable.
 From the histogram, we see little differences between the two net sizes, except for perhaps a spike in catches of about 120 by the small net in August.
@@
 
-From the paired t-test, we find that there is no significant difference ($t$ = `r pairttest$statistic`, $df$ = `r pairttest$parameter`, $p$-value = `r pairttest$p.value` ) in expected counts.
+From the paired t-test, we find that there is no significant difference ($t$ = `r pairttest$statistic`, $df$ = `r pairttest$parameter`, $p$-value = `r pairttest$p.value`) in expected counts.
 
@@
 
-We test for a difference in counts between the nets formally by fitting a model to compare if the count of Mysids caught by the large net is the same as the count of Mysids caught by the large net, after normalizing for the area of the two nets and controlling for covariates.
+We test for a difference in density between the nets formally by fitting a model to compare if the count of *Mysids* caught by the large net is the same as the count of *Mysids* caught by the large net, after normalizing for the area of the two nets and controlling for covariates.
 The class of models known as generalized linear models (glms) offers a rigorous solution for modeling count data.
-These models allow for regression models that control for the effects of covariates like linear regression but do not assume a normal distribution.
+These models allow for regression models that account for the effects of covariates like in linear regression but do not assume a normal distribution.
 For count data, there are two natural distributions to use: the Poisson distribution and the negative binomial distribution.
@@
 In our data, the count means are $(`r I(mns)`)$ and the count variances are $(`r I(vars)`)$, for the small and large nets, respectively.
-Therefore we use a negative binomial model
+Therefore, we use a negative binomial model
 $$
@@
 
-To test for a change in total counts between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling and sampling location.
+To test for a change in total counts between the two net sizes, we construct a negative binomial regression model that examines the effect of net size while accounting for date of sampling and sampling location.
 The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes ($z$-value = `r I(round(summary(nbmod)$coef[2, 3], 4))`, $p$-value=`r I(round(summary(nbmod)$coef[2, 4], 4))`).
@@
 
-2) 
-The next population characteristic of interest is whether we are catching the same size Mysids with each net. First we examine some histograms and get an idea of what the length data look like.
+2) Is there a difference in mean length of *Mysids* caught between the two nets?
+First we examine some histograms and get an idea of what the length data look like.
 
@@
 
-When looking at the distribution of lengths between the two net sizes, we see that the histograms look quite similar across the different dates.
-What stands out is that the small net catches about a quarter of the number of shrimp, as expected by the difference in net sizes, whereas the general shapes of the histograms appear quite similar across net sizes, suggesting there is not a difference in shrimp length distribution between nets across time.
+When looking at the distribution of lengths between the two net sizes, we see that the histograms look quite similar across the different dates, with a general increasing trend through time.
+What stands out is that the small net catches about a quarter of the number of *Mysids*, as expected by the difference in net sizes, whereas the general shapes of the histograms appear quite similar across net sizes, suggesting there is not a difference in *Mysid* length distribution between nets across time.
 
@@
 When we plot the distribution of lengths between the net sizes with respect to sex class, the distributions look quite similar as well.
-This suggests that there is not a lot of difference in length distribution of shrimp between the two net sizes when broken down by sex class.
-
-Next, we plot histograms of the sampling events (80 sampling events – 10 sites, 2 nets, 4 dates) as a paired data set.
+This suggests that there is not a lot of difference in length distribution of *Mysids* between the two net sizes when broken down by sex class.
 
@@
 
-From the histogram of paired differences, it looks like there is some light evidence that the smaller net catches slightly smaller shrimp, although the distribution does seems to be centered near 0.
-From the paired t-test, we find that there is a significant difference ($t$ = `r pairttestlength$statistic`, $df$ = `r pairttestlength$parameter`, $p$-value = `r pairttestlength$p.value` ) in expected counts, although the $p$-value is close to the 0.05 level.
-Despite the statistically significant difference, the practical difference is quite small as judged by the histograms and small difference in shrimp size.
+From the histogram of paired differences, it looks like there is some light evidence that the smaller net catches slightly smaller *Mysids*, although the distribution appears to be centered near 0.
+From the paired t-test, we find that there is a significant difference ($t$ = `r pairttestlength$statistic`, $df$ = `r pairttestlength$parameter`, $p$-value = `r pairttestlength$p.value`) in expected counts, although the $p$-value is close to the 0.05 level.
+Despite the statistically significant difference, the practical difference is quite small as judged by the histograms and small difference in mean *Mysid* size difference (`r pairttestlength$estimate` mm).
 
@@
 
-One could argue that because the larger net catches four times the number of shrimp, the large net is more likely to catch shrimp at the extremes of the size classes (either really small or really large Mysids).
+One could argue that because the larger net catches four times the number of *Mysids*, the large net is more likely to catch *Mysids* at the extremes of the size classes (either really small or really large *Mysids*).
 To account for this, we perform a paired t-test using a trimmed mean, removing the smallest and largest 5% of the lengths before calculating the sample mean.
-From the paired t-test using the trimmed means, we find that there is not a significant difference ($t$ = `r pairttestlengthtrim$statistic`, $df$ = `r pairttestlengthtrim$parameter`, $p$-value = `r pairttestlengthtrim$p.value` ) in expected counts, although the $p$-value is close to the 0.05 level.
-As with the untrimmed means, the practical difference is quite small as judged by the histograms and small difference in shrimp size.
+From the paired t-test using the trimmed means, we find that there is not a significant difference ($t$ = `r pairttestlengthtrim$statistic`, $df$ = `r pairttestlengthtrim$parameter`, $p$-value = `r pairttestlengthtrim$p.value`) in expected counts, although the $p$-value is close to the 0.05 level.
+As with the untrimmed means, the practical difference is quite small as judged by the histograms and small difference in mean *Mysid* size difference (`r pairttestlengthtrim$estimate` mm).
 
 To better account for the different sample sizes (the group means are from unbalanced sample sizes), we fit a linear model and a robust linear model (that assumes overdispersion in the data).
-We start by checking for heterogeniety in the group variances.
+We start by checking for heterogeneity in the group variances.
 
@@
 
-When looking at the distribution of sampling variance by sampling occasion (10 locations, 2 nets, 4 different dates gives 80 total sampling ocassions), we see that there is some difference in variance of lengths by sampling occasion and Bartlett's test ($p$-value = `r bart_test$p.value`) and Levene's test ($p$-values = `r levene_test$Pr[1]`).
+When looking at the distribution of sampling variance by sampling occasion (10 locations, 2 nets, and 4 different sampling dates gives 80 total sampling occasions), we see that there is some difference in variance of lengths by sampling occasion and Bartlett's test ($p$-value = `r bart_test$p.value`) and Levene's test ($p$-values = `r levene_test$Pr[1]`).
 
@@
 
-From the weighted least squares regression, we find that there is a statistically signficant difference in mean Mysid length caught between the two nets ($t$ = `r summary(weightedlm)$coefficients[2, 3]`, $p$-value = `r summary(weightedlm)$coefficients[2, 4]`).
+From the weighted least squares regression, we find that there is a statistically significant difference in mean *Mysid* length caught between the two nets ($t$ = `r summary(weightedlm)$coefficients[2, 3]`, $p$-value = `r summary(weightedlm)$coefficients[2, 4]`).
 Now, after accounting for the location, date, sex distribution, and heterogeneity of variance we find that the difference in mean length caught between the two net sizes is `r summary(weightedlm)$coefficients[2, 1]` mm.
 
-As a final analysis, we apply a robust regression, using an M-estimator model that accounts for heterogeniety in variance and presence of outlying observations.
+As a final analysis, we apply a robust regression, using an M-estimator model that accounts for heterogeneity in variance and presence of outlying observations.
 
@@
 
-For the robust linear model, we still find statistically significant differences in mean length between the net sizes ($t$-value = `r tvalues`, $p$-value = `r pvalues` on `r df` degrees of freedom), however the effect size `r summary(rlmmod)$coefficients[2, 1]` is still about the same as from the weighted linear model
+For the robust linear model, we still find statistically significant differences in mean length between the net sizes ($t$-value = `r tvalues`, $p$-value = `r pvalues` on `r df` degrees of freedom), with the effect size `r summary(rlmmod)$coefficients[2, 1]` about the same as from the weighted linear model
 
@@
 
-3) 
+3)  Does the density when broken down by age and sex class differ between the two net sizes?
 The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.
-A visual inspection of counts by sex class shows no difference in counts by net size, but does show a change in counts over time.
-
-The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.
+A visual inspection of counts by sex class shows no difference in counts by net size, but does show a change in counts over time for each sex class.
 
@@
 load("~/mysis/data/mysisSexCountData.RData")   
+## Normalize the counts <- divide by 4 because of the zeros for the small net
+mysisSexCountData$count[mysisSexCountData$net == "Large Net"] <- 
+  mysisSexCountData$count[mysisSexCountData$net == "Large Net"] / 4
+
 ggplot(data=mysisSexCountData, aes(count)) +         
   geom_histogram() + facet_grid(net ~ gender) + 
-  ggtitle("Histogram of normalized counts by sex and net size")
+  ggtitle("Normalized counts by sex and net size")
 ```
@@
   geom_histogram() + facet_grid(gender ~ date) + 
-  ggtitle("Histogram of normalized counts by sex and date")
+  ggtitle("Normalized counts by sex and date")
 ```
@@
 ```{r, message=FALSE, results="hide"}
-## Normalize the counts <- divide by 4 because of the zeros for the small net
-mysisSexCountData$count[mysisSexCountData$net == "Large Net"] <- 
-  mysisSexCountData$count[mysisSexCountData$net == "Large Net"] / 4
 ## Note: this gives warnings in the model as there are now non-integers
@@
 
- To test for a change in counts broken down by sex category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling, sampling location, and sex class.
-The results show that there is no effect of net size on counts broken down by sex class ($z$-value = `r summary(nbmod)$coeff[2, 3]`, $p$-value = `r summary(nbmod)$coeff[2, 4]`).
+To test for a change in counts broken down by sex category between the two net sizes, we construct a negative binomial regression model that examines the effects of net size, date of sampling, sampling location, and sex class.
+The results show that there is no effect of net size on counts when controlling for sampling date, sampling location, and sex class ($z$-value = `r summary(nbmod)$coeff[2, 3]`, $p$-value = `r summary(nbmod)$coeff[2, 4]`).
 
-We are also interested if the proportion of juveniles is different between the two net sizes.
-Because the data are proportions, we cannot use methods that assume normally distributed data.
-One could use a contingency table to test homogeneity of proportions, but we prefer to use a glm known as beta regression (where a contingency table for homogeneity of proportions is a special case of beta regression).
+We can also examine if the counts within a sex class are different between the net sizes. To begin, we compare the counts of juveniles between the two nets.
 
 ```{r, eval=TRUE, include=TRUE}
-load("~/mysis/data/mysisJuvenileData.RData") 
-ggplot(data=mysisJuvenileData, aes(prop)) + geom_histogram() + facet_grid(net ~ date) + 
-  ggtitle("Histogram of proportion of juveniles by net size")
+juvenileCountData <- data.frame(count= mysisSexCountData$count[mysisSexCountData$gender == "J"],
+                                date=mysisSexCountData$date[mysisSexCountData$gender == "J"], 
+                                net=mysisSexCountData$net[mysisSexCountData$gender == "J"], 
+                                station=mysisSexCountData$station[mysisSexCountData$gender == "J"])
+maleCountData <- data.frame(count= mysisSexCountData$count[mysisSexCountData$gender == "M"],
+                                date=mysisSexCountData$date[mysisSexCountData$gender == "M"], 
+                                net=mysisSexCountData$net[mysisSexCountData$gender == "M"], 
+                                station=mysisSexCountData$station[mysisSexCountData$gender == "M"])
+femaleCountData <- data.frame(count= mysisSexCountData$count[mysisSexCountData$gender == "F"],
+                                date=mysisSexCountData$date[mysisSexCountData$gender == "F"], 
+                                net=mysisSexCountData$net[mysisSexCountData$gender == "F"], 
+                                station=mysisSexCountData$station[mysisSexCountData$gender == "F"])
+
+ggplot(data=juvenileCountData, aes(count)) +
+  geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Normalized counts of juveniles by net size and date")
 ```
 
-Looking at the histogram of proportions of juveniles by net size and by date, we see that there is more variability in the proportions for the small net (which is expected because the counts are smaller for the small net and therefore the proportions are more variable).
-We can see that the proportion of juveniles decreases through time and appears similar between the two net sizes.
+The histogram shows no visual differences in the juvenile count between net sizes but shows a decrease in juvenile count through time, as expected.
+To test this formally, we construct a negative binomial model that controls for sampling date and station.
 
-```{r, message=FALSE, eval=TRUE}
-brmod <- betareg(prop ~ net + date + station, data=mysisJuvenileData)
+```{r}
+## count model for juveniles
+summary(nbmodjuvenile <- glm.nb(count ~ net + date + station, data = juvenileCountData))
+## count model for males
+summary(nbmodmale <- glm.nb(count ~ net + date + station, data = maleCountData))
+## count model for females
+summary(nbmodfemale <- glm.nb(count ~ net + date + station, data = femaleCountData))
 ```
 
-To test for a change in the proportion of juveniles caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and sampling location.
-The results shown below show that there is no evidence of an effect on the proportion of juveniles caught between the two net sizes (z=`r I(round(summary(brmod)$coef$mean[2, 3], 4))`, p=`r I(round(summary(brmod)$coef$mean[2, 4], 4))`).
-
-Also of interest is whether the ratio of females to males caught is the same between the two net sizes.
-Like the model for the proportion of juveniles, we could use a contingency table for heterogeneity of proportions, but we instead use the more flexible beta regression model.
-
-```{r, eval=TRUE, include=TRUE}
-load("~/mysis/data/mysisSexData.RData")
-ggplot(data=mysisSexData, aes(prop)) + geom_histogram() + facet_grid(net ~ date) + 
-  ggtitle("Histogram of proportion of females by net size")
-```
-
-Looking at the histogram of proportions of females by net size and by date, we see that there is more variability in the proportions for the small net (which is expected because the counts are smaller for the small net and therefore the proportions are more variable).
-We can see that the proportion of females appears to increase through time and appears similar between the two net sizes.
-
-```{r, message=FALSE, eval=TRUE}
-brmodsex <- betareg(prop ~ net + date + station, data=mysisSexData)
-```
-
-To test for a change in the proportion of females caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and sampling location.
-The model shows that there is no evidence of an effect on the proportion of females caught between the two net sizes (z=`r I(round(summary(brmodsex)$coef$mean[2, 3], 4))`, p=`r I(round(summary(brmodsex)$coef$mean[2, 4], 4))`).
+From our model, we see that time of year and location are important in influencing the juvenile counts, but net size is not ($z$-value = `r summary(nbmodjuvenile)$coeff[2, 3]`, $p$-value = `r summary(nbmodjuvenile)$coeff[2, 4]`).
+Similar negative binomial regression models for counts of males and females show similar effects ($z$-value = `r summary(nbmodmale)$coeff[2, 3]`, $p$-value = `r summary(nbmodmale)$coeff[2, 4]` and $z$-value = `r summary(nbmodfemale)$coeff[2, 3]`, $p$-value = `r summary(nbmodfemale)$coeff[2, 4]` for males and females, respectively).
+For further details and `R` code of the analysis, see the attached appendix and the online supplement at http://jtipton25.github.io/mysis/.
 
@@
 I prefer to look at the histograms and see that the distributions between net sizes for all of the different subsets of the data are consistent rather than do a formal statistical test (which is much more complicated).
-A quick and dirty extreme value analysis suggests that the model parameters for the generalized extreme value distributions are quite similar for the two net sizes, suggesting that both nets sample the large shrimp equally.
+A quick and dirty extreme value analysis suggests that the model parameters for the generalized extreme value distributions are quite similar for the two net sizes, suggesting that both nets sample the large *Mysids* equally.
 

### 2016-01-05 data/processData.R

@@ -202,7 +202,7 @@ for(i in unique(fullData$Sample.Label)){
 
 ## construct count vector
 count <- c(males, females, juveniles, unknowns)
-gender <- factor(rep(1:4, each=80))
+gender <- factor(rep(c("M", "F", "J", "U"), each=80))
 ## Correct for mislabeling
 date_tmp[70] <- 4
 date_tmp[72] <- 4

### 2016-01-12 _site/responseToReviewers/mysisReviewerResponse.Rmd
attempt at updating github page

--- 2016-01-05 responseToReviewers/mysisReviewerResponse.Rmd
@@
----
-title: "Effect of net size on estimates of abundance, size, age and sex ratio of Mysis diluviana Journal of Great Lakes Research - Statistical Analysis"
-author: "John Tipton"
-date: "January 3, 2016"
-output: word_document
----
-
 ```{r loadLibraries, echo=FALSE, include=FALSE, message=FALSE}
@@
 # knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
-knitr::opts_chunk$set(echo=FALSE)
+knitr::opts_chunk$set(echo=TRUE)
 knitr::opts_chunk$set(include=FALSE)
@@
 
-1) Is there a difference in density of *Mysids* caught between the two nets?
+## 1) Is there a difference in density of *Mysids* caught between the two nets?
 
@@
 pairttest <- t.test(y1, y2, paired=TRUE)
+summary(pairttest)
 ```
@@
 
-```{r totalCountModel, eval=TRUE, results="hide"}
+```{r totalCountModel, eval=TRUE}
 summary(nbmod <- glm.nb(count ~ net + date + station, data=mysisCountData))
@@
 
-2) Is there a difference in mean length of *Mysids* caught between the two nets?
+## 2) Is there a difference in mean length of *Mysids* caught between the two nets?
 First we examine some histograms and get an idea of what the length data look like.
@@
 
-When looking at the distribution of sampling variance by sampling occasion (10 locations, 2 nets, and 4 different sampling dates gives 80 total sampling occasions), we see that there is some difference in variance of lengths by sampling occasion and Bartlett's test ($p$-value = `r bart_test$p.value`) and Levene's test ($p$-values = `r levene_test$Pr[1]`).
+When looking at the distribution of sampling variance by sampling occasion (10 locations, 2 nets, and 4 different sampling dates gives 80 total sampling occasions), we see that there is some difference in variance of lengths by sampling occasion and Bartlett's test ($p$-value `r format.pval(bart_test$p.value)`) and Levene's test ($p$-value `r format.pval(levene_test$Pr[1])`).
 
@@
 
-3)  Does the density when broken down by age and sex class differ between the two net sizes?
+## 3)  Does the density when broken down by age and sex class differ between the two net sizes?
 The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates.

### 2016-01-12 responseToReviewers/mysisReviewerResponse.Rmd
attempt at updating github page

@@
 ---
-title: "Effect of net size on estimates of abundance, size, age and sex ratio of Mysis diluviana Journal of Great Lakes Research - Statistical Analysis"
+title: Effect of net size on estimates of abundance, size, age and sex ratio of Mysis
+  diluviana Journal of Great Lakes Research - Statistical Analysis
 author: "John Tipton"
 date: "January 3, 2016"
-output: word_document
+output: html_document
 ---

### 2016-01-12 _drafts/2016-01-12-Appendix.Rmd
attempt at updating github page

---
title: Effect of net size on estimates of abundance, size, age and sex ratio of Mysis
  diluviana Journal of Great Lakes Research - Statistical Analysis Appendix
author: "John Tipton"
date: "January 12, 2016"
layout: post
---

```{r loadLibraries, echo=TRUE, include=FALSE, message=FALSE}
## We start by loading packages and define helper functions
options(digits=3)
set.seed(123)
library(knitr)
# knitr::opts_chunk$set(fig.path='../images/mysisAnalysis-')
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(include=TRUE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(warning=FALSE)
library(ggplot2)
library(Gmisc)
library(MASS)
library(stringr)
library(mgcv)
library(lme4)
library(ismev)
library(car)
library(betareg)
```

To determine if there was any difference between the two sized nets, we have a few questions of interest. Our main questions of interest are 1) Is there a difference in density of *Mysids* caught between the two nets, 2) Is there a difference in mean length of *Mysids* caught between the two nets, and 3) Does the density when broken down by age and sex class differ between the two net sizes? Note that we will use count and density interchangeable in what follows because we normalize to a one $m^2$ area.

## 1) Is there a difference in density of *Mysids* caught between the two nets?

```{r totalCount, eval=TRUE}
load("~/mysis/data/mysisCountData.RData")                                 ## load count data
```

```{r plotCount, eval=TRUE, include=TRUE}
ggplot(data=mysisCountData, aes(count)) + geom_histogram() + facet_grid(net ~ date) + 
  ggtitle("Normalized counts by net size and date")
```

By looking at the distributions of counts across time, we see that they are similar between net sizes, but the normality of the counts is questionable. From the histogram, we see little differences between the two net sizes, except for perhaps a spike in catches of about 120 by the small net in August. Because counts cannot be less that zero and often are right skewed, the assumption of normality in the data is questionable. Hence, methods like linear regression, t-tests, and ANOVA that assume normal distributions are not ideal, but, for completeness we begin with a paired t-test.

```{r pariedTTest}
y1 <- mysisCountData$count[mysisCountData$net == "Small Net"]
y2 <- mysisCountData$count[mysisCountData$net == "Large Net"]
pairttest <- t.test(y1, y2, paired=TRUE)
pairttest
```

From the paired t-test, we find that there is no significant difference ($t$ = `r pairttest$statistic`, $df$ = `r pairttest$parameter`, $p$-value = `r pairttest$p.value`) in expected counts.

```{r totalCountPoisson, message=FALSE}
mns <- by(mysisCountData[, "count"], mysisCountData[, "net"], mean)       ## mean of counts by net size
vars <- by(mysisCountData[, "count"], mysisCountData[, "net"], var)       ## variance of counts by net size
```

We test for a difference in density between the nets formally by fitting a model to compare if the count of *Mysids* caught by the large net is the same as the count of *Mysids* caught by the large net, after normalizing for the area of the two nets and controlling for covariates. The class of models known as generalized linear models (glms) offers a rigorous solution for modeling count data. These models allow for regression models that account for the effects of covariates like in linear regression but do not assume a normal distribution. For count data, there are two natural distributions to use: the Poisson distribution and the negative binomial distribution. The Poisson distribution has the strong assumption that the mean equals the variance, which is often not met in practical datasets. In our data, the count means are $(`r I(mns)`)$ and the count variances are $(`r I(vars)`)$, for the small and large nets, respectively. Therefore, we use a negative binomial model
$$
\label{neg:bin}
y_i \sim \mbox{NegBin}(\mu_i, \phi),
$$
where $\mu_i$ is the mean of the negative binomial distribution for sampling event $i$ and $\phi$ is an overdispersion parameter that allows for the mean and variance to be different. We model $\mu_i$ with covariates using the canonical log link function
$$ 
\mu_i = log(\mathbf{X}_i \boldsymbol{\beta}),
$$
where $\mathbf{X}_i$ is the set of covariates for observation $i$. Then we perform inference on our coefficients $\boldsymbol{\beta}$, where the interpretation of $\beta_j$ is a percent change in count per unit change in $X_{ij}$.

```{r totalCountModel, eval=TRUE}
summary(nbmod <- glm.nb(count ~ net + date + station, data=mysisCountData))
## get coefficient estimates
coefs <- c(nbmod$coefficients, nbmod$theta)           
```

To test for a change in total counts between the two net sizes, we construct a negative binomial regression model that examines the effect of net size while accounting for date of sampling and sampling location. The results shown below show that there is no evidence of an effect on the number of counts observed between the two net sizes ($z$-value = `r I(round(summary(nbmod)$coef[2, 3], 4))`, $p$-value=`r I(round(summary(nbmod)$coef[2, 4], 4))`).

## 2) Is there a difference in mean length of *Mysids* caught between the two nets?
First we examine some histograms and get an idea of what the length data look like.

```{r lengthData}
load("~/mysis/data/mysisLengthData.RData")
```

```{r, eval=TRUE, include=TRUE}
ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ date) +
  ggtitle("Lengths by net size")
```

When looking at the distribution of lengths between the two net sizes, we see that the histograms look quite similar across the different dates, with a general increasing trend through time. What stands out is that the small net catches about a quarter of the number of *Mysids*, as expected by the difference in net sizes, whereas the general shapes of the histograms appear quite similar across net sizes, suggesting there is not a difference in *Mysid* length distribution between nets across time.

```{r, eval=TRUE, include=TRUE}
ggplot(data=mysisLengthData, aes(y)) + geom_histogram() + facet_grid(net ~ gender) + 
  ggtitle("Lengths by sex and net size")
```

When we plot the distribution of lengths between the net sizes with respect to sex class, the distributions look quite similar as well. This suggests that there is not a lot of difference in length distribution of *Mysids* between the two net sizes when broken down by sex class.

```{r pairedTTestLength}
y1 <- by(mysisLengthData$y[mysisLengthData$net == "Small Net"], 
         mysisLengthData$label[mysisLengthData$net == "Small Net"], mean)
y1 <- y1[!is.na(y1)]

y2 <- by(mysisLengthData$y[mysisLengthData$net == "Large Net"], 
         mysisLengthData$label[mysisLengthData$net == "Large Net"], mean)
y2 <- y2[!is.na(y2)]
pairttestlength <- t.test(y1, y2, paired=TRUE)
pairttestlength
qplot(y1-y2, xlab="difference in means")
```

From the histogram of paired differences, it looks like there is some light evidence that the smaller net catches slightly smaller *Mysids*, although the distribution appears to be centered near 0. From the paired t-test, we find that there is a significant difference ($t$ = `r pairttestlength$statistic`, $df$ = `r pairttestlength$parameter`, $p$-value = `r pairttestlength$p.value`) in expected counts, although the $p$-value is close to the 0.05 level. Despite the statistically significant difference, the practical difference is quite small as judged by the histograms and small difference in mean *Mysid* size difference (`r pairttestlength$estimate` mm).

```{r pairedTTestLengthtrim}
y1trim <- by(mysisLengthData$y[mysisLengthData$net == "Small Net"],
             mysisLengthData$label[mysisLengthData$net == "Small Net"], mean, trim=0.05)
y1trim <- y1trim[!is.na(y1trim)]

y2trim <- by(mysisLengthData$y[mysisLengthData$net == "Large Net"],
             mysisLengthData$label[mysisLengthData$net == "Large Net"], mean, trim=0.05)
y2trim <- y2trim[!is.na(y2trim)]
pairttestlengthtrim <- t.test(y1trim, y2trim, paired=TRUE)
pairttestlengthtrim
qplot(y1trim-y2trim, xlab="difference in trimmed means")
```

One could argue that because the larger net catches four times the number of *Mysids*, the large net is more likely to catch *Mysids* at the extremes of the size classes (either really small or really large *Mysids*). To account for this, we perform a paired t-test using a trimmed mean, removing the smallest and largest 5% of the lengths before calculating the sample mean. From the paired t-test using the trimmed means, we find that there is not a significant difference ($t$ = `r pairttestlengthtrim$statistic`, $df$ = `r pairttestlengthtrim$parameter`, $p$-value = `r pairttestlengthtrim$p.value`) in expected counts, although the $p$-value is close to the 0.05 level. As with the untrimmed means, the practical difference is quite small as judged by the histograms and small difference in mean *Mysid* size difference (`r pairttestlengthtrim$estimate` mm).

To better account for the different sample sizes (the group means are from unbalanced sample sizes), we fit a linear model and a robust linear model (that assumes overdispersion in the data). We start by checking for heterogeneity in the group variances.

```{r}
vars <- as.vector(by(mysisLengthData$y, mysisLengthData$label, var))
qplot(vars, main = "var by sampling location", xlab="var")
bart_test <- bartlett.test(mysisLengthData$y, mysisLengthData$label)
bart_test
levene_test <- leveneTest(mysisLengthData$y, mysisLengthData$label)
levene_test
```

When looking at the distribution of sampling variance by sampling occasion (10 locations, 2 nets, and 4 different sampling dates gives 80 total sampling occasions), we see that there is some difference in variance of lengths by sampling occasion and Bartlett's test ($p$-value `r format.pval(bart_test$p.value)`) and Levene's test ($p$-value `r format.pval(levene_test$Pr[1])`).

To test if there is a difference in length of net sizes while accounting for heterogeneity, we use a weighted least squares regression where the weights are the inverse of the group variances.

```{r}
## weighted linear regression
samps <- by(rep(1, length(mysisLengthData$y)), mysisLengthData$label, sum)
w <- rep(0, length(mysisLengthData$y))
for (i in 1:length(unique(mysisLengthData$label))) {
  if (i==1) {
    w[1:samps[1]] <- rep(1 / vars[i], samps[i])
  } else {
    w[(sum(samps[1:(i-1)])+1):sum(samps[1:i])] <- rep(1 / vars[i], samps[i])
  }
}
summary(weightedlm <- lm(y ~ net + date + gender + station, 
                          data=mysisLengthData, weights=w))
```

From the weighted least squares regression, we find that there is a statistically significant difference in mean *Mysid* length caught between the two nets ($t$ = `r summary(weightedlm)$coefficients[2, 3]`, $p$-value = `r summary(weightedlm)$coefficients[2, 4]`). Now, after accounting for the location, date, sex distribution, and heterogeneity of variance we find that the difference in mean length caught between the two net sizes is `r summary(weightedlm)$coefficients[2, 1]` mm.

As a final analysis, we apply a robust regression, using an M-estimator model that accounts for heterogeneity in variance and presence of outlying observations.

```{r}
summary(rlmmod <- rlm(y ~ net + date + gender + station, data=mysisLengthData, w=w))
tvalues <- summary(rlmmod)$coefficients[2, 3]
df <- summary(rlmmod)$df[2]
pvalues <- pt(abs(tvalues), df, lower.tail=FALSE) * 2
```

For the robust linear model, we still find statistically significant differences in mean length between the net sizes ($t$-value = `r tvalues`, $p$-value = `r pvalues` on `r df` degrees of freedom), with the effect size `r summary(rlmmod)$coefficients[2, 1]` about the same as from the weighted linear model

```{r effect size, eval=TRUE}
SD <- sqrt(((sum(mysisLengthData$net == "Large Net") - 1) *
              sd(mysisLengthData$y[mysisLengthData$net == "Large Net"])^2 + 
              (sum(mysisLengthData$net == "Small Net") - 1) *
              sd(mysisLengthData$y[mysisLengthData$net == "Small Net"])^2) / 
             (sum(mysisLengthData$net == "Large Net") + 
                sum(mysisLengthData$net == "Small Net")))

d <- (mean(mysisLengthData$y[mysisLengthData$net == "Large Net"]) - 
        mean(mysisLengthData$y[mysisLengthData$net == "Small Net"])) / SD
```

Although the statistical tests using both the weighted linear model and the robust linear model show statistically significant differences in mean length caught between the net sizes, this is not unexpected because have very large sample sizes and and the regression coefficients are very small relative to the other sources of variation in the model. Of greater interest is whether the observed effect is of practical significance.

The effect size of the mean length difference between the two net sizes is small (`r I(summary(rlmmod)$coefficients[2, 1])`mm) and the sample size is large (n = `r I(length(mysisLengthData$y))`). Given a large sample size, a hypothesis test will show statistical significance unless the population effect size is exactly zero (this explains why all of the $p$-values in the table above are less than 0.05). Therefore, the practical effect of a difference in mean length of `r I(summary(rlmmod)$coefficients[2, 1])`mm on a species with a mean length of `r I(mean(mysisLengthData$y))`mm is small (this is the smallest effect of all the effects estimated by almost a factor of two) and a difference in means of this size is not of practical interest. Another measure of effect size is Cohen's $d$ which measures the difference in means relative to a pooled standard deviation. For our data, Cohen's $d$=`r I(d)`, which implies that the effect of net size is quite small in terms of practical significance.

## 3)  Does the density when broken down by age and sex class differ between the two net sizes?
The next question we wish to explore is whether the total counts when grouped into sex classes of male, female, juvenile, and unknown vary with net size or other covariates. A visual inspection of counts by sex class shows no difference in counts by net size, but does show a change in counts over time for each sex class.

```{r, eval=TRUE, include=TRUE}
## load data
load("~/mysis/data/mysisSexCountData.RData")   
## Normalize the counts <- divide by 4 because rather than multiply by 4 to lessen the impact of zeros
mysisSexCountData$count[mysisSexCountData$net == "Large Net"] <- 
  mysisSexCountData$count[mysisSexCountData$net == "Large Net"] / 4

ggplot(data=mysisSexCountData, aes(count)) +         
  geom_histogram() + facet_grid(net ~ gender) + 
  ggtitle("Normalized counts by sex and net size")
```

From the histogram of counts broken down by sex class and net size, we see no outstanding differences of counts between net size.

```{r, eval=TRUE, include=TRUE}
ggplot(data=mysisSexCountData, aes(count)) +
  geom_histogram() + facet_grid(gender ~ date) + 
  ggtitle("Normalized counts by sex and date")
```

From the histogram of sex class by date, we see a pattern of juveniles maturing to males and females as time progresses and an increase in unknowns as juveniles grow in size but are not sexually differentiated.

```{r, message=TRUE, warning=TRUE, include=TRUE}
## Note: this gives warnings in the model as there are now non-integers
summary(nbmod <- glm.nb(count ~ net + date + station + gender, data = mysisSexCountData))
```

To correct for the warning in the model above, we work with densities rounded to the nearest whole number (the normalization from counts to densities introduces decimals that need to be rounded). Notice that the results (estimates and $p$-values) do not change in any meaninful manner by rounding.

```{r}
## Now fit the model with rounded counts
mysisSexCountData$count[mysisSexCountData$net == "Large Net"] <- 
  round(mysisSexCountData$count[mysisSexCountData$net == "Large Net"])
## Note: this no longer gives a warning
summary(nbmod <- glm.nb(count ~ net + date + station + gender, data = mysisSexCountData))
```

To test for a change in counts broken down by sex category between the two net sizes, we construct a negative binomial regression model that examines the effects of net size, date of sampling, sampling location, and sex class. The results show that there is no effect of net size on counts when controlling for sampling date, sampling location, and sex class ($z$-value = `r summary(nbmod)$coeff[2, 3]`, $p$-value = `r summary(nbmod)$coeff[2, 4]`).

We can also examine if the counts within a sex class are different between the net sizes. To begin, we compare the counts of juveniles between the two nets.

```{r, eval=TRUE, include=TRUE}
juvenileCountData <- data.frame(count= mysisSexCountData$count[mysisSexCountData$gender == "J"],
                                date=mysisSexCountData$date[mysisSexCountData$gender == "J"], 
                                net=mysisSexCountData$net[mysisSexCountData$gender == "J"], 
                                station=mysisSexCountData$station[mysisSexCountData$gender == "J"])
maleCountData <- data.frame(count= mysisSexCountData$count[mysisSexCountData$gender == "M"],
                                date=mysisSexCountData$date[mysisSexCountData$gender == "M"], 
                                net=mysisSexCountData$net[mysisSexCountData$gender == "M"], 
                                station=mysisSexCountData$station[mysisSexCountData$gender == "M"])
femaleCountData <- data.frame(count= mysisSexCountData$count[mysisSexCountData$gender == "F"],
                                date=mysisSexCountData$date[mysisSexCountData$gender == "F"], 
                                net=mysisSexCountData$net[mysisSexCountData$gender == "F"], 
                                station=mysisSexCountData$station[mysisSexCountData$gender == "F"])

ggplot(data=juvenileCountData, aes(count)) +
  geom_histogram() + facet_grid(net ~ date) + 
  ggtitle("Normalized counts of juveniles by net size and date")
```

The histogram shows no visual differences in the juvenile count between net sizes but shows a decrease in juvenile count through time, as expected. To test this formally, we construct a negative binomial model that controls for sampling date and station.

```{r}
## count model for juveniles
summary(nbmodjuvenile <- glm.nb(count ~ net + date + station, data = juvenileCountData))
## count model for males
summary(nbmodmale <- glm.nb(count ~ net + date + station, data = maleCountData))
## count model for females
summary(nbmodfemale <- glm.nb(count ~ net + date + station, data = femaleCountData))
```

From our model, we see that time of year and location are important in influencing the juvenile counts, but net size is not ($z$-value = `r summary(nbmodjuvenile)$coeff[2, 3]`, $p$-value = `r summary(nbmodjuvenile)$coeff[2, 4]`). Similar negative binomial regression models for counts of males and females show similar effects ($z$-value = `r summary(nbmodmale)$coeff[2, 3]`, $p$-value = `r summary(nbmodmale)$coeff[2, 4]` and $z$-value = `r summary(nbmodfemale)$coeff[2, 3]`, $p$-value = `r summary(nbmodfemale)$coeff[2, 4]` for males and females, respectively). For further details and `R` code of the analysis, see the attached appendix and the online supplement at http://jtipton25.github.io/mysis/.

## Additional extreme value analysis (not in the manuscript) for exploring if the nets sample the extremely large *Mysids* differently

I prefer to look at the histograms and see that the distributions between net sizes for all of the different subsets of the data are consistent rather than do a formal statistical test (which is much more complicated). A quick and dirty extreme value analysis suggests that the model parameters for the generalized extreme value distributions are quite similar for the two net sizes, suggesting that both nets sample the large *Mysids* equally.

```{r}
small_gev <- gev.fit(xdat=mysisLengthData$y[mysisLengthData$net == "Small Net"])
large_gev <- gev.fit(xdat=mysisLengthData$y[mysisLengthData$net == "Large Net"])
```
By comparing the estimates for the tail parameter of the extreme value distribution estimate, we find that there is overlap between the two net sizes. The 95% confidence interval for the small net tail parameter is `r paste("(", round(small_gev$mle[3] - 2 * small_gev$se[3], digits=3), ",", round(small_gev$mle[3] + 2 * small_gev$se[3], digits=3), ")", sep="")` and the large net tail parameter is `r paste("(", round(large_gev$mle[3] - 2 * large_gev$se[3], digits=3), ",", round(large_gev$mle[3] + 2 * large_gev$se[3], digits=3), ")", sep="")`. Thus both net sizes do not show evidence of heavy tail distributions.

### 2016-01-12 _drafts/mysisAnalysis.Rmd
github pages update???

@@
 
- To test for a change in counts broken down by sex category between the two net sizes, we construct a negative binomal regression model that examines the effects of net size, date of sampling, sampling location, and sex class.
-The results show that there is no effect of net size on counts broken down by sex class ($z$-value = `r summary(nbmod)$coeff[2, 3]`, $p$-value = `r summary(nbmod)$coeff[2, 4]`).
+To test for a change in counts broken down by sex category between the two net sizes, we construct a negative binomial regression model that examines the effects of net size, date of sampling, sampling location, and sex class.
+The results show that there is no effect of net size on counts when controlling for sampling date, sampling location, and sex class ($z$-value = `r summary(nbmod)$coeff[2, 3]`, $p$-value = `r summary(nbmod)$coeff[2, 4]`).
 
-We are also interested if the proportion of juveniles is different between the two net sizes.
-Because the data are proportions, we cannot use methods that assume normally distributed data.
-One could use a contingency table to test homogeneity of proportions, but we prefer to use a glm known as beta regression (where a contingency table for homogeneity of proportions is a special case of beta regression).
+We can also examine if the counts within a sex class are different between the net sizes. To begin, we compare the counts of juveniles between the two nets.
 
 ```{r, eval=TRUE, include=TRUE}
-load("~/mysis/data/mysisJuvenileData.RData") 
-ggplot(data=mysisJuvenileData, aes(prop)) + geom_histogram() + facet_grid(net ~ date) + 
-  ggtitle("Histogram of proportion of juveniles by net size")
+juvenileCountData <- data.frame(count= mysisSexCountData$count[mysisSexCountData$gender == "J"],
+                                date=mysisSexCountData$date[mysisSexCountData$gender == "J"], 
+                                net=mysisSexCountData$net[mysisSexCountData$gender == "J"], 
+                                station=mysisSexCountData$station[mysisSexCountData$gender == "J"])
+maleCountData <- data.frame(count= mysisSexCountData$count[mysisSexCountData$gender == "M"],
+                                date=mysisSexCountData$date[mysisSexCountData$gender == "M"], 
+                                net=mysisSexCountData$net[mysisSexCountData$gender == "M"], 
+                                station=mysisSexCountData$station[mysisSexCountData$gender == "M"])
+femaleCountData <- data.frame(count= mysisSexCountData$count[mysisSexCountData$gender == "F"],
+                                date=mysisSexCountData$date[mysisSexCountData$gender == "F"], 
+                                net=mysisSexCountData$net[mysisSexCountData$gender == "F"], 
+                                station=mysisSexCountData$station[mysisSexCountData$gender == "F"])
+
+ggplot(data=juvenileCountData, aes(count)) +
+  geom_histogram() + facet_grid(net ~ date) + 
+  ggtitle("Normalized counts of juveniles by net size and date")
 ```
 
-Looking at the histogram of proportions of juveniles by net size and by date, we see that there is more variability in the proportions for the small net (which is expected because the counts are smaller for the small net and therefore the proportions are more variable).
-We can see that the proportion of juveniles decreases through time and appears similar between the two net sizes.
+The histogram shows no visual differences in the juvenile count between net sizes but shows a decrease in juvenile count through time, as expected.
+To test this formally, we construct a negative binomial model that controls for sampling date and station.
 
-```{r, message=FALSE, eval=TRUE}
-brmod <- betareg(prop ~ net + date + station, data=mysisJuvenileData)
+```{r}
+## count model for juveniles
+summary(nbmodjuvenile <- glm.nb(count ~ net + date + station, data = juvenileCountData))
+## count model for males
+summary(nbmodmale <- glm.nb(count ~ net + date + station, data = maleCountData))
+## count model for females
+summary(nbmodfemale <- glm.nb(count ~ net + date + station, data = femaleCountData))
 ```
 
-To test for a change in the proportion of juveniles caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and sampling location.
-The results shown below show that there is no evidence of an effect on the proportion of juveniles caught between the two net sizes (z=`r I(round(summary(brmod)$coef$mean[2, 3], 4))`, p=`r I(round(summary(brmod)$coef$mean[2, 4], 4))`).
-
-Also of interest is whether the ratio of females to males caught is the same between the two net sizes.
-Like the model for the proportion of juveniles, we could use a contingency table for heterogeneity of proportions, but we instead use the more flexible beta regression model.
-
-```{r, eval=TRUE, include=TRUE}
-load("~/mysis/data/mysisSexData.RData")
-ggplot(data=mysisSexData, aes(prop)) + geom_histogram() + facet_grid(net ~ date) + 
-  ggtitle("Histogram of proportion of females by net size")
-```
-
-Looking at the histogram of proportions of females by net size and by date, we see that there is more variability in the proportions for the small net (which is expected because the counts are smaller for the small net and therefore the proportions are more variable).
-We can see that the proportion of females appears to increase through time and appears similar between the two net sizes.
-
-```{r, message=FALSE, eval=TRUE}
-brmodsex <- betareg(prop ~ net + date + station, data=mysisSexData)
-```
-
-To test for a change in the proportion of females caught between the two net sizes, we construct a beta regression model that examines the effects of net size, date of sampling and sampling location.
-The model shows that there is no evidence of an effect on the proportion of females caught between the two net sizes (z=`r I(round(summary(brmodsex)$coef$mean[2, 3], 4))`, p=`r I(round(summary(brmodsex)$coef$mean[2, 4], 4))`).
+From our model, we see that time of year and location are important in influencing the juvenile counts, but net size is not ($z$-value = `r summary(nbmodjuvenile)$coeff[2, 3]`, $p$-value = `r summary(nbmodjuvenile)$coeff[2, 4]`).
+Similar negative binomial regression models for counts of males and females show similar effects ($z$-value = `r summary(nbmodmale)$coeff[2, 3]`, $p$-value = `r summary(nbmodmale)$coeff[2, 4]` and $z$-value = `r summary(nbmodfemale)$coeff[2, 3]`, $p$-value = `r summary(nbmodfemale)$coeff[2, 4]` for males and females, respectively).
+For further details and `R` code of the analysis, see the attached appendix and the online supplement at http://jtipton25.github.io/mysis/.

### 2016-01-12 _site/countModel/countData.Rmd
github pages update???

--- 2015-10-12 countModel/countData.Rmd
@@
----
-title: "negativeBinomial"
-author: "John Tipton"
-date: "August 14, 2015"
-output: html_document
----
-
 ## Load the data
@@
 #   iter = 5000, chains = 4)
-m <- estimate_params(sm, data = list(N = length(count), y = count, X = X, p=dim(X)[2]),  pars = c("beta", "phi"))
+m <- estimate_params(sm, data=list(N=length(count), y=count, X=X, p=dim(X)[2]),
+                     pars=c("beta", "phi"))
 ```

### 2016-01-12 _drafts/2016-01-12-Appendix.Rmd
updated params.json

@@
 ---
-title: Effect of net size on estimates of abundance, size, age and sex ratio of Mysis
-  diluviana Journal of Great Lakes Research - Statistical Analysis Appendix
+title: Effect of net size on estimates of abundance, size, age and sex ratio of *Mysis
+  diluviana* - Statistical Analysis Appendix
 author: "John Tipton"

### 2016-11-04 responseToReviewers/mysisReviewerResponse.Rmd
bump

@@
 date: "January 3, 2016"
-output: html_document
+output: word_document
 ---
